\documentclass[14pt,a4paper,oneside]{memoir}

\setlrmarginsandblock{2.5cm}{2.2cm}{*}
\setulmarginsandblock{2.5cm}{2.5cm}{*}
\checkandfixthelayout

\usepackage{datetime2}

%\setlrmarginsandblock{3.5cm}{2.5cm}{*}
%\setulmarginsandblock{2.5cm}{*}{1}
%\checkandfixthelayout

% Keep indentations with ragged right.
\makeatletter
\newcommand\iraggedright{%
  \let\\\@centercr\@rightskip\@flushglue \rightskip\@rightskip
  \leftskip\z@skip}
\makeatother

\usepackage{ifxetex}
\usepackage{amssymb} % Essential to be included before xunicode
\usepackage{xcolor}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\ifxetex

\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}
\usepackage{polyglossia}
\setromanfont[Mapping=tex-text]{Liberation Serif}
\setsansfont[Mapping=tex-text]{Liberation Sans}
\setmonofont[Mapping=tex-text]{Liberation Mono}
\setdefaultlanguage{english}

\else % not XeTeX

\iffalse

\usepackage{type1ec}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage[english,russianb]{babel}

\else

\message{XeTeX only document}
\stop

\fi

\fi % not XeTeX

%%%%%%%%%%%%% Common preamble %%%%%%%%%%%%

%\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\usepackage{graphicx}
\usepackage{textcomp}

\usepackage{indentfirst}
\frenchspacing
\clubpenalty=10000
\widowpenalty=10000

\sloppy

\newcommand{\bsk}{\vspace{20pt}}
\newcommand{\sk}{\vspace{5pt}}
\newcommand{\pp}{\textcolor{red}}   % Prepositional Pronoun.
\newcommand{\lp}{\textcolor{brown}} % Lexicalized Pronoun.
\newcommand{\bl}{\textcolor{blue}}  % Just blue.
\newcommand{\nb}[1]{{\small \emph{\textcolor{blue}{#1}}}}

\usepackage{indentfirst}

\usepackage{csquotes}        % Defines \enquote
\usepackage{microtype}       % For hanging indents (no config needed)

%\input{preamble}

\chapterstyle{verville}
\usepackage[hidelinks]{hyperref}
\setsecnumdepth{subsection}

\newcommand{\inda}{\hspace{40pt}}
\newcommand{\indu}{\hspace{20pt}}

\begin{document}

\frontmatter

\begin{titlingpage}
\title{Breaking Out of Beginner's Spanish}
\posttitle{\par\vskip1em{\normalfont\normalsize\maltese\par}\end{center}}
\author{Joseph J. Keenan}
\date{2004}

\maketitle
\end{titlingpage}

\tableofcontents*

\mainmatter

\chapter{Ten Ways to Avoid Being Taken for a Gringo}

A gringo, in much of the Spanish-speaking world, is a person
who comes from abroad, speaks another language, and wears loud
shorts. In certain countries, such as Mexico, it refers specifically to
U.S. citizens, but even there the distinction is hazy. A Canadian or
German who acts like a gringo will be referred to as a gringo, birth certificates be damned. Act like a gringo and you will be called one; don't
act like one and you may be called one anyway. The word is descriptive
first-of a style, a cultural stance, a way of life-and derogatory only
later, if at all.

So how to avoid being taken for a gringo? The truth is, if you
were born outside the Spanish-speaking world, there is probably nothing you can do to hide the fact. You will never fully blend in nor
should you necessarily want to. But as you travel or mingle among
Spanish-speakers, you may wish to smooth over the most obvious differences that set you apart. We all want to be outstanding; standing out
is another matter altogether.

Since you're making an effort to speak and understand Spanish, you've already distinguished yourself from the stereotypical
gringo, that mythical beast of Latin American lore who wears obtrusive shirts, smacks gum, and tends to misplace his or her wallet. But
even if you're not one of them, you can still heed a few simple precautions that will help put you in tune with the local culture, be it in Patagonia or East Los Angeles. Dress codes and behavior are beyond the
scope of this book, but some general pointers may keep others from
pointing at you.

\section{Pronunciation}

Spanish pronunciation should be the easiest thing in the world
to master. Unlike English, where a letter can change its sound seemingly at will, Spanish letters have-with very few exceptions-the exact same sound word after word. To compare, think of the three different sounds of the letter a in the English pronunciation of the name
Abraham; in Spanish the letter a in Abraham has but a single sound,
repeated three times. Still, you'll need to practice to convince your
tongue to make the correct sound, to get your teeth to close or open on
cue, and to master the inflections. At first it will be a struggle, but
there is no reason why anyone can't learn to pronounce Spanish properly after enough use. Here are some tips on how to proceed:

\subsection{}

Spanish teachers always tell their students to practice repeating the vowel sounds: \emph{a, e, i, o, u}. Listen to these wise men and
women and practice, practice, practice. If it's more fun, follow your
litany with the phrase schoolchildren learn south of the border: \emph{El burro sabe más que tú} ("The burro knows more than you"). Move your
mouth as you repeat the vowels. Pretend someone fifty yards away is
trying to read your lips. Clip the vowel sounds short, as you might
imagine a Japanese colonel in a late-night, World War II movie would
do. \emph{A, E, I, O, U, A, E, I, O, U}\ldots{}

\subsection{}

Next come the sounds for the letters \emph{r} and \emph{rr}. The double
one trills, and so does the single one at the start of a word. Thus \emph{carro}
and \emph{rancho} have essentially the same \emph{r} sound. Your tongue won't want
to trill at first; it will make a scene about being made out of concrete
and will refuse to emit such a ridiculous sound. But you're not going to
let your tongue push you around, are you? Trill away! Pretend you're
Charo on "The Tonight Show": "R-r-r-really, Johnny, r-r-r-romance for
me is r-r-r-relaxing on a r-r-r-rug, listening to r-r-r-rock and r-r-r-roll." If
you've studied any French, this sound may be harder for you to get
used to at first. But if you try at all, you will learn it. Many gringos do
and, as far as anyone can tell, we're all born with the same kind of
tongue.

\subsection{}

The d between vowels or at the end of a word sounds more
like the \emph{th} in \emph{thus} than an English \emph{d}. \emph{Nada} is thus pronounced "natha," or close to it. So light is the mid-vowel \emph{d} that sometimes, in colloquial spoken Spanish, it's almost left out altogether; you may even
see \emph{nada} represented as \emph{na'a} or \emph{na'} in written dialogue. You shouldn't
take it that far, but do get the hang of the soft Spanish d by learning a
few words well: \emph{nada, limonada, edad, comida, ciudad, cansado}. All
regular past participles follow the same rule: \emph{hablado, conocido, bebido}, and so on. The \emph{d} at the beginning of a word in Spanish is also a
tad softer than its English counterpart, perhaps more like a \emph{dth} than a
solid \emph{d}. Shout the name David in English and you can almost feel
yourself spit; shouting it in Spanish is a much less moist affair.

\subsection{}

The Spanish letters \emph{c, z, j}, and \emph{ll} vary in pronunciation
from country to country. Don't let this bother you. Either adopt the
sound used in your country of choice or seek a middle ground. For
most purposes, you're safe pronouncing both the \emph{c} and \emph{z} as an English
\emph{s}, the \emph{j} as an \emph{h}, and the \emph{ll} as the \emph{y} of "yes." (Remember, of course, that
the hard \emph{c} in Spanish, as in \emph{ca-, co-}, and \emph{cu-}, sounds like a \emph{k}.) Other
noteworthy regional differences include the use of \emph{vos} instead of \emph{tú}
and the use of \emph{vosotros} instead of \emph{ustedes}. \emph{Vos} is used in much of
Central and South America and requires learning yet more verb endings (\emph{tú quieres = vos querés}). Still, if you are spending time in those
countries, you will probably want to use it. The same goes for \emph{vosotros}, which is used in Spain for the second-person plural and which
you will no doubt learn if you're picking up your Spanish there.

\subsection{}

Other regional differences are best left alone. In many
countries, for instance, there is a tendency to "swallow" the \emph{s} sound at
the end of a syllable, especially before consonants: estoy aquí becomes
e'toy aquí. (It reaches such extremes that there's even a joke about the
Cuban child who asks his mother how to form plurals. "Easy, chico,"
she says, "just add an \emph{s}: \emph{el coco, lo coco}.") There's really no good reason to learn to speak Spanish that way, or with any other regional dialect, unless you're keen on being identified as having studied in a specific country. Imagine an Asian immigrant speaking English like a
Boston cabbie, or a Uruguayan drawling like a Texan, and you'll understand why.

\subsection{}

In Spanish, the letters \emph{b} and \emph{v} sound the same: almost (but
not quite) like the English \emph{b}. Like its d sound, Spanish's \emph{b/v} sound is a
shade softer, especially between vowels. Thus \emph{ave} is not pronounced
either "ah-bay" or "ah-vay" but "ah-bvay." Say "ah-bay" fast, without
giving your lips time to spit out a hard \emph{b} sound, and you'll get the idea.

\section{The wrong word syndrome}

There are dozens of cases in Spanish where you will be
tempted to use a word that is patently wrong. Mostly this is a result of
misleading English cognates---words that look or sound the same in
English and in Spanish but harbor different meanings. Sometimes the
meaning will be close in Spanish, and the lazy language learner is content to use the cognate. But you aren't like that, are you? You want to
speak Spanish well and not be responsible for polluting it with English
usages. Swell. For people like you, Chapter 3 in this book is dedicated
to these "tricksters," and Appendix B covers more subtle nuances.
Don't worry about learning them all at once. Just try to remember
which words are tricky and then try to avoid stepping in them as you
go along.

\section{\emph{Yo-ismo}}

In English, a sentence is incomplete without a noun or pronoun for a subject. In Spanish, the subject of the sentence is often conveyed by the verb and is optional. Often, in fact, it is left out altogether, unless the speaker wants to emphasize the subject of the
statement. This quaint little grammatical fact affects you, the language
learner, in one important case: the first person. Since you are used to
including pronouns, you will tend to preface all of your first-person
comments with yo. But to a Spanish ear, this sounds like you are constantly calling attention to yourself: "\emph{I} want this" and "\emph{I} think that."
This affliction, dubbed "\emph{yo-ismo}," can in extreme cases make people
think you're a pretty snotty individual, when you and I know that's not
true. But why take chances? Try to say \emph{quiero} instead of \emph{yo quiero},
\emph{creo} instead of \emph{yo creo}, and so on. Later, when you've broken the habit,
you can go back to inserting the occasional \emph{yo} to emphasize a truly
personal opinion: \emph{El quiere casarse pero yo no quiero} ("He wants to
get married but \emph{I} don't").

\section{The stumbles}

The stumbles are what you get when you're asked a simple
question and your tongue runs off and hides behind your tonsils. It is a
common ailment of those who have studied some Spanish---and \emph{know}
what they want to say---but lack conversational practice. Remedying
this condition requires practice, of course, but also a careful study of
interjections, pert comebacks, snappy answers, and sentence starters.
These useful words and phrases will get you through almost every
situation requiring sudden tongue work, but they are often neglected
in textbooks. Lists of these gems are included in later chapters. Pick
your favorites (there are usually several acceptable ones for each situation) and store them close to your tongue.
Speaking of the stumbles, you should be especially cautious of
letting your English "crutch words" slip into your Spanish. It sounds
awful: \emph{Quiero\ldots{} um\ldots{} ir a\ldots{} you know\ldots{} the\ldots{} er\ldots{} cine,
okay?} If you must, learn some Spanish crutch words and lean on them
instead: \emph{Quiero\ldots{} este\ldots{} ir\ldots{} o sea\ldots{} al cine, ¿no?} You'll still
sound like a space cadet, but at least a fairly fluent one.

\section{Formalities}

The gringo who makes the effort to get off the beaten path and
find out-of-the-way shops and cafés is almost by definition a fairly gregarious soul. But fear of the language can make this same person seem
timid, uptight, or arrogant, browsing in a shop for half an hour without
saying a word to anyone and leaving without so much as a good-bye. In
general, you should be happy to inflict your fledgling Spanish on anyone who crosses your path. But you should be especially eager to unleash it in cases calling for common courtesy. There is probably no
faster way to separate yourself from the pack of tourists and gawkers
than to look someone in the eye and speak to them in their own language---even if it's only to say "hello" and "good-bye."

For starters, you should always greet people whose lives you
have invaded, if only briefly. This doesn't mean you should walk down
the streets of Buenos Aires saying \emph{hola} to everyone, but it does mean
you should immediately recognize the existence of shop clerks, waiters, secretaries, and guests. The formula is simple. From the time you
wake up until 11:59 a.m., you say \emph{buenos días}; from 12:00 noon until
dark, you say \emph{buenas tardes}; and from dark until bedtime, you say
\emph{buenas noches}. Say it loud, say it proud. You'll be amazed how service
improves and prices drop after a pleasant greeting.

When you leave a place, remember to say \emph{gracias} if it's a commercial establishment, \emph{adiós} or \emph{hasta luego} if it's not. Better yet,
when leaving a shop, say \emph{muchas gracias} or \emph{muy amable}, gracias or
even \emph{muchas gracias muy amable}, all run together. You'll feel much
more cheerful walking out of a place after a heartfelt farewell, even if
the clerk did nothing more than stare at your back the whole time you
were in the store.

In general, Spanish requires more spoken formalities than English (at least as it's spoken today), which is nice because it gives you a
lot of opportunities to practice those key words and phrases. Skipping
over the formalities, on the other hand, will tag you as a gringo from
the get-go, which is what we've decided we want to avoid. For a fuller
treatment of formalities and politeness in general, there's a whole
must-read chapter just ahead with bounteous tips and details.

\section{The volume}

In Baltimore or Toronto or Oxford our friends and neighbors
seem to speak in reasonable tones. So why is it that these same people
seem to start shouting as soon as they get through Customs? This is
one of the great mysteries of cultural intercourse, and I don't foresee
resolving it here. But it is worth mentioning that, as a native English
speaker, you are expected to shout instead of speak and that a whole
continent of Latin Americans will be grateful if you manage to do otherwise. Gringos (remember them?) tend to bunch together and speak
English at volumes appropriate for a rock concert. I've even seen a
Mexican head of family ask the maitre d' to be seated "away from the
gringos" at a restaurant. As when greeting people, speak loud and
proud-but not too loud. Contrary to gringo folk wisdom, comprehension does not increase with volume. Normal speaking tones are best,
and by the time you find yourself shouting, absolutely no one will admit to understanding you at all.

\section{Adjectives}

One of the most frustrating things about learning a new language is not having access to that grab bag of adjectives that we rely on
to express our opinions. And since adjectives are a relatively second-class part of speech (after the big shots like nouns and verbs), many
beginning students of language tend to put off learning them until
some later phase of their study. In practical terms this produces human
beings whose whole range of descriptions goes from "very, very" on
one extreme to "not very" on the other. "How was the film?" \emph{Muy,
muy, muy buena}. "How about the meal?" \emph{No muy buena}.

You should try to break out of this rut as quickly as possible,
learning alternative ways of expressing your likes and dislikes. Pay special attention to the use of prefixes and (especially) suffixes in modifying adjectives in Spanish by learning a few suffixes and attaching
them to words you already know, you can quickly multiply the coverage of your vocabulary. \emph{Grande}, for example, can go up in size to \emph{grandísimo} and even \emph{grandotote} or deflate to a semisardonic \emph{grandecito}.
Also learn some words for that vast middle ground between good and
bad where, sad to say, most of our experiences tend to fall. Chapter 4
will try to help you do just that in your encounters with people.

\section{Speed kills}

Just as loud doesn't equal intelligible, fast definitely doesn't
equal fluent. Take it easy. You wouldn't try to break speed records on a
Kawasaki when you're just learning to ride, so why try with your Spanish? Spoken English has raised slurring practically to an art form, and
it's considered normal to modify (i.e., mispronounce) consonants or
vowels that get in our way. The usual result is a series of phrases like
"I dunno" and "Waddaya-wamee-tudo?" In spoken Spanish, each vowel
and consonant retains its particular, unalterable sound, no matter how
fast you're speaking. True, if you speak fast enough, people may not
catch your errors. But they won't catch your drift either, and you'll end
up having to repeat everything. If that's your strategy for getting extra
practice, \emph{adelante}. Just say it slowly the second time.

\section{Body language}

If slurring in English is almost an art form, then downwardly
mobile dressing in English-speaking (and other) countries is long overdue for a major museum exhibition. Far be it from me to tell you how
to dress on vacation or when prowling the barrio, but do remember the
. Spanish \emph{dicho: Como te ven, te tratan} ("How they see you, they treat
you"). Dressing down has not yet caught on in most Latin cultures,
perhaps because millions of people dress that way for reasons not related to fashion. If you stay close to the tourist bus, what you wear
isn't so important---but who wants to do that? You don't have to dress
to the nines to go buy a Coke, but you should at least be in the low
sevens. Otherwise, your clothes will be saying things about you that
your mouth never would.

What gringos often do, since we seem to be talking about
them, is to convert their Sunday barbecue outfits back home into all-purpose wear for their introduction to Latin culture. What they would
never wear to church they don't think twice about wearing to a Colombian cathedral or Guatemalan village church. If you intend to show the
local people that you respect their culture, the best way to start is by
letting your clothes speak for you. And the clothes that speak best are
the ones that cover knees and shoulders---at least away from coastal
cities. You don't have to care about any of this, of course, but if you
do, remind yourself that sin and skin are still closely linked in many
minds.

\section{Those crazy gringos}

Every now and then, it won't matter how well you say something in Spanish if what you are saying is so patently absurd that it
transcends language altogether. And what is judged as absurd can vary
widely from place to place. I remember witnessing a frustrated tourist trying to request, in flawless Spanish, \emph{hielo hervido} for his soft
drink. Now, if you're familiar with tourists, you'll probably realize that
"boiled ice" is a sort of shorthand way of saying "ice made from boiled,
or purified, water." But if you're a waiter in a small-town bus-station
restaurant in Latin America, you may not make that conceptual leap at
all. Instead, you will try your darndest to make sense of your customer,
and then will probably go back to the kitchen and heat up a couple of
ice cubes. Many innocent requests like this can turn into Major Cultural Confusions if you're not careful. Asking for ice to put in a soda
that is already cold is considered downright silly in many places, for
instance. Asking for a "pizza with meat on it," in at least one place I've
been, can lead to a pizza with a slab of steak lying on top. And so on.

This, of course, is part of the beauty of getting to know foreign
cultures: learning that what you had considered a given all your life is
often not a given at all. So if you find that nothing you say in Spanish
seems to get your point across, consider changing tack. What you are
saying, not how you are saying it, may be the culprit. Then change
your order to \emph{pollo frito} and a beer and forget about it. Just pray that
they don't come in a bowl, and together.

\chapter{Minding Your Verbal Manners}

Being polite is something many people, especially the young,
associate with visits to Grandma. In daily life, only bootlickers and
dweebs make a special effort to be polite; the rest of us are as we
are---take us or leave us.

Actually much of what goes for politeness is implicit in our
behavior and requires no special effort. Society has carved on our
minds the notion that if we don't follow certain preestablished, communal norms, it will use harsh and unfriendly epithets to describe us
behind our backs. So we take it for granted that you should open the
door for the elderly, avoid using expletives in public places, and refrain
from cutting in front of people on the exit ramp. We don't think of it as
being polite; we just do it because society says so.

Spanish-speaking society has its own set of unspoken norms
that you, as an outsider, won't have had beaten into your head from
birth. This means that you will have to pay attention to them and actually work at being polite. In addition, you will want to master the
subtleties of verbal manners that you now unthinkingly control in English. Consider these examples. Do you say the same sweet-sounding
phrases to a mean-faced bureaucrat as you do to a pleasant cashier? Of
course not. Nor do you use the same words or tone with an elderly
person that you use with someone more your age. Getting a feel for
subtleties in Spanish requires getting a handle on the language that is
,used to express manners.

In Spanish, there is really no good translation for "polite."
\emph{Cortés, amable}, and \emph{pulido} all come close but are better translated by
their English cognates: "courteous,' "amiable" (or "nice"), and "polished." Instead, a Spanish speaker will talk about someone's \emph{educación}, which goes beyond a person's schooling to cover upbringing in
general and manners in particular. \emph{Es una persona educada} means that
so-and-so is a person who has good manners and is polite in dealing
with others. The person in question could be a grease monkey in the
neighborhood lube shop or a physics professor, a kindergarten dropout
or a triple Ph.D.; \emph{educado} simply means that the person has decent
manners. "Well bred,' though somewhat out of favor in modern egalitarian societies, conveys the right idea.

In passing, it's worth noting that \emph{rudo} is not the equivalent of
"rude," nor is it a good opposite for \emph{educado}. A Spanish-speaker would
probably use \emph{mal educado, sin educación}, or \emph{de poca educación}, or
would resort to \emph{grosero}. This last word, in context, refers to a foulmouthed individual, but in a more general sense it comes closer to
"rude." \emph{Se portó muy grosero con nosotros} = "He was very rude
to us."

Being polite, of course, is more than simply uttering elegant
phrases at key points in a conversation. To achieve the rank of \emph{educado} and skirt all that is associated with \emph{grosero}, you'll need one part
proper language and one part common sense. We've already addressed
the rudiments of good manners in our brief review of greetings and
good-byes (see Chapter 1). The phrases \emph{buenos días, buenas tardes},
and \emph{buenas noches}, in conjunction with \emph{gracias} and \emph{hasta luego},
will get you through 90 percent of your daily encounters. But that's
about all they'll do. If your goal is to go beyond the point of just getting by, you'll want a more in-depth look at the universe of Spanish
formalities.

\section{Meetings and greetings}

Politeness begins upon meeting a person. You meet 'em, you
greet 'em. The question is, how?

The answer depends on the person you're meeting. If there's
little or no chance of ever seeing the person again, you're safe and sufficient with the \emph{buenos días\ldots{} hasta luego} formula. If you think the
person may figure in your future life, or if the meeting is the result of
an introduction, you're expected to go beyond that. From a second encounter onward, except in the case of repeated encounters with employees (shop clerks, the doorman, waiters, the gardener), you should
usually employ a more personal greeting than just "good day" or "good
night."

Most students of Spanish have been drilled in the basic forms
of greeting, but they're worth a quick review. On being introduced to a
person, you have at your disposal a number of stock responses: \emph{mucho
gusto, tanto gusto}, and \emph{encantado} (or \emph{encantada}) in roughly descending order of frequency. For less formal introductions and situations
\emph{¿qué tal?} and \emph{hola} work well. Save \emph{muchísimo gusto} for someone
you've been dying to meet.

Once you've been introduced to a person, you'll naturally be
expected to greet this individual at all future encounters, be it on the
street or at a party. How you do this reflects (a) who the person is and
(b) who you are in relation to that person. Here are some of the common options, in more or less descending order of formality:

\bsk

1. \emph{¿Cómo está?} or \emph{¿Como está usted?}

2. \emph{¿Cómo le va?}

3. \emph{¿Qué tal?}

4. \emph{¿Cómo estamos?}

5. \emph{¿Cómo estás?}

6. \emph{¿Qué hay de nuevo?}

7. \emph{¿Qué pasó?} or \emph{¿Qué pasa?} (varies by country)

8. \emph{¿Qué me cuentas?} or \emph{¿Qué me dices?}

9. \emph{¿Qué onda?} (Mexico) or \emph{¿Quiúbole?} (mostly Mexico and
the Caribbean)

10. Slangy variations of the preceding, such as \emph{¿Qué pasotes?}
or \emph{¿Qué pasión?} (from \emph{¿Qué pasó?}), \emph{¿Qué hongos?} (from ¿Qué onda?),
and so on. Save these for the people whose street gang you're looking
to join.

\bsk

It bears noting that all these expressions---except \emph{¿Cómo está?}
and \emph{¿Cómo está usted?}---imply some level of friendliness. Said another way, if you're not on especially friendly terms with the person,
stick to the first expression listed above. It is the only appropriate
form for greeting a person whose social, familial, occupational, or political position warrants your respect, and thus is the safe choice for
those who aren't, strictly speaking, your buddies. \emph{¿Cómo estamos?} has
paternalistic overtones and is often used by older people to greet
younger ones---even if the younger ones are thirty or forty years old.
This greeting is also a safe one when you're on good terms with the
person but aren't sure whether to use tu or usted (more on that bugaboo in a bit).

A common way of sprucing up any greeting is to use the person's name, title, or both. The commonest titles are \emph{Don} and \emph{Doña}
(for older people) and professional titles like \emph{Doctor, Contador} (accountant or C.P.A.), \emph{Ingeniero, Profesor, Maestro} (any teacher or craftsperson and sometimes even mechanics and plumbers), and the ubiquitous \emph{Licenciado} (virtually anybody who wears a tie). These are used far
more often than their English equivalents, especially in the workplace.

As an example of how greetings work, let's take the case of
Juan Doe, assistant director in charge of flange production, arriving at
his office. For simplicity's sake, let's presume all of the males in his
workplace are named Alberto Alvarez and all the females Teresa Ruiz.
Juan parks his car on the street and walks toward the office building. In
order, he meets and greets the following:

\bsk

\inda The eighty-year-old doorman:

\indu \emph{Buenos días, Don Alberto.}

\inda The security guard:

\indu \emph{Buenos días.}

\inda The sixty-year-old elevator operator:

\indu \emph{¿Cómo le va, Doña Tere?}

\inda The receptionist:

\indu \emph{Hola, Tere. ¿Cómo estás?}

\inda A same-aged colleague in the hall:

\indu \emph{¿Qué tal, Alberto?}

\inda A younger colleague at her desk:

\indu \emph{Buenos días, Tere. ¿Qué hay de nuevo?}

\inda A visiting branch manager:

\indu \emph{Buenos días, Señora Ruiz. ¿Cómo le va?}

\inda A co-worker and best friend:

\indu \emph{¿Quiúbole, Beto?}

\inda The immediate boss:

\indu \emph{Hola, Alberto. ¿Cómo estás?}

\inda An older co-worker:

\indu \emph{¿Qué me cuenta, Don Alberto?}

\inda An employee:

\indu \emph{Buenos días, Alberto. ¿Cómo estamos?}

\inda The division director:

\indu \emph{Muy buenos días, Señor Alvarez. ¿Cómo le va?}

\inda The office boy:

\indu \emph{¿Qué pasó, Beto?}

\inda The factory owner and CEO:

\indu \emph{Buenos días, Don Alberto. ¿Cómo está usted?}

\inda The secretary:

\indu \emph{Buenos días, Tere. ¿Qué tal?}

\bsk

By now, as you might imagine, Juan is exhausted and it's time for his
coffee break.

A couple of general tips on greetings are in order. First, note
that when greeting so many people, you will naturally gravitate toward
new ways of saying the same thing. That's because saying \emph{buenos días}
to twenty-five consecutive people can be extremely boring.

Second, use nicknames only if the person is accustomed to being called that. In other words, pay attention to whether others call a
certain Jose "Pepe" before you call him that. Use generic nicknames---such as \emph{viejo, compadre}, and \emph{jovenazo}---only when you feel certain
that the person won't be offended by your informality.

Third, if you're a male, avoid affectionate pet names for female
friends, employees, and co-workers. In Latin America it is common to
hear men calling women co-workers and employees things like \emph{linda}
and \emph{cariño}. To most North Americans this treatment is patronizing at
best and at worst borders on sexual harassment. Men in Latin America
have been slow about concerning themselves with these matters, but
that's no reason for you to imitate them.

Fourth, if you're a woman, stick to more formal modes of address until you're sure that your friendliness won't be taken as encouragement by the wolfish male mind. It's unfortunate that you have to
consider this issue, but that doesn't make it any less real.

And fifth, greet everyone possible, especially when meeting
a group of people. If you've met the people before, you are expected
to take the trouble of greeting each of them individually. Not to do
so can be interpreted as an offense. The same goes for saying goodbye. If you're in too much of a hurry or there are simply too many
people involved, make sure you issue an all-encompassing \emph{¿Qué tal,
cómo están?} or \emph{Hasta luego}, and make sure it's interpreted as all-encompassing.

\section{\emph{Usted} versus \emph{tú}}

From the moment you greet a person, you will start to think
about whether so-and-so is an "\emph{usted} person" or a "\emph{tú} person." Native
Spanish speakers make this decision instinctively; you will have to
think it through, and repeatedly. It is a concept that doesn't have an
easy English equivalent, but it is usually not that hard to keep straight.
Perhaps the most functional system for converting the concept into
English is to use usted in Spanish with anyone you would address with
"Mr.," "Mrs.," "Ms.," or "Miss" in English. Mr. Brown is your neighbor, so you use \emph{usted} with him. Once you get to know him and call
him "Fred," you can switch to \emph{tú}. Your lawyer is Ms. Smith, so she's
an \emph{usted} person; if you call your lawyer "Betty," then you would also
probably use \emph{tú} with her. And so on.

This rule of thumb works even when you don't actually know
the person's name. Your waiter's name is Juan Perez, but you probably
wouldn't know that. If you did, though, would you call him "Juan" or
"Mr. Perez?" That will generally depend upon his age, your age, and so
on. If it would feel awkward calling a twenty-two-year-old "Mr. Perez,"
go ahead and use \emph{tú} with him. Then again, if the twenty-two-year-old
happens to be an undersecretary for tax policy in the finance ministry---or a traffic cop---you would probably call him "Mr. Perez" and
thus use \emph{usted}.

As a rule, people aren't bashful about telling you to use \emph{tú}
with them if they feel it's appropriate. Almost no one will tell you, to
use \emph{usted} when you're using tu---it's the equivalent of putting you in
your place. So when in doubt, you're far safer using \emph{usted} and waiting
until you're told to do otherwise.

The best way to choose the right form is to listen to the conversation around you. Let's say you're meeting a group of friends at a
restaurant. Upon arriving, someone you don't know is sitting with
your friends. You are introduced to this person as Betty (your name),
and she to you as Yolanda (her name). This is your first clue: almost
certainly you are expected to \emph{tutear} (use \emph{tú} with) this person. To play
it safe, though, you can return the greeting with a noncommittal \emph{mucho gusto} and keep your ears open. If Sam asks Yolanda, \emph{¿Dónde estás
trabajando ahora?} then you can feel safe using \emph{tú} yourself. Likewise, if
you're addressed with \emph{usted}, you should respond with \emph{usted}.

That said, there are a few cases in which the \emph{tú-usted} relationship is not reciprocal---that is, when you will use \emph{tú} with the person
and he or she will use \emph{usted} with you, or vice versa. Almost always
this is the result of a considerable age difference. You might use \emph{usted}
with your friends' parents, for instance, and they will likely use \emph{tú}
with you. (This, incidentally, fits under the "Mr. and Mrs." rule.) Turn
the formula around for your children's friends.

\section{Magic words}

Besides greeting people, expressing thanks is probably the
daily act that most requires a modicum of civility. \emph{Gracias} is the
obligatory comment, of course, but you can spice it up with a \emph{muchas}
before it or a \emph{muy amable} after it or both, as noted in Chapter 1. By
doing so, you'll sound both more polite and more fluent. \emph{Muy gentil} is
also used, but it sounds somewhat strained.

When being thanked, you can respond with \emph{por nada, de
nada, no hay por que}, or \emph{no hay de que}. They're all about the same
and mean "You're welcome." You will hear \emph{para servirle} a lot from
clerks and waiters, but don't use it yourself unless it is genuinely
your job or duty to serve the person, or unless you're feeling especially subservient.

Another common linguistic nicety is asking permission. In
Spanish, as in English, the most typical way of asking permission is
essentially to excuse oneself for having the audacity to ask. Thus we
ask a person's "permission" to squeeze by them in an aisle. Here are
some options for communicating your humility while asking someone
to move over or let you by, again in descending order of formality:

\bsk

1. \emph{Con permiso}

2. \emph{¿Me permite?}

3. \emph{Perdón}

4. \emph{¿Se puede?}

5. \emph{Comper'} (a slangy version of con permiso)

6. \emph{Hágase un poco para allá, por favor}

7. \emph{Abreme espacio} or \emph{Abreme cancha}

8. \emph{Hazte pa'llá}

\bsk

The first five expressions are formal enough for just about any occasion. \emph{¿Me permite?} is sometimes used ironically, as when someone is
clearly blocking the way; say it very innocently for greatest effect. \emph{¿Se
puede?} is also the common way to ask to see something in a store; it
presumes you will wait for an affirmative response before, say, taking a
painting down off the wall or an earring out of a glass case. Unless the
object you wish to see is obvious (you're pointing at it, for instance),
you should use the full phrase: \emph{¿Se puede ver?}

The last three expressions on the list convey informality or
rudeness, depending on the person you are speaking to and your
tone of voice. \emph{Hazte pa'llá}, for instance, can be used for either "Scoot
over a little" (with a friend) or "Get out of the way" (With a stranger).
\emph{Abreme cancha} is very slangy.

The phrase for "Coming through!"---as when carrying a two-hundred-pound sofa down the hall---is \emph{¡Golpe avisa! Excúsame}, incidentally, does exist as a Spanish expression, but it doesn't mean "Excuse me." Use something (anything!) else.

Here's a cultural tip. One nicety that many foreigners
have trouble learning is to say gracias when someone sends \emph{saludos}
through them to another person: \emph{Salúdame a tu esposa} ("My regards to your wife") is something you will hear constantly if you are
married, for instance. Your English-speaking reflex will be to answer
"Okay, I will," but in Spanish it is customary to thank the person for
this \emph{detalle} ("thoughtfulness" or "consideration").

As \emph{gracias} is the universal word for "thank you," \emph{por favor} is
all you will ever need for "please." Always remember to use it when
asking for something. If you're tired of it and want to flex some Spanish muscles, use instead \emph{si es tan amable}, generally placed after your
request: \emph{Un café, si es tan amable}. It means \emph{por favor}.

\section{Sugar versus saccharine}

Spanish speakers are famous for going a bit overboard with
their floridness and politeness, and it is a matter of opinion whether
foreigners learning the language should leap in after them. Use your
own judgment. For instance, Spanish speakers will often refer to their
house as \emph{su casa} or \emph{tu casa}---"your house"---meaning that now that
you know them, you should consider their abode to be yours. This can
get pretty confusing at times: someone will be giving you directions to
their home and at the end they will point to the map they've improvised and say, "And here is your house." You'll be tempted to respond,
"But my house is nowhere near there!"---until you recall this subtle
gesture of hospitality. It can get even more confusing when a nonnative speaker, from whom such a gesture is generally not expected,
tries to communicate it.

Slightly less silly-sounding in the mouths of foreigners is the
statement \emph{Está usted en su casa} when someone comes to visit. It's really nothing more than a way of saying "Make yourself at home" and
shouldn't be made to sound more grandiose than that. To use it really
correctly, save it for when a houseguest makes a simple request, such
as "Can I use the phone?" To this you reply, with a wave of the hand,
\emph{Estás en tu casa}.

A word on homes is in order at this point. In much of the
Spanish-speaking world, homes are considered private reserves, and it
is not especially common to receive an invitation to visit someone's
house. So first of all, don't expect to receive such an invitation. And
don't be surprised if your offers of hospitality---"Hey, Pedro, how about
popping by after work for a beer?"---are taken as a nice gesture instead
of as a genuine invitation. Furthermore, never drop in unannounced on
a friend in the Spanish-speaking world. Finally, you may be told while
visiting someone's house that some object you admire \emph{es tuya} ("it's
yours," "take it"). Not only shouldn't you accept such offers, you
should be careful about making them. Sooner or later a literal-minded
guest might take you up on it!

Other sweet-sounding phrases that border on the sickly sweet
include solemn declarations of humility (saying \emph{su servidor} instead of
"I"), exaggerated requests for cooperation (\emph{Tenga usted la bondad de
traerme un café}), and overly formal greetings (\emph{Me es grato tener la
oportunidad de conocerlo}). All sound like you're reading out of a
phrasebook---and a phrasebook written for royal weddings, at that.

\section{Asking and getting}

Politeness is particularly important when asking someone for
something, since otherwise you may not get it. Bicultural lore is full of
anecdotes about Party A getting something in twenty minutes that
took Party B three weeks to get, simply because Party A asked politely
and Party B was viewed as rude. Whether you plan to use your Spanish
to get government authorizations or good directions, your command of
these niceties is critical.

It's easy to sound rude or clumsy when making a special or
even routine request in Spanish, especially since most students of the
language are taught the imperative as the sole way of asking for things:
Thus many students of Spanish will tell their hostess, \emph{Tráigame un café} ("Bring me a coffee"), thinking that's the correct, formal, and polite way to petition one. A good hostess will bring you one anyhow,
but on some interior level she's thinking, "Sure, here's one in your lap,
schmuck." Fortunately, it's quite easy to sound natural when asking
for something, but most Spanish texts neglect to mention the most frequently used form of the imperative in daily life: the indicative. That
is, most people don't say \emph{Tráigame un café, por favor} but \emph{¿Me trae un
café, por favor?} An added advantage to this form is that you don't have
to worry about those strange imperative forms and can stick to the
tried-and-true indicative instead.

The \emph{¿me trae? + por favor} formula is all you'll ever need for
restaurants and the like. But this use of the indicative also works for
most other situations and can employ a number of different verbs in
addition to \emph{traer}, especially \emph{permitir, dar, prestar}, and \emph{regalar}. \emph{Permitir} is the most formal: \emph{¿Me permite un cigarro?} will get a cigarette off
your future father-in-law. \emph{Prestar} and \emph{regalar} are the least formal, and
the latter implies you're going to keep what you're given: \emph{¿Me prestas
tu pluma?} means "Lend me your pen." \emph{¿Me regalas tu pluma?} is "Can
I have your pen (forever)?" In colloquial use, \emph{pasar} is common and
equates with \emph{prestar}: \emph{¿Me pasas el cenicero?} means "Pass the ashtray
over here, would ya?" Another common formal way of asking for
things is equivalent to "May I" in English: \emph{¿Puedo tomar un cigarro?}
("May I have a cigarette?").

\section{Etc.}

Phone Spanish is generally even more polite than face-to-face
Spanish. Listening to it, you will hear a lot of phrases like \emph{si es tan
amable} and \emph{si no es mucha molestia} tacked on to simple requests.
Once you get the hang of it, you'll be doing the same thing. To ask for
someone on the phone, you can be formal---\emph{¿Me puede comunicar can
el Sr. So-and-so?}---or informal---\emph{¿Está por ahí So-and-so?}. It will depend on whom you're calling. To say "Speaking" when someone calls
and asks for you, just say \emph{Él} (or \emph{ella}) \emph{habla}.

Interrupting is considered bad form in any language, of course,
but some foreigners seem to do it more when the conversation they are
interrupting is in a foreign language---i.e., one they don't understand
that well. Be aware of this, and be conspicuously polite when you do
need to interrupt, directing yourself to the party whom you are momentarily cutting out of the picture. Act, in other words, as if you were
cutting in on a dance. \emph{¿Me permite un momento?} you can ask.

Certain situations call for specific graces from you. Some are
verbal graces and some aren't, but all come under the heading of \emph{buena educación}.

\bsk

1. When you pass by someone who is eating, and presuming
you are at least vaguely acquainted with the person, wish him or her
\emph{provecho} or \emph{buen provecho} ("bon appetit," as we'd say in English).
Don't say it to total strangers in restaurants, though.

2. North Americans like to toss things. "Here's the pencil you
asked for," they'll say as they wing a fine-pointed no. 2 lead pencil by
your ear. "This is no good," they'll say as they crumple a sheet of paper
and lob it at the wastebasket. In the Spanish-speaking world, this behavior is considered barely short of barbaric.

3. Ditto for pointing at people.

4. Be conscious of not "giving your back" to people. Many
people from non-Latin cultures do it without intending offense. But in
Spanish-speaking cultures it's common to see people realize there is
someone behind them listening, do a half-turn, say \emph{perdón}, and continue speaking in a way that includes the previously excluded person.

\section{Sweet sorrow}

Departures are easily handled in Spanish by any of a number
of words, but \emph{adiós} and \emph{hasta luego} are sufficient for almost every circumstance. \emph{Adiós}, as you may have been taught, is generally used for
more lasting farewells. \emph{Nos vemos} is also a common colloquial send-off, equating with "See you later." \emph{Ciao} (or \emph{chao}) and \emph{bye} are making
rapid headway into Spanish from their respective languages.

Slightly more formal leave-taking makes use of phrases like
\emph{Que le vaya bien}, used only when the person you say it to is doing the
leaving. More fancy is \emph{Vaya con Dios}, but unless you're a nun or a
priest, it comes close to the saccharine category. When taking leave of
a group or passing by people on your way out, it's considered nice
manners to say \emph{Con permiso} to the people who are staying on. Usually it is
used for people whom you didn't actually say hello to on your way in.
Leaving the dentist's office, for instance, you say \emph{Gracias} to the dentist
and the receptionist and \emph{Con permiso} to the people sitting in the waiting room.

For less formal departures and with the younger set, you can
say things like \emph{Cuídate} (roughly, "Take it easy") and \emph{Pórtate bien} ("Behave yourself") to people when they are leaving. If it's nighttime and
the person is presumably leaving to go home to sleep, you can say \emph{Que
descanses} ("Rest up"). As a rule, try to respond with a farewell that is
different from the one used by the other person. If someone says \emph{Hasta
luego}, you say \emph{Nos vemos}; if they say \emph{Nos vemos}, answer \emph{Hasta luego}.

\chapter{Tricksters}

Every non-native speaker of a foreign language is afraid of
making mistakes. And good thing, too. A little fear forces us to concentrate harder and causes the memory of our mistakes to linger-long
enough to correct them the next time we open our mouths.

Most mistakes are grammatical and are eliminated only after
long periods of trial and error. Other mistakes, though, are almost the
fault of the languages themselves. Spanish and English, because of
their long history of coexistence and their many common origins, contain a lot of words that are similar on the surface but are used quite
differently. This makes instinctive translation at times as dangerous as
a cross-eyed knife-thrower.

Fortunately things aren't quite that bad. A lot of these false
cognates are so frequently tripped over by students of both languages
that they can be identified in advance. In the following list of "tricksters," you'll find some of the Spanish words most commonly misused
by native English speakers. Review the words and keep them in the
back of your brain. You won't overcome all of your instincts overnight,
but knowing your enemies-in this case, the tricksters-is the first
step toward conquering them.

\section{\emph{acostar}}

%ACOSTAR
Not "to accost," which is acosar. Acostar simply
means "to lie down" and is usually used in the reflexive. \emph{Me voy a
acostar} = "I'm going to lie down" or "I'm going to bed." If by doing
this you are accosting someone, you may be in the wrong bed.

\section{\emph{actual}}

%ACTUAL
Not "actual" but "current" or "present." \emph{La actual
administración} means "the current administration" or the one in power
at the moment. To convey "actual" in the sense of "factual" or "genuine," you would resort to \emph{verdadero, autentico, genuino, real}, etc. "This
is an actual Picasso" is expressed by \emph{Este es un auténtico Picasso}.

\section{\emph{actualmente}}

%ACTUALMENTE
Worth a special mention because "actually"
is so common as a sentence-starter in English: "You must be starving."
"Actually, I just ate." \emph{Actualmente} won't work here, since it means
"at present" or "currently." Confusing the two can lead to some
strange situations. Think of someone telling you, "You and your
brother are invited for lunch"---to which you want to answer, "Actually, he's my husband." If you use \emph{Actualmente, es mi esposo}, what
you are saying is "At present, he's my husband." For "actually," try the
phrase \emph{la verdad es que}: \emph{La verdad es que acabo de comer}. When "actually" is used in mid-sentence for emphasis, you can use either
\emph{realmente} or some other construction altogether. For example, "He actually ate it!" can be expressed as \emph{¡Realmente se lo comió!} or \emph{¡Sí se lo
comió!} or \emph{¡De verdad se lo comió!}

\section{\emph{afección}}

%AFECCIÓN
This doesn't mean "affection" but usually a
medical condition. \emph{Tiene una afección cardiaca} = "He (or she) has a
heart condition." Use \emph{afecto}---or better yet, \emph{cariño}---to convey your
affection.

\section{\emph{americano}}

%AMERICANO
Almost more of a political issue than a semantic one. Still, the use (and misuse) of \emph{americano} by "Americans"---i.e.,
U.S. citizens---should be kept in mind by anyone eager to avoid unnecessary offense. In Spanish and Spanish-speaking countries an \emph{americano} is simply a person from "the Americas," including South and
Central as well as North America. Thus if you tell a Chilean that you
are an \emph{americano}, you might get a smirk and the comment \emph{Yo también} ("Me too") in return. Incidentally, in the Spanish-speaking world
the correct way to refer to "the Americas" is in the singular: \emph{América}.
Schoolchildren are taught that, from the Bering Strait to Tierra del
Fuego, it is but one continent.

Oddly enough, there really is no perfect way for U.S. citizens
to state their origin in Spanish. \emph{Norteamericano} is the most common
word, but technically it includes Canadians and Mexicans as well. \emph{Estadounidense} has caught on of late, but besides being a mouthful it
overlooks the fact that other countries (the United Mexican States, for
instance) are also technically "United States." Even so, saying \emph{Soy de
los Estados Unidos} is probably the easiest way of explaining your
plight.

This confusion, incidentally, goes a long way toward explaining why \emph{gringo} is so common a word in countries like Mexico and why
you yourself will probably start to use it after a short time in these
places. The word can be used in an offensive way, but usually the
word's negative connotation is a result of the tone of voice. In Mexico,
especially, \emph{gringo} is very commonly used as an adjective as well; imports from the United States can easily (if colloquially) be referred to
as \emph{productos gringos}.

\section{\emph{argumento}}

%ARGUMENTO
A classic trickster. This word doesn't work
well as "argument" in the usual sense of a "heated discussion" or a
"quarrel." Instead, it should be reserved for prepared arguments (such
as lawyers' summations), debates, and carefully reasoned, often written
arguments. \emph{Argumento} describes a logical process, not a rather illogical throwing of pans and vases. For that, use \emph{pleito, disputa}, or \emph{disgusto} (see below), roughly in descending order of intensity. Perhaps the
best all-purpose translation of "argument" is another trickster, \emph{discusión}, which refers to a far more heated exchange than what we consider a "discussion" in English.

\section{\emph{asistir}}

%ASISTIR
Not "to assist" but "to attend" or "to be present at."
\emph{Asistencia} is the proper word for "attendance" (see \emph{atender} and \emph{audiencia} below), and an \emph{asistente} is technically just someone who is attending something. Nonetheless, you may come across \emph{gerente asistente} as a way of saying "assistant manager," though you'll probably
come across it at a local branch of a US. company. A more natural
Spanish construction would be \emph{subgerente}. For "to assist," use \emph{ayudar}.

\section{\emph{atender}}

%ATENDER
Not "to attend" (see \emph{asistir} above) but "to attend
to"---a significant difference. Remember, you can \emph{atender} a patient but you can't \emph{atender} a concert.

\section{\emph{audiencia}}

%AUDIENCIA
Another word in the "attendance" versus "assistance" imbroglio. Correctly, an \emph{audiencia} is usually a private interview granted to you by someone more important than you. In this
sense, it is usually used with \emph{conceder}: \emph{El ministro nos concedió audiencia} = "The minister granted us an interview." This meaning is
still in use in English, but it is usually reserved for the pope. In Spanish the town sewage commissioner can grant you one. \emph{La audiencia}
can be used for "the audience"---in a concert hall, for instance---but \emph{la
asistencia, el auditorio}, and \emph{el público} are all preferred.

\section{\emph{balde}}

%BALDE
This doesn't refer to hair loss but to a bucket. It's also
commonly seen in the expression \emph{en balde}, which means "in vain."
For "bald," you can use either \emph{calvo} (polite) or \emph{pelón} (irreverent).

\section{\emph{balón}}

%BALÓN
The common word for "ball," from about grapefruit
size on up. Smaller balls are called \emph{bolas} or \emph{pelotas}. "Balloon" is
\emph{globo}, just so you know.

\section{\emph{billón}}

%BILLÓN
A false cognate that many overlook. \emph{Un billón} is
1,000,000,000,000, or 10 to the twelfth power ($10^{12}$). It is equal to the
U.S. "trillion." (In England, this quantity is a "billion.") To convey the
US. "billion," or 10 to the ninth power (UK. "milliard"), you must say
\emph{mil millones}, or "a thousand millions."

\section{\emph{bizarro}}

%BIZARRO
A fairly archaic term for "chivalrous" or "brave"
(not "bizarre" in the sense of "peculiar"). Both "bizarre" and \emph{bizarro}
come from the Basque word for "bearded," which apparently suggested
strangeness to some and gallantry to others. You don't need to know
\emph{bizarro} to speak Spanish, but you should be aware that it doesn't mean
"bizarre" if you tend to translate your thoughts fairly literally from English to Spanish. For "bizarre," use \emph{raro} or \emph{extraño}.

\section{\emph{"capable"}}

%"CAPABLE"
This word doesn't even exist in Spanish, but if
you were to use it, it would be understood as "castratable." Saying
someone is \emph{un hombre capable}, therefore, does not mean he is "capable" but rather in imminent danger of being rendered "incapable."
The right word for "capable" is \emph{capaz}; "incapable" is \emph{incapaz}, which
covers "incompetent" as well.

\section{\emph{cargo}}

%CARGO
Almost works but not quite. \emph{Carga}, with the feminine ending, is the right word for "cargo." \emph{Cargo} in Spanish generally
means "job position" or "post." \emph{El encargado} is "the guy in charge."

\section{\emph{carpeta}}

%CARPETA
Usually a "portfolio" of the sort for keeping and
carrying papers. It could also be a simple manila folder or a file. It is
never a "carpet," which is covered by \emph{tapete} or \emph{alfombra}. Nonetheless,
thanks to the influence of English, it is said (perhaps apocryphally) that
Spanish-speaking residents of the United States say things like \emph{Voy a
vacunar la carpeta} for "I'm going to vacuum the carpet." In fact, this
statement means "I'm going to vaccinate the portfolio." (See Appendix
B for more English-influenced words and phrases.)

\section{\emph{chocar}}

%CHOCAR
A tricky trickster. This word has long existed in
Spanish to mean "to crash," as in what happens to a car that is driven
recklessly. A few hundred years ago, though, the English word "shock"
began to take on some trendy new scientific meanings, and a handful
of these were assigned to the old standby \emph{chocar} and its derivatives.
Thus while un choque has always meant "a crash," in recent years it
has been expanded to cover "a state of shock" (such as the driver's condition after the car's choque). It is also in widespread use for a powerful electrical shock, though \emph{chocar} as a verb is not used for "to shock
(electrically)." For that, use \emph{dar choque} or, in many countries, \emph{dar toques}. (Generally, \emph{un choque} will kill you and \emph{un toque} won't.) \emph{La lámpara me dio toques} = "The lamp gave me a shock." To confuse matters more, sometimes shock is used instead of \emph{choque}.

\section{\emph{chocante}}

%CHOCANTE
As an adjective, this works pretty well as
"shocking" but usually conveys a distinct disapproval that isn't implicit in the English word. In some contexts, especially in reference to
people, it comes close to meaning "offensive" or "rude." \emph{Me choca}, by
extension, is a common colloquial way of saying you strongly dislike
something: \emph{Me choca el chocolate} = "I hate chocolate."

\section{\emph{complexión}}

%COMPLEXIÓN
Not your skin texture, as in English, but your
general physical structure and shape, or "build"---in other words, fat,
skinny, pear-shaped, or just about right. Government forms in Spanish-speaking countries often ask about your \emph{complexión}, and as long as
you remember not to put "cleared up years ago," you'll be fine. For
skin condition, stick to \emph{piel} (all skin) or \emph{cutis} (especially facial skin).
"You have a nice complexion" is expressed by \emph{Tienes buen cutis}.

\section{\emph{compromiso}}

%COMPROMISO
Yes, it works as "compromise," but a far more
common usage is to mean "commitment." The verb \emph{comprometer} is
similarly double-edged. \emph{Me comprometo con las mujeres} could be a
politician's way of saying he or she is committed to his or her female
constituents and connotes no "compromising" situations. The same
sense of obligation is present in the common marketplace remark
\emph{sin compromiso}, which means you can try on a blouse, for instance,
"without committing" yourself to buy it. \emph{Compromiso} is also an
"appointment" or "engagement." \emph{Tengo un compromiso después de la
comida} = "I have an appointment after lunch." "Commitment" can.
also be used this way, but it sounds a bit like power-breakfast English,
whereas in Spanish it is an everyday expression.

\section{\emph{copa}}

%COPA
If you ask for your "cup of coffee" as a \emph{copa de café} in
the Spanish-speaking world, don't be surprised if you're served your
coffee "Irish" or with a coffee liqueur. In general usage, a \emph{copa} is a
stemmed glass or goblet of the sort used for wine or champagne. Thus
\emph{copa}, in a restaurant setting, almost always suggests an alcoholic beverage of some sort. Just as you wouldn't think of asking for coffee in a
goblet, your waiter won't think of serving something in a \emph{copa} without
a little booze in it. \emph{Taza} is the correct word for a coffee-style "cup."

\section{\emph{corriente}}

%CORRIENTE
This word is only partly tricky, but when it's
tricky, it's dangerously so. Basically you can use it safely as a noun to
mean "current"---any sort of electrical, river, or political currents---but not as an adjective. As a modifier, \emph{corriente} suggests "cheap" or
"trashy"; when used for people, it is a definite insult, equating with
"rude" or "vulgar." To convey "current" in the sense of "now in progress," you should take care to use \emph{presente} or \emph{actual} (see above).

\section{\emph{decepción}}

%DECEPCIÓN
A classic trickster. \emph{Decepción} means "disappointment" or "disillusionment," often with no suggestion whatsoever
of deceit. Likewise, \emph{decepcionar} means "to disappoint," and \emph{decepcionado} means "disappointed." \emph{Me decepcionó su novio} means you
were unimpressed by someone's boyfriend, not that he talked you out
of your inheritance. For "to deceive" and its derivatives you're better
off with \emph{engañar}. \emph{Defraudar} can work either way: "to defraud (monetarily)" or "to disillusion," "to let (someone) down."

\section{\emph{disgustar}}

%DISGUSTAR
Not "to disgust" but "to cause displeasure"---a
subtle but important difference. Turned around, with the speaker as
the indirect object (as with gustar), it means simply "to dislike." \emph{Me
disgustan los pepinos} means you don't like cucumbers, not necessarily
that they make you sick. For "to disgust," \emph{asquear} or \emph{dar asco} is appropriate: \emph{Los pepinos me dan asco}. Similarly, "disgusting" would be
\emph{asqueroso}. Used to describe a person, it conveys the idea of extreme
sleaziness. \emph{Es un tipo asqueroso} = "He's a real slimeball."

\section{\emph{"dum dum da dum dum, tump tump"}}

%"DUM DUM DA DUM DUM, TUMP TUMP"
A non-verbal trickster of the most dangerous sort. It's hard to convey the sound pattern
in writing (one rendering, perhaps from vaudeville days, is "shave and
a haircut---two bits"), but the pattern is familiar to everyone. Imagine
yourself knocking it on a door or tapping it out on your car horn. Got
it? Now consider that in certain Latin countries, Mexico especially,
what you've just said is, essentially, "Fuck your mother." Knock that
on a door in Mexico and be prepared to see someone with a shotgun
open it; tap it on your car horn and you'll have twenty vehicles gunning for you. Tap it on your car horn when there's a police car in front
of you and you've got serious problems.

\section{\emph{embarazado}}

%EMBARAZADO
The most famous trickster of all, leading to
all sorts of colorful anecdotes from travelers. The word actually can
mean "embarrassed" in certain contexts and in certain expressions,
but it also means "pregnant." By using it, even correctly, you open
yourself to no end of silliness and smirks. Better to stick to the common ways of saying "embarrassed": \emph{apenado} and \emph{penoso}, both from
the word \emph{pena} (see below). \emph{Dar pena} is good for "to embarrass." If you
don't want to speak in public because it embarrasses you, bow out with
a \emph{Me da pena}. For "How embarrassing!" (as in "What a fool I've made
of myself!"), a simple \emph{¡Que pena!} will suffice. A stronger concept like
"shame," often with moralistic overtones, is covered by \emph{vergüenza}.
\emph{Pena} is more the embarrassment that comes of shyness or prudishness.

\section{\emph{en absoluto}}

%EN ABSOLUTO
Not "absolutely" but the opposite---"absolutely not." If you want to avoid using it altogether, that's fine, too.
Use \emph{claro} for "absolutely" and \emph{claro que no} for "absolutely not." But
learn the correct meaning of \emph{en absoluto} for those times when it's
used on you.

\section{\emph{enfrente de}}

%ENFRENTE DE
A sneaky trickster and one that can cause
crossed signals with the best of them. It means "in front of" in the
sense of "across the way (or street) from" or "facing." Thus, if you tell
your friends to pick you up \emph{enfrente del cine} and you're waiting in
front of the movie theater, expect to see your friends waiting across
the street. \emph{Frente a} is equally misleading, meaning the same as \emph{enfrente de}. For "in front of" as we use it in English, try \emph{en la puerta de}
("at the door of") to avoid any misinterpretations. \emph{Al frente de} can also
be used, but why risk the confusion? (See also Chapter 11 under
"Front.")

\section{\emph{excitado}}

%EXCITADO
An easy one to slip in unawares, this trickster
probably conveys more than you bargained for. The English translation
is "aroused," sexual overtones included. For "excited," use \emph{emocionado}; for "exciting," \emph{emocionante}.

\section{\emph{éxito}}

%ÉXITO
Not "exit" but "success." In the music industry \emph{éxitos}
are "hits," lest you think \emph{Los grandes éxitos de Frank Sinatra} refers to
his greatest stage exits. "The exit" is \emph{la salida}.

\section{\emph{fábrica}}

%FÁBRICA
Not "fabric" but "factory." This is a good one to remember if you're going to get involved in business in Latin America, as
a lot of people are these days. If you're looking to buy fabric, on the
other hand, the word you want is \emph{tela}.

\section{\emph{gentil}}

%GENTIL
Not really "gentle" but "kind" or "courteous." \emph{Que
gentil} is a somewhat stuffy (and often ironic) way of saying "How
nice" or "That's great." To say "gentle," you'll want to use \emph{cuidadoso}
or even simply \emph{cuidado}: \emph{Mucho cuidado por favor con esa caja} =
"Please be gentle with that box." A "gentle" wind might be \emph{suave},
while a "gentle" person would be \emph{tierna}.

\section{\emph{informal}}

%INFORMAL
This word has come to have pretty much the
same meaning as its English cognate, with one important exception:
when used to refer to people, informal means "unreliable." It's the
word you use to refer to the plumber who swore that he'd be by last
Thursday to unstop your drains, only to vanish from the face of the
planet instead. To say that someone is "informal"---i.e., "laid-back"---you could use \emph{relajado} ("relaxed") or \emph{despreocupado}. For uses other
than personal ones, \emph{informal} is widespread, though purists tend to dislike it. \emph{Fue una reunión informal} = "It was an informal get-together."

\section{\emph{injuria}}

%INJURIA
An "injury," yes, but a moral and psychological
one---better known in English as an "insult" or "offense," and usually
a real dinger of one at that. Don't use it for "injury," which, if major, is
the noun \emph{herida}. If it's a small injury (a sprained ankle, for example),
the verb \emph{lastimarse} is more appropriate. Why is Pablo limping? \emph{Es que
se lastimó el pie jugando tenis} ("He hurt his foot playing tennis").

\section{\emph{intoxicado}}

%INTOXICADO
Unless it's a real binge you're talking about,
this word doesn't work for "drunk." It means "poisoned"---unintentionally, as a rule---and covers food poisoning, industrial toxins, overdoses, and the like. Another whole book could be written on how to
say "intoxicated" in the sense of "drunk" in Spanish. \emph{Ebrio} is the
best equivalent for "intoxicated," while \emph{borracho} is closest in tone to "drunk."

\section{\emph{introducir}}

%INTRODUCIR
You'll be tempted to use this for "to introduce,"
but don't. It means "to introduce" only in the sense of "to insert" or
"to add something in." Often it is used as a fairly fancy substitute for
\emph{meter} or "to stick in." For "to introduce to" with people, always use
\emph{presentar}. \emph{Ven, quiero presentarte a un amigo} = "Come on, I want to
introduce you to a friend of mine."

\section{\emph{largo}}

%LARGO
Not "large" but "long." Just a reminder.

\section{\emph{librería}}

%LIBRERÍA
Not "library" but "bookstore." Another reminder.
A "library" is a \emph{biblioteca}.

\section{\emph{media}}

%MEDIA
In Spanish, this (presumably) has nothing to do with
your favorite newscaster. It means "stocking," and in the plural,
"pantyhose." In math it means "mean." "Media," in the collective
sense of newspapers, magazines, radio, and television, is usually covered in Spanish by \emph{los medios} (de \emph{comunicación}).

\section{\emph{molestar}}

%MOLESTAR
In Spanish this word carries no sexual overtones
whatsoever. It's perfectly safe and extremely common for "to bother,"
\emph{No me molestes} = "Don't bother me." As an adjective, it works well
as "upset," "angry," or "uncomfortable." \emph{¿Estás molesto por algo?} =
"Are you upset about something?"

\section{\emph{ordinario}}

%ORDINARIO
Be careful using this term in regard to people.
Far from meaning "ordinary," it means "vulgar," "rude," and "crass."
\emph{Es un tipo muy ordinario} = "He's a slob." To describe an average, run-of-the-mill person just like you and me and a million other people, use
\emph{normal} or \emph{común:} \emph{Es un tipo normal}. \emph{El hombre común} is a good
translation of "the man in the street." (See also Chapter 4.)

\section{\emph{parientes}}

%PARIENTES
Not "parents," though they're \emph{parientes} too, but
"relatives"---all of them, from your grandparents and children to in-laws and cousins. What you share with these people is called \emph{parentesco}, or "kinship."

\section{\emph{pena}}

%PENA
Not "pain" (the physical kind) but "sorrow" and "embarrassment." \emph{Pena} is the word you want to learn so as to avoid using
\emph{embarazado} (see above). "Pain" is usually handled by \emph{dolor}, as is
"ache." A "headache" is a \emph{dolor de cabeza}. In the figurative sense of
a "pain in the neck" (or even lower), a good equivalent is \emph{lata}. \emph{¡Qué
lata!} = "What a pain!" \emph{Dar (la) lata} = "to be a pain," "to pester." To
people who are bothering you, you can say \emph{No des (la) lata} ("Stop being a pain").

\section{\emph{primer piso}}

%PRIMER PISO
If you're used to thinking of the ground floor as
the first floor, be prepared to think again in Spanish, where \emph{primer piso}
means "one flight up." The ground floor is usually called \emph{1a planta
baja}, or PB in the elevators.

\section{\emph{quitar}}

%QUITAR
You may try to use this for "to quit," especially
since Spanish at first glance doesn't seem to have a good word for "to
quit." Unfortunately, quitar isn't that word either. It means "to take
(something) off," "to take away." The word you need for "quit" depends on what you're quitting: if it's your job, the word is \emph{renunciar}, if
it's smoking or some other activity, \emph{dejar de}. To "quit" a computer
program, most translated software programs simply use \emph{salir}.

\section{\emph{realizar}}

%REALIZAR
There is some overlap with this word and its English cognate. But for the most common use of "to realize" in English,
\emph{realizar} does not work. The correct phrase is \emph{darse cuenta (de)}. "I realized too late he had a gun" is expressed by \emph{Me di cuenta demasiado
tarde de que traía pistola}. "Realize" in the common English sense can
often be handled perfectly competently by \emph{saber} ("to know") in Spanish. "I realize you're busy" thus becomes \emph{Sé que estás ocupada}. "I
didn't realize you were married" would be \emph{No sabía que estabas
casado}.

\section{\emph{receta}}

%RECETA
Almost always a medical prescription or a cooking
recipe in Spanish. It is never a "receipt," which would be \emph{recibo} or
\emph{nota}.

\section{\emph{ropa}}

%ROPA
Not "rope" but "clothes." "Rope" is usually either
\emph{soga} or \emph{cuerda}.

\section{\emph{sano}}

%SANO
This goes beyond mental health to cover all aspects of
health. In other words, it means "healthy." The word for "sane" is
\emph{cuerdo}. For "insane," use \emph{loco}. It's a bit unscientific and insensitive, but then so is "insane."

\section{\emph{sensible}}

%SENSIBLE
It means "sensitive," not "sensible." For "sensible" you should use \emph{sensato}. \emph{Una persona sensible} is "a sensitive
person." This is one of those confusing words you'll just have to
remember.

\section{\emph{sopa}}

%SOPA
Not "soap" but "soup." Creamy soups are simply
called \emph{cremas}, as in \emph{crema de champiñones} ("cream of mushroom
soup"). "Soap" is \emph{jabón}.

\section{\emph{soportar}}

%SOPORTAR
A common word in Spanish for "to tolerate,"
though it is more frequent in the negative in the sense of "can't stand."
\emph{No soporto la televisión} = "I can't stand television." Mixing it up will
get you some funny looks: \emph{Mi familia me soporta} doesn't mean your
family pays your bills but that your family tolerates you (barely). For
"to support" in the bill-paying sense use \emph{mantener}. In the sense of
physically supporting something (what a wall does for the ceiling, in
other words), \emph{sostener} is more accurate.

\section{\emph{sumar}}

%SUMAR
This verb may pop into your mind as a neat translation for "to sum up," but you should pop it right back out of there if
you want to be understood. \emph{Sumar} is the word for "to add" or "to add
up." \emph{¿Cuánto suma?} = "How much does that add up to?" For summaries and summations, stick to the verb \emph{resumir} (\emph{para resumir} = "to
sum up," \emph{en resumen} = "in sum"), although \emph{en suma} is safe for "in sum" as well.

\section{\emph{suplir}}

%SUPLIR
Not "supply," as you might think. Instead, it means
"to substitute" or "to fill in for." \emph{Suplo en la oficina a mi hermana
cuando está de viaje} = "I fill in for my sister at the office when she's
out of town." A \emph{suplente} is a "substitute"---a substitute teacher, for
instance. \emph{Substituto} and \emph{substituir} also exist and mean about the
same thing, but \emph{suplir} and its derivatives are more frequent (and far
easier to pronounce).

\section{\emph{tremendo}}

%TREMENDO
Often a close fit for "tremendous" but not always. Tremendo often comes closer to "outrageous" and can mean
"outrageously bad," "terrifying," or "terrible" as well as "outrageously
good" or "tremendous." The word is especially tricky around children,
it seems. \emph{Es un niño tremendo} describes a monstrous child capable of
the worst mischief. In short, be aware of the negative connotations
that frequently surround this word.

\section{\emph{tuna}}

%TUNA
Here's a real menu trickster. Tuna is not the fish but
the fruit of the prickly pear cactus, or \emph{nopal}---that is, the "prickly
pear" itself. "Tuna" is \emph{atún}. Both are perfectly edible, but if you have
your heart set on a tunafish sandwich, a serving of prickly pears may
not quite fill the bill.

\section{\emph{últimamente}}

%ÚLTIMAMENTE
In correct usage not "ultimately" but "recently." \emph{Últimamente ha estado enferma} = "She's been sick of late."
For "ultimately," use \emph{al final} or \emph{a fin de cuentas}. \emph{A fin de cuentas, es
su decisión} = "Ultimately it is his decision."

\section{\emph{vacunar}}

%VACUNAR
See carpeta above---and fast!

\section{\emph{vago}}

%VAGO
This word means "vague" when applied to things, but
it means "bum" or "tramp" when used for people. "Vagabond" is a
good cognate for this usage. To express "vague," use \emph{vago}, carefully,
and use \emph{impreciso} when referring to people or when there's any chance
of being misinterpreted.

\section{\emph{voluble}}

%VOLUBLE
Those of you who studied your vocabulary lessons
in high school will remember that this means "talkative." In Spanish,
though, it means "unreliable," "flighty," or "fickle" and is a good deal
more common than its English cognate.

\section{\emph{zorra}}

%ZORRA
"Foxy lady," even before Jimi Hendrix's time, has
been a slangy way to refer to an attractive female. And "a fox," by itself, means much the same thing for either men or women. Slang is a
dangerous thing to translate literally, however, and there is no better
example of this than the Spanish word \emph{zorra}, or "fox" (i.e., a canine
of the genus Vulpes). When applied figuratively to human beings, it
means "shrew," "slut," or "prostitute." What better way to learn your
tricksters than to get a slap in the face? If you must use corny come-ons, stick to \emph{guapo} and \emph{guapa}.

\chapter{Our Fellow Human Beings}

As we go about our daily lives, we are often put to the task of
describing our fellow human beings. And the vast range of personalities these humans represent requires an equally vast vocabulary of descriptive terms. Rare is it indeed to encounter a person whose behavior
can be wrapped up in so simple a concept as "good" or "evil." Instead,
human behavior runs the gamut from charming to wicked, honest to
malicious, overbearing to submissive, and happy to forlorn---sometimes even on the same day. For each of these many states and traits,
you will need words in Spanish.

Of course, learning all of the epithets employed to describe human personalities would be the intellectual equivalent of memorizing
a thesaurus. You could do that, naturally, but since you'll be wanting
to save a little brain space for such things as nouns and verbs, it's better to concentrate on a few common ways of referring to other individuals. Thus what follows is by no means an exhaustive list of all Spanish words describing people but rather a collection of some of the most
useful ones---ones you will probably hear, and may even be called, in
the company of Spanish speakers.

Where appropriate, rough English equivalents are provided.
The translations serve more to highlight the Spanish word's relative
strength and connotations than to provide a literal translation. As in
any language, you should choose your words carefully when referring
to other people. When in doubt, a good rule is to use them for the first
time when the person being described is not within hearing range.
Outside of a language-learning context, of course, this is called gossip.
But since you're still learning, it's allowed.

\section{The good}

Unless you are at a particularly low point in a mood swing,
you will probably concur that when all is said and done, there are still
a few good people out there. And with luck and perseverance, you may
even meet one of them. When you do, it's important to be ready with
the appropriate verbal match for your smile and that warm feeling in
your belly.

A good way to convey this sensation is by telling a person you
like them. Conveying this in Spanish, though, requires careful attention to differences in degree and intensity of the words. If you guide
yourself by the dictionary, for instance, you will think that \emph{querer}
means "to like." This is true---up to a point---but what a dictionary
often doesn't tell you is just how strong the "liking" reflected by
\emph{querer} really is. The distinction is important, since saying \emph{te quiero} to
someone whom you just "like" could provoke a considerably stronger
reaction than you bargained for. \emph{Te quiero} means "I love you"---in
some contexts even "I want you"---and is virtually synonymous with
the magic words \emph{Te amo}. Suppose someone asks, "What do you think
of my boyfriend?" If you try to say "I like him" with a response like \emph{Lo
quiero}, expect to be misunderstood: what you are really saying is "I
want him," as in "I want him for myself." Such slips of the tongue
could prove fatal.

So how do you say "Hey, you're not such a bad bloke" in Spanish? The simplest formula throughout the Spanish-speaking world is to
use \emph{caer bien}---literally "to fall well." \emph{Ese señor me cae bien} is the common, colloquial, and non-intimate way of saying "I like that fellow." It
is equally safe in direct speech. \emph{¿Sabes? me caes bien} = "You know, I
like you." This useful phrase will never fail to get your message across
loudly, clearly, and without confusion and misinterpretation.

But, you say, my dictionary also says to use gustar for "to
like," as in \emph{Me gustan las tortillas}. True---up to that point, again. Gustar with people is a slightly different matter. \emph{Me gusta Paco} does mean
"I like Paco," but it carries many of the connotations and commitments implied in \emph{Quiero a Paco}. "I fancy Paco" might be a good English equivalent. If you're male and you say \emph{Me gusta Paco}, expect to
receive funny looks. \emph{Paco me cae bien} is presumably what you want
to say.

As long as you're looking through the dictionary, you may also
discover \emph{agradar}. This verb does work for "I like you" (\emph{Me agradas})
without the romantic overtones, but it can sound a little forced or
formal.

Presumably your descriptions of others will occasionally go
beyond the fact that you like them or dislike them. You'll want to
describe, for the benefit of your listener, what kind of person Juan or
Juanita is. If Juan is a "nice guy," for instance, you could say \emph{es un
buen tipo, es buena persona}, or \emph{es buena gente} (closer to "he's good
people"). These work with Juanita as well, with one important exception: when referring to women, note that \emph{tipa} is derogatory. Thus \emph{Es
una buena tipa} sounds inherently contradictory---kind of like saying
"She's a nice bitch"---and should be avoided. Regional expressions
of general approval abound. In Mexican slang a likable person will often be described by saying simply \emph{Es buena onda}, roughly "He's (or
she's) cool."

Once you've mastered how to say you like a person, you'll
want to delve deeper and learn words for specific personality traits. For
the purposes of learning some of these, we'll divide positive personality traits into four entirely arbitrary groups: the \emph{amables}, or "nice"
people; the \emph{simpáticos}, or "cool" people; the \emph{listos}, or "sharp" people;
and the \emph{serios}, or "solid" people. Try to learn at least one description
from each, and you'll be well on your way to explaining why exactly a
given person "falls well to you."

Note as you go that all of these words, unless otherwise stated,
should be used with \emph{ser}. In fact, using some of these words with \emph{estar}
can lead to confusion and may change the meaning altogether. Where
this is the case, it is noted.

\section{The \emph{amables}}

Let's face it: some people are just downright nice. God knows
how they got that way or how they manage to stay that way, but the
fact is that they are kind, warm, and caring. And---why not admit
it?---we like them. On the other hand, they may never become our best
friends, and our dealings with them may never transcend the simply
social. In Spanish such a person could be called \emph{amable}, equating with
"kind." \emph{Amable} is a useful, generic, and somewhat bland word, perhaps most common in the stock phrase \emph{gracias, muy amable}. \emph{Gentil}
is used in much the same way, though it rings more formal. Note also
that \emph{gentil} is a trickster (see Chapter 3); it can mean "gentle," but it often covers much more.

Here are other commonly used favorable descriptions that fall
roughly in this category:

\subsection{\emph{generoso}}

%GENEROSO
Works safely as "generous."

\subsection{\emph{bondadoso}}

%BONDADOSO
Comes closer to "charitable" and "giving."

\subsection{\emph{desprendido}}

%DESPRENDIDO
A nice, multisyllabic word for "generous" in
the sense of "disinterested."

\subsection{\emph{atento}}

%ATENTO
Means "thoughtful."

\subsection{\emph{detallista}}

%DETALLISTA
Means "thoughtful," too, but goes further and is
used for the sort of people who send thank---you notes and tasteful, personalized gifts. It's the same in the masculine and feminine.

\subsection{\emph{cortés}}

%CORTÉS
Means "courteous" or "polite." A common maxim
in Spanish is \emph{Lo cortés no quita lo valiente}, which, very loosely translated, means something like "Real men can be polite, too."

\subsection{\emph{una persona considerada}}

%UNA PERSONA CONSIDERADA
A very safe cognate for "a considerate person."

\subsection{\emph{una persona comprensiva}}

%UNA PERSONA COMPRENSIVA
Similar to \emph{una persona considerada} but distinct, referring more to a very "understanding" or "compassionate" person.

\subsection{\emph{dulce}}

%DULCE
Means "sweet," a good word for some people.

\subsection{\emph{es un alma de Dios} and \emph{es un pan de Dios}}

%ES UN ALMA DE DIOS and ES UN PAN DE DIOS
Expressions
for extremely good-hearted people, of the sort who have never had a
harsh word for anyone.

\bsk

Also falling into the \emph{amable} category are a number of words
and phrases typical of polite society:

\subsection{\emph{una fina persona}}

%UNA FINA PERSONA
Suggests a kind, considerate person.
Turned around, \emph{una persona fina} is a well-bred, "fine" person.

\subsection{\emph{educado}}

%EDUCADO
Often used much like \emph{una persona fina}. Remember in passing that \emph{educado} has a far broader meaning in Spanish than
"educated" does in English (see Chapter 2). Sometimes \emph{educado} comes
closer to "classy" or "a class act" in English.

\subsection{\emph{una persona culta}}

%UNA PERSONA CULTA
Goes a step further, describing a dignified, tasteful individual.

\section{The \emph{simpáticos}}

This is the category that would include your best friends, and
the most descriptive word for them is \emph{simpático}. This word is also a
trickster, looking like "sympathetic" but meaning something quite different. A single English word doesn't really do it justice; probably you
would say something like "He (or she) is great!" The younger set might
find "cool" a close equivalent. Easier than translating the epithet is
imagining the sort of person who would be worthy of it: a happy,
friendly, attractive, charming, witty, altogether likable person.

About the only quirk to be noted about \emph{simpático} is its use
with the verb \emph{estar}, especially in reference to members of the opposite
sex and babies. In these contexts \emph{está simpático} can sound like a backhanded compliment. For babies, it would be like saying, "What an
interesting-looking child." For members of the opposite sex, especially
"eligible" ones about your own age, it's on a par with saying "Well, he
(or she) certainly has a good personality." Perhaps a parallel English
word would be "cute," in cases when it's understood that you're deliberately not using a more flattering word. If you're trying to be nice, use
simpatico with \emph{ser}.

If you want to pin down a \emph{simpático}'s personality further,
there is a wealth of words at your disposal:

\subsection{\emph{alegre}}

%ALEGRE
Covers "happy," "happy-go-lucky," "outgoing," and
the like.

\subsection{\emph{amigable}}

%AMIGABLE
A safe cognate for "friendly," but the Spanish
word is less common and more specific.

\subsection{\emph{atractivo}}

%ATRACTIVO
A straightforward cognate of "attractive." Like
the English word, it stresses the person's physical charms.

\subsection{\emph{encantador}}

%ENCANTADOR
A close fit for "charming."

\subsection{\emph{relajado}}

%RELAJADO
Describes a simpático who is "relaxed" or "laid-back."

\subsection{\emph{sociable}}

%SOCIABLE
Means "sociable" and is used to describe the outgoing, partying type.

\subsection{\emph{genial}}

%GENIAL
A useful word, though it almost ranks as a trickster.
It usually means "clever" or "great" when referring to things or ideas,
and "a character" or "a wild-and-crazy guy (or gal)" when referring to
people. It is usually a favorable assessment. For "genial," rely on \emph{amable} or \emph{amigable}.

\section{The \emph{listos}}

Some people rank high in our esteem because they are notably
"clever," "sharp," or "bright." Calling them that presupposes some intelligence on their part, but it's not the same thing as calling them
"smart" or "intelligent." "Witty" might come closer, in some cases. A
catch-all Spanish word for these people is \emph{listo}, used with \emph{ser} (with \emph{estar} it means "ready"). This is the word for people who appeal to you on
an intellectual level, people who enlighten, challenge, and entertain,
people you generally like to have around.
Not all bright and witty people are likable, of course, but the
ones who are can merit additional words:

\subsection{\emph{gracioso} and \emph{chistoso}}

%GRACIOSO and CHISTOSO
Stress the humorous aspect of the
person's personality.

\subsection{\emph{astuto} and \emph{ágil}}

%ASTUTO and ÁGIL
Emphasize sheer mental acumen.

\subsection{\emph{despierto}}

%DESPIERTO
A close fit for "bright," suggesting one part "brilliant" and one part "bright-eyed and bushy-tailed," "lively," "energetic." Used with \emph{estar}, of course, \emph{despierto} simply means "awake."

\section{The \emph{serios}}

Some people we like simply because they seem to be "good
people." They are honest, forthright, hard-working, and guileless. What
you see is what you get. These people tend to be our most esteemed
co-workers and our counselors in times of trouble, and the best word
for them in Spanish is \emph{serio}, used, as usual, with \emph{ser}. \emph{Es una persona
seria} = "He (or she) is a real straight-shooter" or "He (or she) has his
(or her) act together" or even is a "together person." \emph{Serio} does not
necessarily convey the idea of a droopy-faced soul or a rigid, unyielding
sort (although with \emph{estar} it can). Instead, it refers to a reliable, trustworthy,solid individual. \emph{Formal} means much the same, but is more, well, formal.

Since \emph{serio} is often used to describe the ideal qualities in a
professional colleague, many words in this category pop up frequently
in reference to the workplace:

\subsection{\emph{capaz}}

%CAPAZ
Means "capable." (Remember that \emph{capable} is a dangerous trickster; see Chapter 3.)

\subsection{\emph{trabajador}}

%TRABAJADOR
Means "hard-working."

\subsection{\emph{responsable} and \emph{dedicado}}

%RESPONSABLE and DEDICADO
These are our old friends "responsible" and "dedicated."

\subsection{\emph{cumplido}}

%CUMPLIDO
An extremely useful word for "reliable" or "competent." Saying \emph{Es muy cumplida} about someone means "She gets the job done."

\bsk

Some other qualifiers are slippery and should be used with care:

\subsection{\emph{muy vivo}}

%MUY VIVO
Sounds very complimentary---especially considering the alternative---but this somewhat slangy expression hints that
the person is "clever," "nobody's fool," and perhaps even a tad "shady,"
"underhanded" or "crooked." \emph{Despierto}, noted above, is a safer choice.

\subsection{\emph{movido}}

%MOVIDO
Suffers the same fate as \emph{muy vivo}. It refers to a very
active person---a real "go-getter"---but on the downside it can imply
that you're never really sure where this person is "moving," with
whom, or according to whose rules.

\bsk

Other "serious" words stress moral rectitude and solidness of
judgment:

\subsection{\emph{recto}}

%RECTO
Highlights the person's basic decency and honesty.

\subsection{\emph{honesto}}

%HONESTO
Includes "honest" but often covers a wider range
of moral traits; "decent" might be a better translation. For simply
"honest," \emph{honrado} is more accurate.

\subsection{\emph{una persona de confianza}}

%UNA PERSONA DE CONFIANZA
A common way to say a person can be trusted or confided in, either in the sense of personal honesty or competence.

\subsection{\emph{sensato}}

%SENSATO
Means "sensible" and is a good word to describe
someone's sound judgment and solid character.

\section{The bad}

All of us know that if you don't have anything nice to say
about a person, you shouldn't say anything at all. None of us practices
this belief, though, which is why this section is needed. For unless you
lead an extraordinarily blessed existence, you will probably find yourself in need of one or more of the following words sooner or later. After
all, even if you don't take pleasure in verbally thrashing your fellow
human being, you'll still need a few good words to describe those
who do.

Spanish, alas, is a rich language for belittling others. The range
of words, phrases, and expressions that can be employed to this end is
practically infinite and, furthermore, varies widely from region to region. Not all of these epithets fall within the range of dignified abuse,
of course, and some are so harsh and vulgar that they can reflect as
poorly on the person using them as on the person for whom they are
intended. For crude, harsh, and undignified abuse, flip ahead to
Chapter 10.

In this section, we'll concern ourselves with descriptions for
people whom you simply don't like. These people, in Spanish, can be
said to "fall badly to you"; \emph{te caen mal}. The converse of \emph{caer bien},
\emph{caer mal} is the safest and most universal way of expressing dislike in
Spanish. A watered-down version is \emph{no caer bien}. \emph{Pedro no me cae
bien} = "I not too fond of Pedro." A "watered-up" version is \emph{no caer
nada bien}. \emph{Pedro no me cae nada bien} = "I don't like Pedro at all." In
many countries, local and colloquial modifications of \emph{caer mal} have
been invented. In Mexico it is common to hear that so-and-so \emph{me cae
gordo}: "falls fat to me," literally. \emph{Me cae pesado}, or "he falls heavy to
me," is the same idea. Remember to use the correct gender depending
on who is "falling to you": \emph{él me cae pesado} but \emph{ella me cae pesada}.

As when expressing your likes, you are on shaky ground using
the verb \emph{querer} to convey dislike. \emph{No lo quiero} means "I don't love
him" or "I don't want him"---leaving a lot to the imagination of the
listener. \emph{Gustar} is also dubious. \emph{No me gusta} suggests "He (or she) is
not for me" or "is not my type."

To capture what it is about a person that "falls badly" to you,
you will need to enter the world of words for negative personality
traits. Again, we can divide them into four arbitrary groups: the \emph{pesados}, or "obnoxious" people; the \emph{imbéciles}, or "jerks"; the \emph{malvados},
or "mean" people; and the \emph{cochinos}, or "slobs." Remember that while
these words are less likely overall to provoke a brawl than their
four-lettered cousins in Chapter 10, you should still use them advisedly.
Pay attention to such intangibles as tone and situation. The same word
said lazily to a friend can offend or inflame a stranger. Unless otherwise noted, these epithets are best used with \emph{ser}.

\section{The \emph{pesados}}

This group covers the obnoxious boor, and \emph{pesado} is the word
we've chosen to carry the banner. \emph{Una persona pesada} can be anyone
from a nagger to an intolerable snob. What all \emph{pesados} have in common is that they get on our nerves after a while---in many cases after
a very short while. They are not really malicious; they're just underequipped in the personality department.

\emph{Pesados} tell crude jokes when no one wants to hear them,
make repeated passes at the same woman, use foreign words that nobody understands, drop names, sneer at your wardrobe, flaunt theirs,
gossip a lot, argue about everything, and never accept their mistakes.
Thank God we're not like them! In a lighter vein, you can use pesado
among friends or family to tell someone to behave themselves or to
"lighten up." \emph{No seas pesado}, said without much conviction, carries
the same message as "Don't be a pain."

Other words explore the \emph{pesado} phenomenon in more detail:

\subsection{\emph{pedante}}

%PEDANTE
For intellectually snobbish people (the ones who
use those foreign words). When we were kids, we called these people
"know-it-alls." Some people may refrain from correcting your Spanish
for fear of appearing \emph{pedante}. The word can be generalized to cover
anyone who has a higher opinion of himself or herself than others do.
"Stuck-up" comes to mind in English.

\subsection{\emph{presumido}}

%PRESUMIDO
The word you've been searching for to say
"show-off." Kids who stick their neat toys in your face but don't let
you play with them are \emph{presumidos}. Ditto for adults with fancy cars.

\subsection{\emph{prepotente}}

%PREPOTENTE
Implies a powerful person who abuses his or
her authority to the detriment of others. It's a good word for cops,
judges, politicians, and bosses, providing they don't overhear you. A
good English equivalent doesn't really exist.

\subsection{\emph{arrogante}}

%ARROGANTE
Safe for "arrogant," "excessively proud."

\subsection{\emph{vanidoso}}

%VANIDOSO
Means "vain" but is more commonly used than
its English equivalent.

\subsection{\emph{snob} or \emph{esnob}}

%SNOB or ESNOB
This term, for better or worse, is creeping
into Spanish. You may hear it, but try to use a Spanish equivalent to
avoid being taken for one yourself.

\subsection{\emph{creerse}}

%CREERSE
This verb, along with an appropriate adjective, will
deflate a pretentious person in a hurry. Someone who tells bad jokes
\emph{se cree chistoso}; someone who wears fancy clothes \emph{se cree elegante};
someone who uses foreign phrases \emph{se cree inteligente}; and so on. \emph{Se
cree mucho}, by itself, covers the lot of these people. In other words,
they believe themselves to be "great shakes," but this opinion is not
widely held.

\bsk

Regional words and slang expressions to cover snootiness are
frequent and often the most colorful. In Mexico \emph{sangrón} is widely if
somewhat slangily used to convey "obnoxious." In Mexico, too, you
can say of a snob, \emph{Le echa mucha crema a sus tacos}, or "He (or she)
puts a lot of cream on his (or her) tacos." Once I heard, in reference to
a particularly stuck-up young man, that \emph{se cree la última Pepsi del
desierto} ("he thinks he's the last Pepsi in the desert"). Keep your ears
open, wherever you are, and you should be able to add fun new descriptions to the stock you acquire here.

\section{The \emph{imbéciles}}

As a rule, none of us likes stupid people. It's not a matter of
their I.Q.---in fact, they may be quite intelligent---but we call them
"stupid" nonetheless. These are people who do stupid things and make
stupid comments. Sometimes they make us wonder how on earth anyone can be so downright, undeniably, irretrievably STUPID!
When you reach that point in your regard of a fellow human
being, the word you want is \emph{imbécil}. In truth, \emph{estúpido} exists and is
widely used as well, but it lacks the punch that \emph{imbécil} packs. Say it
with heavy, accentuated, spitting scorn on the middle syllable: \emph{im-BE-cil}. Now throw your hand in the air as you say it, as if casting this
person from the planet: \emph{¡im-BE-cil!} Isn't this fun?
\emph{Imbécil}, like "stupid," doesn't necessarily mean that the object of your scorn is unintelligent. This is a good thing, because it
means you can use it on an even wider range of people: college graduates, college professors, or even college presidents, if you like. An English equivalent might be "jerk" or perhaps "stupid jerk" or even "You
stupid jerk!" depending on the amount of spit and scorn you want to
add. It is a multipurpose, and strong, pejorative. I once watched as an
unseemly man pestered a young, well-dressed blonde woman on the
subway in a major Latin American city; taking her for a tourist, he
kept asking her "Whey a you from?" in barely pidgin English. Finally
the woman got fed up with this harassment, turned on the man, and
spit \emph{De aquí, ¡imbécil!} in his face. The man slinked off, humbled, and
the crowd was delighted. \emph{Estúpido} just wouldn't have worked.
The \emph{imbécil} category includes a number of other useful descriptions for those who make a distinctly bad impression on us:

\subsection{\emph{idiota}}

%IDIOTA
Similar to \emph{imbécil} but perhaps a shade weaker. Still,
native English speakers should be warned that this word is not used
nearly as casually in Spanish as in English---as in "Oh, don't be an
idiot" or "What a blunder! I feel like an idiot!" Same in masculine and
feminine: \emph{él es un idiota, ella es una idiota}.

\subsection{\emph{infeliz}}

%INFELIZ
Less a "jerk" and more a "klutz" or a "schmuck." It
is used to describe a sort of hapless clod whose luck is mostly bad, at
least in part because of a lack of willpower to make it better. An \emph{infeliz}
is not necessarily a scoundrel or even an entirely unpleasant person,
but you probably wouldn't want to have one as a friend. To soften the
blow, qualify it with \emph{pobre}. \emph{Un pobre infeliz} = "a poor sap," "a loser," a real "sad sack."

\subsection{\emph{tonto}}

%TONTO
Works as "silly" or a watered-down "fool."

\subsection{\emph{payaso}}

%PAYASO
Means "clown" and is a stronger word for "fool,"
particularly a clumsy, oafish one.

\subsection{\emph{baboso}}

%BABOSO
Less harsh than payaso but no less clear in its implications (it comes from \emph{babear}, "to slobber").

\subsection{\emph{tarado}}

%TARADO
A handy word suggesting that something crucial to
cerebral functioning may be missing; "moron" might fit.

\subsection{\emph{tarugo}}

%TARUGO
A term that means "a block of wood." Applied to
people, it equates nicely with "blockhead."

\subsection{\emph{burro}}

%BURRO
Meaning "donkey," the word calls to mind "dumbbell" or "dunce" and, like those words, is a favorite among schoolkids.

\subsection{\emph{bobo}}

%BOBO
In the same class as burro, suggesting "dummy."

\subsection{\emph{ignorante} and \emph{cretino}}

%IGNORANTE and CRETINO
Safe cognates that are heard on occasion.

\subsection{\emph{simple}}

%SIMPLE
Mostly a false cognate, especially when talking
about people; it means "simple-minded" or "simpleton" more than
"simple." If "simpleton" is the word you seek, \emph{simplón} is even more
expressive.

\subsection{\emph{torpe}}

%TORPE
Means "clumsy oaf" or "klutz."

\subsection{\emph{necio}}

%NECIO
A useful word that refers more to stubbornness than
stupidity, though the two often go hand in hand; "jackass" may be a
good equivalent.

\section{The \emph{malvados}}

In this group we lump people whom we consider mean, offensive, malicious, or simply "bad" people. \emph{Malvado} is a good catchall to
describe these people, especially if they have done us some harm. \emph{Maldito} is stronger, though without being crudely so, and is invariably the
word chosen in subtitled films to translate "bastard." (In Spanish \emph{bastardo} is usually reserved for persons born out of wedlock and thus
works poorly as a general-purpose insult.) Here are some other useful
descriptions of this type:

\subsection{\emph{mal parido} and \emph{mal nacido}}

%MAL PARIDO and MAL NACIDO
Meaning literally "born bad,"
these phrases convey the same ignominy as \emph{malvado} but sound a bit
snazzier.

\subsection{\emph{sinvergüenza}}

%SINVERGÜENZA
Literally meaning "without shame" or
"scoundrel," the word suggests that this person actually enjoys being
offensive. Very commonly used. Same form for masculine and feminine.

\subsection{\emph{canalla}}

%CANALLA
Refers generally to men, often a man who mistreats women. It works well for "lout." 1I Same form for masculine and
feminine.

\subsection{\emph{desgraciado}}

%DESGRACIADO
A common term of generic opprobrium, generally meaning an unpleasant person who has tried to do us some
wrong. It also implicitly tries to write the person off as trivial and unworthy of genuine scorn. It equates fairly well with "wretch" or "sap"
in English.

\subsection{\emph{infame}}

%INFAME
A good strong word used to describe someone who,
by his or her actions, has earned everlasting infamy---at least in our
eyes. Murderous tyrants are \emph{infames}; so, by extension (and slight exaggeration), is anyone who causes you genuine harm with malicious
intent.

\subsection{\emph{mentiroso}}

%MENTIROSO
A strong word, stronger even than "liar." It suggests the person is a habitual liar and uses lies to defame and gain advantage. Be careful using it.

\subsection{\emph{maléfico}}

%MALÉFICO
Describes a real evil sort, a pernicious "ne'er-do-well."

\subsection{\emph{malviviente}}

%MALVIVIENTE
Literally meaning "bad-liver" (one who lives
badly), the word suggests an incorrigible rogue with a long police
record.

\subsection{\emph{delincuente}}

%DELINCUENTE
Much the same as \emph{malviviente}, the term
means "delinquent" but stresses the criminal lifestyle that such a person has chosen. As a result of increasing drug use and the crime that
goes with it, words like \emph{mariguano} are coming into vogue to mean the
same thing. To an English speaker the word calls to mind a fading hippie, but in Spanish it describes a dangerous "druggie" and is becoming
the common word for scandalous and vandalous teenagers.

\section{The \emph{cochinos}}

For uncouth, classless slobs, \emph{cochino} is the word. It is one of
four common Spanish words for "pig" (the animal)---\emph{cerdo, puerco},
and \emph{marrano} are the others---and all work well for the human equivalent. If you dislike someone because he or she throws trash out of car
windows, takes up three train seats, or burps at the dinner table, this is
the category you should check. (You should be aware that the word
\emph{marrano} has a long anti-Semitic history, though it is rarely used that
way today.)

\subsection{\emph{grosero}}

%GROSERO
The catchall word for "rude." \emph{Se portó bastante
grosero} would be the common way of saying "He acted quite rude,"
"He treated us badly."

\subsection{\emph{vago}}

%VAGO
Describes an unkempt person or "bum." (It is also the
word for "vague"; see Chapter 3.)

\subsection{\emph{bajo}}

%BAJO
Widely used in the phrase \emph{un tipo de lo más bajo}, a
strong denunciation that translates well as "a low-life," "creep," or
even "scum." Note, however, that \emph{Es un tipo bajo} usually just means
"He's a short guy."

\bsk

A number of other common words refer to general "low-class"
behavior. Though most modern-day English speakers are not accustomed to think of behavioral traits in class terms, in much of the
Spanish-speaking world the class distinction is still a strong one.
Words like \emph{corriente, vulgar}, and \emph{ordinario} denote little more than
"common" or "characteristic of the masses," but all three connote
"uncouth," "slobbish," or "classless" (as in "he has no class") when
used to refer to people. \emph{Un hombre corriente} can be expected to meet
his dinner guests unshaven, in boxer shorts and scratching his paunch.
\emph{Inculto} falls into this group as well. It means "uncultured" but is
much broader and more frequently heard than its English cognate; an
\emph{inculto} not only doesn't appreciate Mozart but plays his radio too loud.

\section{The indifferent}

We have looked at the human race from both sides now. We
have met the lowest of the low, the snobs and scoundrels, and learned
what to call them. We have also been introduced to the creme de la
creme, the well-bred gentlefolk, and have had a few charmed words
with them. But what about the rest of the world's inhabitants---that
gray, faceless mass that hovers uncertainly between good and evil?
What do you say to describe a "nowhere type," a fair-to-middling sort,
a person who's nothing to write home about and just like the next guy,
only less so? What do you say about people when there isn't much to
say about them one way or the other?

A number of unspectacularly adequate Spanish words rise
humbly to the task. \emph{Regular, normal}, and \emph{común} all make the grade,
though you should be alert to their idiosyncrasies. \emph{Regular} with \emph{estar},
for example, is a notch or two below "regular" on the scale of descriptive tags; whereas in English it means "average," in Spanish it is closer
to "fair" or even "not so hot," especially when referring to objects.
\emph{Esta regular el camino} does not mean "It's an average road" but rather
"It's a pretty lousy road." "Regular" in the sense of "consistent" is
\emph{regular} with \emph{ser}. Be alert to the distinction: \emph{Es un cliente regular}
would be "He's a steady customer," while \emph{Ese cliente está regular} suggests the customer is a lousy tipper or otherwise "fair." Using \emph{asiduo}
instead of \emph{regular} to describe the steady customer can avoid a potential
misunderstanding and win you brownie points for using a good Spanish word.

\emph{Normal} is the best all-around word for the "nothing-special"
category. \emph{Mediocre} carries a clear connotation of insufficiency, as does
its cognate in English. Other words---specifically \emph{corriente} and \emph{ordinario}---are listed in dictionaries as synonyms for "normal" or "ordinary,"
but in fact can be charged with negative meaning when used to label
people, as noted above. \emph{Común} works in the expression \emph{un hombre común}, meaning "common folk" or "the man on the street." \emph{Común y
corriente} is a good translation for "run-of-the-mill," without the negative connotations of \emph{corriente} by itself. Still, your safest bet in almost
every circumstance involving people is normal.

In situations where physical appearance is being commented
on, a different set of words takes over. A person whose looks are "nothing special" could be described as \emph{pasadero} or \emph{pasable}, meaning "acceptable" or "tolerable" in the sense of "if nothing better comes
along." A useful, if regional, slang phrase for "not bad" (or "not great")
is \emph{dos tres}; it works equally well for people, places, and things and
sounds, to my ear, a lot better than \emph{así así}, which every textbook will
tell you means "so-so." \emph{Más o menos} (technically, "more or less") is
also commonly used in Spanish, even when there's no indication of
what is being compared: \emph{¿Qué tal la película? Más o menos}. = "How
was the film?" "Not bad." Sometimes in slang \emph{más o menos} gets
shortened to \emph{ma' o meno'} in this usage.

Some fun expressions for saying "nothing special" employ neither-nor constructions. A Peruvian, for instance, might say someone is
\emph{ni chicha ni limoná}, where \emph{chicha} is a local alcoholic drink and \emph{limoná} is "lemonade." The idea is that the person is neither alcoholic
nor nonalcoholic-that is, isn't anything well-defined. "Neither here
nor there" comes out as \emph{ni de aquí ni de allá}. \emph{Ni pinta ni da color}
means that someone (or something) "neither paints nor adds color,"
suggesting that there is no real reason for this person or thing to exist
at all.

Another common way of saying "no great shakes" is \emph{nada del
otro mundo} (literally, "nothing from the other world". Another handy
phrase is \emph{Es cosa de cada domingo} ("It's an every-Sunday thing"),
equating with "a dime a dozen" and showing that you are distinctly
unimpressed. \emph{No es nada fuera de lo común} is good but a bit stilted
for "It's nothing out of the ordinary." When you are completely unimpressed by something or someone, but still unwilling to turn it or
them down, resort to \emph{peor es nada}---the correct rendering for "better
than nothing."

\section{Temporary states}

Since we've spent so much of this chapter name-calling, it's
only decent of us to finish by noting that most bad traits are only
temporary and that most people, if we'd only give them time and get
to know them, would soon return to being their real sweet, lovable
selves. With this is mind, you'll need the proper vocabulary to describe
these warm and caring individuals who just happen to be acting like
miserable worms at the moment.

One way to get this across is by using \emph{estar}, the Verb of Temporary States, but not simply as a substitute for \emph{ser}. Instead, by inserting \emph{de} between estar and your adjective, you change "He's an S.O.B.
(or whatever)!" to "He's being an S.O.B.!" Thus \emph{Es un grosero} describes
a permanently rude, coarse, and unpleasant man; \emph{Está de grosero} describes a man who is acting rudely. \emph{Es un presumido} = "He's a showoff"; \emph{Está de presumido} = "He's showing off."

This construction is especially useful for capturing moods. For
instance, \emph{Está de pesado} = "He's in a lousy mood." Two other key
phrases for moods employ the generic terms \emph{buenas} and \emph{malas}. \emph{Está
de buenas} means someone is "in a good mood." \emph{Está de malas} is the
dreaded opposite.

\emph{Estar de} can also be used instead of \emph{ser} for ironic effect. For
instance, \emph{Está de generoso} suggests that someone who usually isn't
generous is for some reason suddenly giving away the store. Saying
that a child \emph{está de obediente} suggests that the child's obedient behavior is indeed most unusual. \emph{Estar de} can likewise be used to explain
away behavior as a temporary aberration: \emph{Estoy de tarado} = "I'm acting a little stupid." It should be noted that some adjectives by their
very meaning resist the \emph{estar de} construction; it is hard, for instance,
to imagine someone being \emph{infame} for only a few minutes or so.

\chapter{The Secret Life of Verbs}

Many students of Spanish can still recall their first encounter
with Spanish verbs. Often the reaction was something like, "What do
you mean it has different endings in every tense? What on earth for?
Aaaarrgh!"

Alas, for verb endings there is little alternative but to buckle
down and memorize them. But even when you've managed to separate
\emph{-aban} from \emph{-aron}, another unappetizing task awaits: using each ending
and each tense and mode in the right situation. To English speakers
the idea of an imperfect tense is unfamiliar, while the notion of a frequently used subjunctive mode can seem downright perverse. Can't
verbs be made just a little easier?

The answer is yes. There are several pointers on the use of the
tenses that will make learning them a less doleful task. And there are
a couple of obvious pitfalls that anyone wandering into the world of
tenses should be alert to. This chapter presumes you have some background knowledge on the use of tenses in Spanish; even so, I'll try to
summarize some of the basics of tense usage to refresh your memory.

\section{The present}

The present is the most straightforward of tenses in Spanish
and corresponds almost perfectly to the present tense in English: \emph{bailo}
is "I dance," \emph{estoy bailando} is "I am dancing," and so on. For the English compound present ("I do dance"), in Spanish you can slip a \emph{sí}
("yes") into the phrase. "He does eat a lot" = \emph{El sí come mucho}. "But,
honey, I do love you" = \emph{Pero cariño, sí te quiero}. Note that in English
the present progressive is used far more than in Spanish. Thus "She's
going" should almost always be rendered \emph{Se va} and only rarely \emph{Se esta
yendo}.

One tip on the present tense: it is used much more in Spanish
for future events than its English equivalent. For instance, the common way to ask "Are you coming tomorrow?" is simply \emph{¿Vienes mañana?} "We'll see you later" often gets rendered \emph{Nos vemos}---literally,
"We see each other." This trick works even when you're specifying a
future time or date. \emph{Te lo doy el martes} = "I'll give it to you Tuesday."
In Spanish the context makes it clear that the future is being referred
to. Compare these examples:

\bsk

\indu \emph{Te cuento}. = "I'll tell you (now)."

\indu \emph{Te contaré}. = "I'll tell you (someday)."

\indu \emph{Mañana te cuento}. = "I'll tell you tomorrow."

\indu \emph{Mañana te cantaré}. = "I shall tell you tomorrow."

\section{The future}

The future is not a particularly hard tense to learn, and when
to use it is pretty obvious. Still, it can be simplified considerably by
remembering the following rule: ignore it.

What? Just forget about the future? Well, maybe not altogether. There are a few uses for the future in common spoken Spanish.
But most of the time you can avoid it outright, and you'll even sound
more fluent by doing so.

The two most frequent substitutes for the future tense in
Spanish are the "present-as-future," discussed above, and the compound future using the verb \emph{ir} ("to go"), just as in English. \emph{Mañana voy
a llamar a mi hermano} = "Tomorrow I'm going to call my brother."
In common Spanish usage, one of these two constructions almost always replaces the future tense, although the future can be used and is
certainly understood: \emph{Mañana llamaré a mi hermano}. Sometimes,
though, the future sounds formal, stiff, and even awkward. \emph{Vuelvo en
seguida} = "I'll be right back." \emph{Volveré en seguida} = "I will return
promptly."

One special use of the future that anyone aspiring to fluency
must learn is what is called the "future of uncertainty." It is a very
common construction that has no real equivalent in English---which is
why many students shudder at the mere thought of it. The best way to
get a grip on it is by example. You often hear it, for instance, when
someone knocks unexpectedly on the door. "Who could that be?"
you'd blurt out in English. \emph{¿Quien será?} you'd say in Spanish. Other
examples: \emph{¿Habrá más?} = "Might there be more?" \emph{¿Estará en casa?} =
"Do you suppose he's at home?" And so on.

The future of uncertainty, applied to the past, uses the compound future perfect tense. That sounds difficult, but really it works
out to sticking \emph{habrá} or an equivalent in front of the past participle of
your choice. \emph{¿Quién habrá sido?} ("Who might that have been?"), you
might wonder on hearing a strange voice on your answering machine.
\emph{¿No habrán querido asustarme?} = "You don't suppose they wanted to
scare me, do you?"

\section{The conditional}

The conditional is a very predictable tense, conveying
thoughts in Spanish that in English rely on the auxiliary "would." It's
an easy tense to understand intuitively. "I'd like to eat now" is expressed as \emph{Me gustaría comer ahora}. "You would have liked it" = \emph{Te habría gustado}."

One special warning applies for the conditional: note that in
English we occasionally use "would" for repeated past actions, as in
"His father would eat every night at seven sharp." This construction
in Spanish calls for the imperfect---in fact, it's a textbook example of
when the imperfect is needed: \emph{Su padre cenaba todas las noches a las
siete en punto}. If you think about it long enough, it becomes clear why
this is not a true conditional, in either English or Spanish. But if you're
simply translating your "woulds," it's easy to make this mistake.

\section{The preterit versus the imperfect}

This face-off is one of the trickiest in Spanish, mostly because
in English we often gloss over the distinction. We do have the progressive past, of course, in such constructions as "He was flying a kite,"
but not all "ongoing" past activities use it. In many cases, you'll have
to slow your translating computer down a few megas and think about
exactly what kind of past action you are describing.

Basically the Spanish imperfect covers two constructions:

\subsection{}

\textbf{The progressive past}, including actions that are taking
place over a period of time in the past, usually in relation to some
other action that happened suddenly. For example, "He was sleeping
when the alarm went off." Or "She was overseas when the new president was elected." Keep your eye (and your mind's eye, if you're speaking)
out for these juxtapositions; they represent one of the most common uses of the imperfect. Because of the explicit juxtaposition, these
are also the easiest situations to recognize as an opportunity for the
imperfect.

\subsection{}

\textbf{Actions that took place over a period of time in the past},
often keyed to the English constructions "used to" and, on occasion,
"would" (see "The Conditional" above). This second common use of
the imperfect is indispensable in describing the way things were: "I
used to work nights" is expressed as \emph{Yo trabajaba de noche}. The key
words aren't always present, though. "When I was a kid, the teachers
beat the students" would be \emph{Cuando yo era niño los maestros golpeaban a los estudiantes}.

The problem of the disappearing key words can be seen in the
following sentences. Just as you could say "My father ate at seven every night," "My father used to eat at seven," or "My father would eat
at seven," you could just as easily state, pure and simple, "My father
ate at seven." Grammatically this last example is correct, but it is confusing without a clarifying context. Did he eat at seven o'clock just
once, or did he always eat at seven? Spanish lets you make the distinction in the verb itself: \emph{cenó a las siete} is clear in communicating that
he ate at seven on a certain occasion; \emph{cenaba a las siete}, using the
imperfect, indicates that it was his custom to eat at seven night after
night.

Understanding this distinction will clear up a lot of the conflicts between imperfect and preterit, but incorporating that knowledge
into your storytelling skills will take some time and practice. Often,
when relating a story, you'll have to jump nimbly back and forth from
imperfect to preterit, and this requires analyzing each action as it
pops up.

Let's say this father usually ate at nine, but on one particular
night he ate at seven; while he was eating, he found a fly in his soup
and fainted. What tenses will make this meaning clear in Spanish?
First, we have to explain that the father used to eat or usually ate (\emph{cenaba}) at nine; on the night in question, though, he ate (\emph{cenó}) at seven.
While he was eating (\emph{cenaba}---the imperfect again), he found a fly in
his soup (\emph{encontró una mosca en la sopa}---preterit) and fainted (\emph{se desmayó}---also preterit).

Sometimes, the differences between past and imperfect are
subtle to the point of near invisibility. In these cases, native Spanish
speakers will sense the distinction, but they have a hard time explaining it you. Let's say you answer the phone and your boyfriend is on the
other end. He could ask, \emph{¿Qué estabas haciendo?} Or he might ask,
\emph{¿Qué estuviste haciendo?} What's the difference? In the first example,
using the imperfect in the auxiliary, he is asking what you were doing
in relation to a sudden action---presumably, the ringing of the phone.
That is, "What are you doing right now (besides talking on the telephone)?" In the second example, he is asking what you were doing,
with the idea that you may have finished by now. "What have you
done all day?" or "What have you been up to?" is closer. Of course, if
he wanted to be completely clear, he could ask \emph{¿Qué hacías?} or \emph{¿Qué
hiciste hoy?} or \emph{¿Qué has estado haciendo?} and so on. But where's the
romance in that?

The important thing in these borderline cases is not to learn
the "right" form. (In fact, your boyfriend would probably just ask \emph{¿Qué
haces?}) The important thing is to search for the distinctions in these
close cases to get a feel for past tenses that will serve to guide you
when no key words or juxtapositions are there to help you along.

Let's take a final borderline example. You were at home last
night. But do you use the imperfect or the preterit to convey that to
your listener? Again, it depends. Both could be correct, but \emph{estaba en
casa} suggests you were there for the duration and something sudden or
specific happened during this time: \emph{Estaba en casa cuando se fue la
luz} ("I was at home when the lights went out"). The juxtaposition may
be implied, not stated, but you create the expectation of one by using
the imperfect. In other words, \emph{Estaba en casa anoche} in effect prompts
the question "And what happened?" \emph{Estuve en casa} is a much more
basic, self-contained, declarative sentence. It just says you were at
home, period. In this sense, it's a safer, more all-purpose choice than
\emph{estaba}.

\section{Special cases}

In Spanish the tense of certain verbs is almost as important as
word choice in getting your point across correctly, and students of the
language are generally unaware of the subtle twists they can give these
verbs by their choice of tense. Yet by manipulating the tenses well,
you can discover ways to express familiar English phrases in correct,
colloquial Spanish.

For instance, \emph{querer} ("to want") changes its meaning on the
trip from preterit to imperfect. Consider the case of \emph{quisieron} (preterit)
and \emph{querían} (imperfect). Both mean "they wanted," but a native Spanish speaker hears a difference. The former suggests that they wanted to
do something (and they did it). Thus their "wanting" came to an end,
at least for a while, so the verb goes into the preterit as a done deal.
\emph{Querían}, the imperfect, suggests they wanted to do something and,
evidently, they still want to. That is, they wanted but were unable to
do something. Thus, \emph{Quisieron ir al cine} and \emph{Querían ir al cine} both
mean "They wanted to go to the movies," but with a difference. The
preterit conveys the idea that they wanted to go the movies, so they
went. The imperfect suggests that they wanted to go the movies but
didn't---maybe after seeing the ticket prices.

The imperfect can also be used to say that they wanted to do
something and perhaps eventually did do it, but not before something
else intervened making that possible. For example, \emph{Querían ir al cine
y los mandé}, ("They wanted to go to the movies, so I sent them"). The
use of \emph{quisieron} implies that they would have gone without waiting
for someone to send them. As in the earlier example about being at
home,an imperfect on its own almost seems to raise the question \emph{¿Y
qué pasó?} It is, so to speak, the tense in which the other shoe is always
waiting to drop.

In the negative, the preterit-imperfect distinction is sharper
still. \emph{No quería ir a la fiesta} implies "I didn't want to go to the party
(but I went along anyway)," perhaps to avoid offending the host. \emph{No
quise ir a la fiesta} says bluntly "I didn't want to go to the party (so I
didn't)." Very often, \emph{querer} in the preterit and negative translates best
as "to refuse." \emph{No quisieron dar sus nombres} = "They refused to give
their names."

Similar changes occur to the verb \emph{poder} ("to be able") when it
makes the switch from preterit to imperfect. \emph{Mi hermano podía romperme la cabeza} suggests "My brother had the physical force necessary
to bust my head," while \emph{Mi hermano pudo romperme la cabeza} suggests that on the occasion in question he put this theoretical force to
the test. In the preterit, in other words, \emph{poder} means you not only
could do something but actually did do it. \emph{El podía nombrar todas las
capitales del mundo} ("He could name all of the world's capitals")
might be said of a smart lad. \emph{El pudo nombrar todas las capitales del
mundo} means the tyke was actually put to the test---and succeeded
("He was able to name all of the world's capitals").

As with \emph{querer}, the distinction is sharper in the negative. \emph{No
podía caminar} might mean that your feet hurt after a long hike and
you spent a day cuddled up on the couch watching sitcoms. \emph{No pude
caminar} suggests you actually tried to walk and fell on your face. In
general, the preterit with \emph{no poder} implies a fairly formidable obstacle
and suggests that an unsuccessful effort was at least made. The imperfect with \emph{no poder} is much less definite and does not hint that an effort was made. To take an extreme example, \emph{No podía volar del techo
de la casa} is stating the obvious: "I wasn't able to fly off the roof of my
house." \emph{No pude volar del techo de la casa} are words that, if uttered
at all, would most likely be uttered in an ICU ward by a sheepish lunatic---that is, after trying to fly and failing.

The last common verb whose meaning varies significantly
between the imperfect and the preterit is \emph{saber} ("to know"). In fact, this
variation causes students of Spanish no end of confusion. It seems unjust that such everyday statements as "I didn't know that!" and "Did
you know he was coming?" require a brain-racking choice of tense.

In fact, with \emph{saber} the difference is usually quite clear-cut.
Here's a rule of thumb that works well most of the time: use the imperfect. In the preterit \emph{saber} usually means "to find out." Some examples are in order. \emph{No sabia eso} = "I didn't know that." \emph{Sabia que
llegarías} = "I knew you'd show up." \emph{¿Sabías hablar español cuando
llegaste aquí} = "Did you know how to speak Spanish when you got
here?" In the preterit, in contrast, \emph{saber} generally refers to sudden
knowledge about a specific event. \emph{Supe que habías llegado} = "I heard
(I found out, word reached me) that you had arrived." \emph{¿Supiste\ldots{} ?} in
particular is almost always used for "Did you hear\ldots{} ?" and implies
some new gossip, revelation, or fast-breaking news. \emph{¿Supiste que gané
la lotería?} = "Did you know (hear) that I won the lottery?"

\section{\emph{ser} versus \emph{estar}}

In Spanish the question is not so much "to be or not to be?"
but "to be (\emph{ser}) or to be (\emph{estar})?" These two verbs are the source of
constant headaches and frequent errors for even intermediate and
advanced students of Spanish. Native Spanish speakers intuitively
choose the correct form without so much as a thought. You should
be so lucky.

One modern Spanish dictionary, in its introduction, makes
this very point (and rather smugly, I thought): "[Foreigners] should
know, so that they realize that the distinction between ser and estar is
clear and precise and that it is just a matter of managing to penetrate
the distinct nature of both verbs, that Spaniards, even the most uncultured ones, never use them wrong."

As a foreigner, of course, you will use them wrong, and about
10 percent of the cases will still seem mystifying to you even years
after you learn the common usages. But in at least 90 percent of the
cases the distinction between \emph{ser} and \emph{estar} is "clear and precise"---or
at least pretty easy to guess. As for that other 10 percent, well, you
gotta leave something to learn as you get older!

\section{The easy ones}

\emph{Ser} is the verb "to be" for things that are That Way, period.
They're not that way in relation to something else, or at certain times
of day, or in the spring or the fall, or only in election years. They are
that way because they were born that way and they will presumably
remain that way until the day they die. Ser is a solid, upstanding
verb---one that you can rely on to give you the same answer time
and time again.

\emph{Estar}, in comparison, is a flake. It is the variable, flighty, here-
today-gone-tomorrow verb "to be." \emph{Estar} covers personality traits that
are ephemeral and ethereal. It describes things that change from one
minute to the next. It's an all-over-the-place, outta-control kind of
verb. It's untrustworthy. It's slippery. You would never buy a used car
from a verb like \emph{estar}.

Let's take an example. Say your boss is a fool. \emph{Es una tonta},
you might say (though perhaps not to her face). But let's also say that
she spent all morning collecting mud samples and is now absolutely
filthy. \emph{Es una tonta and esta mugrienta}. What's more, in a moment of
inspired honesty, you told her that she looked like something that just
crawled out from under a rock, and now she's mad \emph{está enojada}. Just
like her to get so upset about a casual observation, you think. She's so
sensitive---\emph{es tan sensible}.

As you can see, we're getting a good picture of your boss: \emph{es
una tonta} and \emph{es muy sensible} (all the time), and \emph{esta mugrienta} and
\emph{esta enojada} (this afternoon). Probably not the best time to be kicking
back and reading a book, come to think of it.

Some words flat out change in meaning depending on whether
they are governed by \emph{ser} or \emph{estar}, and they can help us "penetrate"
those "distinct natures" we've been told so much about. Here's a rule
that can be applied in most cases: if you can add a "now" or "at the moment" to your description, you should be using \emph{estar}. If not, leave it to
\emph{ser}. \emph{Es un borracho}, for instance, means that someone is a "drunkard"---
a habitual drunk or a wino. \emph{Está borracho}, on the other hand, means
"He is drunk (at the moment)." \emph{Es callado} refers to a man who is
"quiet," not at any given moment but as a way of life---it's his nature;
he is a person who keeps to himself and speaks softly and rarely. To say
of another man \emph{está callado} means something quite different: he is
quiet---now. We are given no insight to his overall personality; we just
know that in this place and at this time, he's keeping his mouth shut.

Learning and reviewing examples is a good way to absorb the
essential difference between \emph{ser} and \emph{estar}. But a few other specific tips
may be helpful as well.

\subsection{}

Use \emph{ser} for general, permanent physical appearance: tall,
dark, handsome, short, light-skinned, ugly. (An exception will be dealt
with in a moment.) Use \emph{estar} for any temporary physical condition:
pale, flushed, disheveled, unshaven, and so on.

\subsection{}

For quantities, numerical or otherwise, always use \emph{ser}: \emph{somos veinte personas, es mucha, era poco, son dos}.

\subsection{}

For possession, use \emph{ser}: \emph{es mío, es de él, son de las señoras,
son suyos}.

\subsection{}

Location is always the province of \emph{estar}. This might throw
you if you think of the location of, say, a building as fairly unchanging.
But location is an implicit recognition of an object's relation to other
things---not a reflection of its indelible self---and thus is a job for
\emph{estar}.

\subsection{}

With all adverbs, adverbial expressions, and present participle forms or gerunds (the "-ing" form), use \emph{estar}: \emph{está bien, están en
buenas condiciones, está lloviendo, estoy nadando.}

\subsection{}

With all nouns, use \emph{ser}. If you have trouble recognizing
nouns, a good device is to key on the presence of the indefinite article
(\emph{un, una, unos, unas}). Thus \emph{es un doctor, eres una tonta, es un santo},
and so on. When the article is missing, as it often is in Spanish (that is,
\emph{es doctor}), you'll just have to remember that it could be used in that
situation and therefore it's a noun that requires \emph{ser}.

\section{Getting tricky: the past participles}

%GETTING TRICKY: THE PAST PARTICIPLES
With past participles (the "-ed" form in English, the \emph{-ado} and
\emph{-ido} forms in Spanish), things start to get tricky. Both \emph{estar} and \emph{ser}
can be used, but they mean different things. With \emph{estar} the participle
is generally being used as an adjective and to describe a passing state.
\emph{Estaba agotado} means "He was worn out (at that moment)."

With \emph{ser} the past participle is generally used to form a passive
construction or a predicate noun. In the case of the passive, you implicitly ask (and often must explicitly state) "whodunnit?"---that is,
who or what caused the action. \emph{Fue agotado}, for instance, means "He
was worn out," meaning something or someone wore him out. A few
past participles are used with \emph{ser} without any explicit causal agent, including \emph{conocido, sabido, tardado}, and \emph{parecido}.

Some examples may help clarify the distinction between \emph{ser}
and \emph{estar}. Say you went on an expedition to a remote patch of rainforest. When you got there, though, you found that it had recently been
bulldozed. On your return, someone may ask, "How was the forest?"
You could reply using either \emph{Estaba destruido}, referring to its destroyed state, or \emph{Fue destruido}, meaning essentially "It has been destroyed" and calling attention to the fact that someone or something
destroyed it. In English both senses can be covered by "It was destroyed." Spanish makes a finer distinction.

Another example: you can say both \emph{Las tiendas son cerradas a
las nueve} and \emph{Las tiendas están cerradas a las nueve}. What's the difference? With \emph{ser} you are saying that the stores are physically closed
by someone at nine 'o clock sharp. That is when the doors are shut and
the keys turn in the locks. With \emph{estar} you are saying that if you go to
the commercial district at nine you will find the stores closed. They
may have been closed at eight, or at six, or at five minutes to nine, but
in any case you will find that they are closed at nine. In other words,
\emph{Son cerradas a las nueve} = "They close at nine"; \emph{Están cerradas a las
nueve} = "They are closed by nine."

Finally, an example you will want to study assiduously if you
are of the married persuasion: \emph{soy casado} versus \emph{estoy casado}. Some
will argue that there's no big difference here. Others will say there's a
world of difference. Basically, \emph{soy casado} is "I'm a married man." It
describes a permanent state. \emph{Estoy casado} means "I'm married," but
some feel it implies "for the moment" or "at present," something akin
to "I am passing through a married phase at the moment." (\emph{Estoy de
casado} would say that unambiguously.) Nonetheless, a man wouldn't
say \emph{soy casado con} (wife's name) but \emph{estoy casado con} (wife's name).
In the past tense, the distinction becomes very clear: \emph{Fui casado con
Maria} means "I was (forcibly) married to Maria (and may still be)." \emph{Estuve casado con Maria} means "I was married to Maria (who is now my ex-wife)."

Look at the following examples and practice separating them
in your mind:

\bsk

\indu \emph{Fue cambiado}. = "It was changed (by someone)."

\indu \emph{Estaba cambiado}. = "It was (looked) changed." 

\indu \emph{Fue dormido}. = "It was put to sleep."

\indu \emph{Estaba dormido}. = "It was asleep."

\indu \emph{Fue rota}. = "It was broken (by someone)."

\indu \emph{Estaba rota}. = "It was (already) broken."

\section{The hard ones: descriptive adjectives}

Don't worry about mastering the gray areas between \emph{ser} and
\emph{estar} from the start. It's enough to know why the differences exist so
as to incorporate them intuitively as you go along. With practice, the
light gray areas will get progressively lighter and the pitch-black regions will soon turn a sort of dark gray. Examining examples and asking yourself why? is the best way to start shedding light on the matter.
Some dubious cases of \emph{ser} versus \emph{estar} follow. Absorb them at your
own pace.

Perhaps you've noticed in your dealings in Spanish that to compliment someone on, say, his beauty, you use \emph{estar}: \emph{Estás guapo}. But
aren't you in a sense suggesting that his beauty is just a temporary
state, that you're saying, "You are beautiful today (but not as a general
rule)?" Some compliment! There's a kernel of truth in your suspicion,
but perhaps because of human vanity, such a comment is generally
taken favorably to mean "You look especially beautiful today."

Compliments highlight one of the largest zones of overlap between \emph{ser} and \emph{estar}, the descriptive adjectives. For instance, if someone is tall, they are presumably tall all the time, and we would correctly expect \emph{es alto} to convey that. So what the devil are we to make
of \emph{está alto}, which you will undoubtedly come across sooner or later?
Certainly you can't temporarily be tall?

In general, using \emph{estar} with adjectives is a way of highlighting
the immediate and subjective nature of a perception---"This is my
impression" or "This seems especially that way to me now." \emph{Es alto}
means "He is tall." \emph{Está alto} means, more or less, "He's so tall," "He's
much taller than I thought," "Gosh, he's tall." To say "He's tall for his
age," for instance, you would use \emph{Está alto para su edad}. If he were
tall, period---in other words, a tall person---you would simply say
\emph{Es alto}.

Let's take another adjective. \emph{Es feo} would be "He is ugly"-no
debate permitted or even needed. Look up "ugly" in the dictionary and
you'll find his picture alongside. Follow him home and he'll have ugly
parents. So what's left for \emph{Está feo?} It could mean "He's temporarily
ugly"---because of a horrible haircut, for instance. Or it could suggest
"He sure looks ugly to me" or "He really is ugly."

Becoming adept at making this distinction is a matter of time
and exposure. What's the difference, for instance, between \emph{es difícil}
and \emph{está difícil?} Roughly, \emph{es difícil} describes something that is always
difficult and notoriously so: say, swimming the English Channel or
learning Chinese. \emph{Está difícil} suggests that something that has come
up is difficult or that something is harder than was expected. \emph{Es difícil aprender la diferencia entre ser y estar}
 = "It is difficult to learn the
 difference between ser and estar." \emph{Está difícil aprender la diferencia
 entre ser estar} = "I'm having real trouble with this \emph{ser} and \emph{estar}
business."

Now how about \emph{es viejo} versus \emph{está viejo?} The first example
means someone is old, period---a senior citizen. The second, with
\emph{estar}, is much more subjective and can cover a wide range of English
translations, including "He feels old," "He looks (seems) old," and
"He is too old (for some specific task)." A similar case is \emph{ser joven}
versus \emph{estar joven}. \emph{Soy joven} means "I am young (i.e., a member
of the group of young people)." \emph{Estoy joven} covers anything from "I
feel young" to "I'm young (for my position)" to "I'm still young" (a
washed-up pitcher to his coach), and so on. Note the difference in the
past: \emph{Cuando eras joven} = "When you were young (a youngster)."
\emph{Cuando estabas joven} = "When you still had some pep (weren't over-the-hill)."

How about \emph{es buena} versus \emph{está buena} in reference to, say, a
film? Again, the key is in the subjective appreciation. After seeing it
and liking it, you might say \emph{Está buena la película} to communicate
your personal approval. You could use \emph{Es buena}, too, but there you'd
be declaring "It is a good film"---well done, professionally made, with
good actors, and the recipient perhaps of several awards. If after seeing
the film you say \emph{Es una buena película}, you are subtly implying "It
was good, but\ldots{}" In my experience, people seem to say \emph{Es una
buena película} in reference to arty films---ones they presume to be
"good" but didn't understand or particularly enjoy.

Some final examples will call attention to an advanced aspect
of the distinction with descriptive adjectives. If a person has a permanent physical illness (polio, for instance), he or she can still be described using \emph{está enfermo} or \emph{está enferma}. \emph{Está enferma desde niña} =
"She's been ill since childhood." Likewise, the adjective \emph{loco} is generally used with \emph{estar}, even when referring to someone who has spent
fifty years in an insane asylum. You would say \emph{está loco} of this person,
less commonly \emph{es loco}. Why? Because illness and insanity, as in these
cases, are not an essential part of the person's character but an exceptional, uncharacteristic condition. That is, it is not in their very nature
as people to be sick; it is a condition, a state, an exception. When adjectives like \emph{enfermo} and \emph{loco} are made into nouns, though, they are
used with \emph{ser}---but usually only with a preceding indefinite article: \emph{Es
un enfermo} = "He's sick (a sick person)."

The extreme application of the "essential nature" principle is
\emph{está muerto} and \emph{está muerta}, which is the only way to say "He (or
she) is dead." Students, understandably, balk at this one, since for most
of us death is considered a fairly permanent state, worthy of \emph{ser}. Actually, the use of \emph{estar} makes sense if you take the perspective of the
individual involved: being dead may be a lasting experience, but it's
not an essential aspect of the individual's nature. When the person is
remembered and eulogized years later, people won't say "He (or she)
was a good person, a kind person, and a dead person." Besides, \emph{Es
muerto} means "He is killed"---the use of \emph{ser} and a past participle in a
passive construction. An illustrative if redundant example containing
both would be \emph{Fue muerto a tiros, y ahora está muerto}, literally "He
was killed by shots, and now he's dead."

\section{Sorting out \emph{ser} and \emph{estar} in the imperative}

Imperatives are a source of some added confusion with \emph{ser} and
\emph{estar}. I've never found a good rule governing their usage in the imperative, so I'll invent one: avoid using either of them in the imperative,
but if you must, always use \emph{ser}. It's a fairly drastic rule, and exceptions
can of course be found if you want to get picky. But it will do for the
most part.

Why avoid imperatives with the "to be" verbs? Because Spanish, unlike English, does not lend itself to them as a rule. In English
you can without hesitation say "Be good," "Be on time," "Be there,"
"Don't worry, be happy," and so on. To translate these constructions
into Spanish, you would almost always resort to a verb other than the
"to be" verbs: \emph{Pórtate bien, Llega a tiempo, Asiste, Anímate}. If you
insisted on using a "to be" verb in Spanish, you would almost always
use \emph{ser}, even when referring to a transitory state: you would say \emph{Sé
amable}, for instance, to express "Be friendly," even if you meant it
only for a short while. \emph{Sé amable con tu abuela, sólo nos visita de vez
en cuando} = "Be friendly to your grandmother, she only visits us from
time to time."

Encouraging you to avoid using "to be" in the imperative in
Spanish is not to say that you won't hear it. It's not an especially common construction, but neither is it rare. Here are some examples you
may run across:

\bsk

\indu \emph{Sé puntual}. = "Be on time."

\indu \emph{Sé buena gente}. = "Be a nice guy (or gal)."

\indu \emph{Estate quieto}. = "Be still." (said to children)

\indu \emph{Estate callado}. = "Be quiet." (said to children)

\bsk

The imperative with both ser and estar is much more frequent
in the negative in Spanish:

\bsk

\indu \emph{No seas malo}. = "Be a pal"

\indu \emph{No seas tonto}. = "Don't be a fool"

\indu \emph{No seas imbécil}. = "Don't be a jerk."

\bsk

With estar the negative imperative almost always is constructed with the present participle:

\bsk

\indu \emph{No estés molestando}. = "Quit bugging me."

\indu \emph{No estés gritando}. = "Quit shouting."

\bsk

Even in the negative, though, imperatives tend not use a "to
be" verb at all, as we have seen:

\bsk

\indu \emph{No te enojes}. = "Don't be mad."

\indu \emph{No llegues tarde}. = "Don't be late."

\indu \emph{No te aloques}. = "Don't be crazy."

\indu \emph{No hagas ruido}. = "Don't be noisy."

\bsk

For the student, the \emph{ser-estar} confrontation is a real and constant struggle. Penetrating these verbs' distinct natures can be time consuming and, frankly, a real pain in the backside. Effort is definitely
required, but it is also repaid, since the intuitive understanding of
Spanish you gain in separating \emph{ser} from \emph{estar} will prove indispensable
to true fluency. As that dictionary quoted above goes on to say about
the \emph{ser-estar} problem: "If [foreigners] feel irritated with Spanish for
this difficulty, they should consider that the differentiation between
the essence and the state of things in everyday speech is but one more
demonstration---perhaps the most brilliant one---of the logical sense
of this language."

\section{A matter of perspective}

Learning how to make your English turn into correct Spanish
is sometimes a matter of mastering the vocabulary and sometimes a
matter of mastering a concept. But sometimes---rarely---it's a matter of
mastering a whole new way of looking at things. It's a matter, in short,
of effecting a change of perspective. Long after basic fluency has been
achieved, many foreigners still have trouble remembering to make this
change. Most foreigners, it could honestly be said, never make the
change completely.

Nowhere does this perspective problem crop up with greater
frequency than with the indispensable verbs \emph{llevar} and \emph{traer}. The key
to getting this distinction right is to remember and implement a very
basic rule: in Spanish you can't "bring" something from where you are
to where you aren't. If you are going to a dinner at a friend's house, you
must ask if you should "take" (\emph{llevar}) something (a salad, a bottle of
wine, etc.). "Bringing" is only for cases when something away from the
speaker is being moved toward the speaker.

In English we tend to play loosely with what is essentially the
same rule. That is, we use "bring" regardless of whether the implied
movement is toward the speaker or away from the speaker. If we are
going to a party, we will offer to "bring" a salad; we will "bring" a
cooler with us when we go on a picnic. In Spanish you have to "take"
the salad and "take" the cooler.

\bsk

Imagine the following phone conversation:

\bsk

\inda JOSÉ:

\indu "Hey, Carlos, I think I left my wallet at your house last night.
Could you bring it over today?"

\inda CARLOS:

\indu "Sure. I'll bring it over in the afternoon."

\bsk

Now, in Spanish:

\bsk

\inda JOSÉ:

\indu Oye, Carlos, creo que deje mi cartera en tu casa anoche. ¿Me la
puedes \emph{traer} hoy?

\inda CARLOS:

\indu Claro, te la \emph{llevo} por la tarde.

\bsk

It's important to pay attention to the distinction not just to
sound better but to avoid sounding rude and demanding when you're
not. Imagine, in our phone conversation, that Carlos had no reason
to go by Jose's house that afternoon---in fact, imagine that it was an
hour out of his way. Imagine further that they have a class together
at five o'clock at the university. Now Jose, by asking Carlos to \emph{traer}
the wallet, is being dreadfully uncouth. Jose is saying, in effect, "Bring
it to me here at my house" when he may have meant "Take it to me
there at school"---in which case he would have to say \emph{¿Me la puedes
llevar hoy?}

The same problem of perspective comes into play with \emph{ir} and
\emph{venir}. In English we can call home and ask if the plumber "came" that
day; we can say we'll fix the sink when we "come" home; when we are
called to the phone, we say "I'm coming." In all of these cases in Spanish, however, you have to use \emph{ir} ("to go"), not \emph{venir} ("to come"). To
rephrase the \emph{llevar-traer rule}, you can't "come" (with \emph{venir}) to a location that is somewhere other than where you are at that moment.
\emph{Venir} can only refer to your present location---where you are sitting or
standing or, in a larger sense, to the city or country you are in. \emph{¿Fue el
plomero hoy?} = "Did the plumber come today?" \emph{Arreglaré el lavabo
cuando vaya} (or \emph{llegue}) \emph{a casa} = "I'll fix the sink when I come home."
\emph{Voy} = "I'm coming (to the phone)."

\chapter{The Twilight Zone}

First, an explanation. Why title this chapter, which is about
the subjunctive mode in Spanish, "The Twilight Zone?" The answer is
simple and twofold. First, the concept of a hazy, ephemeral Twilight
Zone accurately conveys the spirit of the subjunctive. And, second, if
it were called "The Subjunctive Mode," no one would read it.

What is it about the subjunctive that inspires such fear and
loathing in students of Spanish? Mostly, it is the task of retraining the
mind to recognize a concept that has no readily obvious equivalent in
English. After all, it's bad enough that Spanish puts different endings
on its verbs to denote mode, tense, and person. But to invent a whole
new mode outright---one that needs endings all its own---is nearly
criminal.

Spanish, of course, did not invent the subjunctive. In fact, the
subjunctive is widely used in English, though not nearly as frequently
as it is in Spanish. Take, for example, a sign hanging in the Sears restrooms in Waco, Texas: "It is important to us that our restrooms be
clean." A nicer, neater subjunctive was never seen.

But in Spanish it's often hard to get a grasp of why the subjunctive is needed and when. Thus "The Twilight Zone." For that, essentially, is what the subjunctive is: the Twilight Zone of the verb universe. The subjunctive gets the job of describing "could-have-beens,"
"might-bes," and "maybe-never-weres.'~Anything that has happened,
is happening, or may happen on the borders of our consciousness gets
handled by the subjunctive. Without the subjunctive, Garcia Marquez
would read like Hemingway. The subjunctive is more than a verb
mode; it is a complete separate reality.

Although beginning and intermediate students of Spanish find
it difficult to believe, many English speakers who have learned to live
with the Spanish subjunctive will tell you that it can actually be quite
enjoyable. With the mere flick of a verb ending you can cast doubt or
aspersions, relegating a simple occurrence to a different realm of understanding. It is a realm you can flirt with and explore, avoid when
you want (sometimes), and revel in at will. Almost certainly you will
find yourself wanting it more and more as you explore the Spanish-speaking world. Magical realism finds its home here, as does the seeming surrealism of much of daily life in the Spanish-speaking world.

So how do you go about learning it? Basically, there are two
equally important approaches. One approach is to learn the
rules-especially with the Twilight Zone concept in mind. A second
approach is to learn the common cues for the use of the subjunctive.
Practice both, and the subjunctive will appear to you one day-perhaps
even in a dream. From that day forward, it will be your constant companion, your escape hatch into the unreal. Signpost up ahead: Subjunctive Mode!

\section{Indirect commands (shallow twilight)}

This group covers giving orders, asking others to do things,
and engaging in other bossy behavior. Thus in a sentence like "Tell the
mariachis to go away," the English infinitive "to go away" must be rendered in the subjunctive in Spanish. Why? Because the action of the
mariachis going away doesn't become a reality until they actually go
away. Until then, it must be considered an entirely suspect notion,
lurking off in the unknown: Will the mariachis go away? Will they
stay? How about if we pay them to go away?

We can lump implicit indirect commands in this category, including "wishing" and "hoping" that the mariachis will go away. "I
hope the mariachis go away" is, after all, nothing more than a cowardly version of "Tell the mariachis to go away." And, as in the first
example, simply hoping that they will go away is no guarantee that
they will actually do so. They may want to sing "La Bamba" again.
You have no way of knowing, so you have to rely on the subjunctive. In
passing, note that you won't always have an obvious "telling" or "hoping" verb directing the action. Sometimes, the order is implicit or impersonal, and it often begins with que all by itself: \emph{Que se vayan los
mariachis} = "Have the mariachis go away." Similarly, \emph{ojalá} will often
initiate a wishing construction: \emph{Ojalá que se vayan los mariachis} =
"I sure hope that the mariachis will leave."

\section{The eternal mystery (deep twilight)}

For this category, we must venture even further into the Twilight Zone. This is the realm of doubt, uncertainty, suspicion, and
downright disbelief. For example, \emph{Es posible que los mariachis se
vayan} ("It's possible that the mariachis will go"), \emph{Dudo que se vayan}
1"1 doubt that they will go"), \emph{No creo que se vayan jamás} ("I don't
think they are ever going to go"), and so on. In each of these cases, we
return to the elemental problem of the mariachis' departure as a mystery, an eternal uncertainty, an action belonging to a separate realm.

Statements of negation also lurk in this shadowy world. At
first, their presence here confounds us: aren't they statements of fact
and thus perfect candidates for the indicative? On closer examination,
however, we can see why they are here. Negations are declarations of
something that never happened, actions that only exist in somebody's
mind. Here, with a little study, we can see the careful distinction that
turns a harmless indicative statement into an unruly, ethereal subjunctive. \emph{No asalté el banco} = "I didn't rob the bank" Straightforward and
indicative: I didn't do it. But when the sentence structure forces us to
make the action of bank robbery stand alone, it acquires its true character---that of an action that was not. \emph{Niego que haya robado el banco}
= "I deny that I robbed the bank." Here "I robbed the bank" is a nonevent, an untrue claim, a load of nonsense. It simply didn't happen---I swear! And since it didn't happen, it must be exiled to that world
where all the things that never happened---the "could-have-beens,"
"might-bes," and "maybe-never-weres"---reside. In short, it must go
to the Twilight Zone.

A similar treatment awaits things you "don't believe" or
"don't think." Since you don't believe them, you certainly don't have
to consider them real. \emph{No creo que esté aquí} = "1 don't think she's
here." Her presence here is something that for you, the speaker,
doesn't belong in your universe of hard facts. Thus into the subjunctive
it goes.

\section{\emph{Que} cues}

By now, being a sharp reader, you will have noticed that every
use of the subjunctive so far has been preceded by a certain word: \emph{que}.
And you're thinking, "Hey, maybe I'm on to something." In fact, you
are---sort of. Que is a good cue for using the subjunctive, though not
an entirely reliable one. That is, almost every time the subjunctive
appears, there will be a \emph{que} preceding it. But \emph{que} will not always be
followed by the subjunctive every time it appears. Still, if you pay
attention to when you use \emph{que}, you will be on your way to spotting
opportunities for showing off your subjunctive.

The context surrounding \emph{que} is the deciding factor in whether
the subjunctive should indeed follow. And often this context is little
more than the proper combination of words with \emph{que}. Thus certain
impersonal expressions followed by \emph{que} almost invariably take the
subjunctive:

\bsk

\indu \emph{Es posible (probable, factible, concebible) que}

\indu \emph{Es mejor (conveniente, preferible, oportuno) que}

\indu \emph{Es importante (necesario, preciso, urgente, obligatorio,
forzoso) que}

\bsk

As a matter of fact, only a handful of common adjectives can
be placed between es and que and still produce the indicative. Some of
these exceptions are \emph{claro, obvio, evidente}, and the like, which stress
that a fact is a fact is a fact. \emph{Es obvio que estoy aquí} = "It's obvious
that I'm here." So obvious a fact certainly has no business in the Twilight Zone.

Many of the adjectives in the \emph{es} + adjective constructions
used above also have verb forms. With our old friend \emph{que} these verbs
also take the subjunctive:

\bsk

\indu \emph{Urge que} (it's urgent that)

\indu \emph{Conviene que} (it suits/appropriate)

\indu \emph{Precisa que} (it states/specifies that)

\indu \emph{Prefiero que} (I'd rather)

\bsk

Along these lines are other impersonal expressions that take
the subjunctive:

\bsk

\indu \emph{Más vale que} ([you/we'd] better)

\indu \emph{Lástima que} (it's a pity/shame/too bad)

\bsk

So do certain "impersonalized" or reflexive constructions:

\bsk

\indu \emph{Se espera que} (always; it is expected)

\indu \emph{Se cree que} (sometimes; it is believed that)

\bsk

If you change these constructions from impersonal to personal, in
most cases you will still need the subjunctive. Some common verbs
that, when followed by \emph{que}, usually require the subjunctive include
\emph{esperar, sentir, querer, pedir, mandar, dejar}, and \emph{permitir}.

Note that when there is no change in subject, the infinitive
can be substituted for the subjunctive clause, as it is in English. These
are the constructions you won't have any trouble with:

\bsk

\indu \emph{Quiero ir}. = "I want to go."

\indu \emph{Esperan ganar}. = "They hope to win."

\bsk

Changing the subject of the second clause will require the subjunctive,
however:

\bsk

\indu \emph{Quiero que vayas}. = "I want you to go."

\indu \emph{Espero que gane ella}. = "I hope she wins."

\bsk

With some indirect command verbs, especially \emph{mandar, permitir}, and \emph{dejar}, the imperative can also be rigged together with the
infinitive to avoid the subjunctive altogether:

\bsk

\indu \emph{Manda traer el dinero}. = "Send for the money to be brought."

\indu \emph{Déjale traer el dinero}. = "Let him bring the money."

\bsk

A more natural-sounding construction in these and other cases of indirect commands is simply starting your sentence with \emph{que} and following it with the subjunctive, as in the earlier example \emph{Que se vayan los
mariachis}. This equates with the English "Have\ldots{}," which is one of
the most common indirect command forms in English.

\bsk

\indu \emph{Que traiga el dinero}. = "Have him bring the money."

\indu \emph{Que venga a las seis}. = "Have her come at six."

\bsk

\emph{Que} is also a reliable cue for the subjunctive when paired with
other words to form certain conjunctions. Most textbooks will give
you a laundry list of these conjunctions, half of which you will probably never need. Here are the important ones to remember:

\bsk

\indu \emph{para que} = "so that," "in order that"

\indu \emph{a menos que} = "unless"

\indu \emph{a pesar de que} = "despite," "even though"

\indu \emph{antes de que} = "before"

\section{Non-\emph{que} cues}

Another common use of the subjunctive is generally not introduced by que, so you'll have to be alert for it. Instead, it uses cuando,
donde, como, and other adverbs. The best guide in this case is the English translation. When you could substitute "-ever," as in "whenever," "wherever," or "however," follow the adverb with the subjunctive
in Spanish. If it helps you remember, memorize one of the classic
lines used to challenge someone to a fight in Spanish: \emph{Cuando quieras,
donde quieras, y como quieras} ("Whenever you want, wherever you
want, however you want"). (The expression is reputedly in use as well
as a "pick up" line, so make sure the person you use it on knows
whether you're a lover or a fighter!) Most likely, though, you will be
called upon to use this subjunctive construction in these more mundane situations:

\bsk

\indu \emph{¿Cuándo quieres ir? Cuando tú quieras}. = "When do you
want to go?" "Whenever you want."

\indu \emph{¿Adónde vamos? Donde quieras}. = "Where are we going?"
"Wherever you want."

\indu \emph{¿Cuándo me vas a dar el dinero? Cuando yo quiera}, = "When
are you going to give me the money?" "Whenever I
feel like it."

\section{The subjunctive with \emph{ser}: \emph{sea}}

\emph{Ser} is also commonly used in "-ever" constructions, and expressions with \emph{sea} are good to slip into your conversational Spanish.
\emph{Cuando sea, como sea, donde sea}, and \emph{quien sea} are equivalents for
"whenever," "however," "wherever," and "whoever" when used alone.
Often more common in English is to use an "any-" word---"anywhere," "anyhow," and so forth:

\bsk

\indu \emph{¿Con quién quieres ir al cine? Can quien sea}. = "Whom
do you want to go to the movies with?" "With
whomever."

\indu \emph{¿Dónde quieres comer? Donde sea}. = "Where do you want to
eat?" "Wherever (anywhere)."

\indu \emph{¿Cómo quieres la carne: con salsa, sin salsa, can papas, sin
papas, término medio, bien cocida? Como sea}. =
"How do you want your meat: with sauce, without
sauce, with potatoes, without potatoes, medium, well
done?" "Any ol' way will do."

\bsk

Often, the best English translation of expressions using \emph{sea}
would be a slangy expression like "It's up to you," "You name it," "I
don't care," or "It doesn't matter." All of these can be conveyed by the
Spanish subjunctive.

Although \emph{ser} and \emph{querer} are the two commonest verbs used in
"-ever" expressions, virtually any verb can be used:

\bsk

\indu \emph{¿Cuándo vas a llegar a la fiesta? Cuando pueda}. = "When are
you going to get to the party?" "Whenever (as soon as)
I can."

\indu \emph{Yo quiero salir ahora. Bueno, lo que tú digas}. = "I want to
leave now." "Okay, whatever you say."

\bsk

Once you get a feel for the \emph{cuando quiera- donde quiera- como quiera complex}, you'll be close to mastering one of the trickiest uses of the subjunctive: a clause containing the adverb plus the
subjunctive to refer to the future. Here you should keep the Twilight
Zone idea in mind. In Spanish, for instance, you would say \emph{Cuando
termine el libro, te llamaré} for "When I finish the book, I'll call you."
In this case, \emph{cuando} is followed by the subjunctive because it refers to
an event in the future that may never happen. A meteor could strike
the reader one page from the book's end, so the notion of "when I finish the book" must be considered uncertain.

Only when you are referring to a habitual action should you
use the indicative. In these cases, note that you are not so much referring to the future as to the past. \emph{Cuando termino de leer en las mañanas, voy a la tienda} = "When I finish reading in the morning, I go
to the store." Here the indicative is safe because presumably you have
done this sort of thing before and thus know that it can happen and has
happened.

\section{The traveler's subjunctive}

A final note on the subjunctive is especially useful for those
traveling in the Spanish-speaking world. A common question format
for lost, bewildered, or just-curious travelers goes as follows: "Is there
a such-and-such near here that does such-and-such?" For instance, you
might want to ask, "Is there a store near here that sells wine?" or "I'm
looking for place where I can leave my luggage." In all cases like these
you must use the subjunctive in Spanish, since the place you are seeking mayor may not exist. Or, put another way, it won't exist until your
question is answered, "Yes, there is such a place." Until such an answer is given, the place belongs in the never-never world of the Twilight Zone.
\emph{¿Hay una tienda por aquí que venda vino? Busco un lugar
donde pueda dejar mi equipaje}. The same logic applies when you ask
about people. ¿Hay alguien aquí que hable inglés? = "Is there anyone
here who speaks English?"

When the answer is in the negative, the place remains nonexistent and therefore must still be referred to in the subjunctive. \emph{No hay
una tienda cerca que venda vino} = "There is no store nearby that
sells wine." This is, after all, but a simple statement of negation, like
the ones we saw above. Ditto for nonexistent people: \emph{¡Na hay nadie en
esta ciudad que me entienda!} = "There's no one in this city who understands me!"

\chapter{Sixty-four Verbs, Up Close and Personal}

Pocket dictionaries will generally give you simple, one-word
equivalents for Spanish verbs. Better dictionaries will give you a list of
other possible meanings and maybe some examples. But rarely is the
dictionary reader given guidance on what usages are common---i.e.,
worth the bother of learning---and which are poetic or archaic and thus
irrelevant. And besides, who wants to read a dictionary?

What follows is a pared-down list of sixty-four Spanish verbs
whose basic meaning you probably already know, but whose inner secrets and common usage go far beyond that. The list cuts through the
clutter and highlights unexpected usages that a native English speaker
may not be on the lookout for or that are "out of character" for a given
verb. Each section also explains a chosen few idiomatic expressions,
selected for their frequency in everyday spoken Spanish. There's a lot
to absorb here, but the alphabetical listing will allow for hour after
hour of repeated consultation. Take heart---it could be worse. You
could be reading the dictionary!

\section{\emph{acabar}}

This is a synonym for \emph{terminar} and, like that word, is a good
equivalent for most uses of "to end" and "to finish" in English. Used
with \emph{con} as an intensifier, both verbs work well for "to finish off":
\emph{Acabé con la leche}. Often you will hear \emph{acabarse} in the reflexive to
mean "to run out of." \emph{Se nos acabó el dinero} = "We've run out of
money," \emph{Se acabo} by itself, meanwhile, means either "It's over" or
"I'm out of it," You'll hear it a lot in stores, at newsstands, and the like
when what you want to buy is no longer in stock. \emph{Terminar} is likewise
used this way; \emph{agotar} is also heard, especially in the phrase \emph{Está agotado} ("We're out of it"). \emph{Está acabado} usually isn't used in this sense,
since its meaning is closer to "It's finished" or, colloquially, "He's
washed up." \emph{Acabar} has one other very common use that \emph{terminar}
doesn't have: in the present tense, with \emph{de}, it means "to have just,"
as in Acabo de comer ("I've just eaten"). Used in the imperfect, it becomes "had just." \emph{Acababa de comer cuando llegaste} = "I had just
eaten when you arrived."

\section{\emph{amanecer}}

This somewhat uncommon word, which means "to dawn," is
included here because of one expression that has perplexed generations
of language students, especially those who live for a time with a
Spanish-speaking family when studying abroad. The expression is
\emph{¿Cómo amaneciste?}---which translates literally as "How did you
dawn?" but which means "How did you sleep?" The answer is (usually) \emph{Bien, gracias}. A fun Spanish expression for "to die in one's sleep"
uses this verb: \emph{Amaneció muerto}; if it translates at all, it would have
to be rendered "He woke up all dead."

\section{\emph{andar}}

Dictionaries say it means "to walk," which of course it does,
but that won't help you when you hear \emph{anda corriendo} or \emph{anda en
coche} for the first time. In English we would probably be inclined
to say "go around" for most uses of \emph{andar}. \emph{Pedro anda gritando tu
nombre} = "Pedro's going around shouting your name." \emph{Andar} also
covers slangy expressions like "to hang out" or "to hang around." \emph{Ya
no ando con ellos} = "I don't hang around with them anymore." \emph{¿Por
dónde andas?} works well for "Whereabouts are you?" or the colloquial
"Where are you at?" And in some countries \emph{anda} lends itself to the
common idiomatic expressions \emph{¡Ándale!} and \emph{¡Anda!} Said with vigor,
they mean "Let's get a move on!" or "Way to go!" Said in passing, they
mean the same as "okay" or "all right." In Mexico, for instance, you'll
hear \emph{ándale} all the time for "that's fine," "that's right," and even
"good-bye": \emph{Nos vemos mañana. Ándale}. Throw a \emph{pues} on the end
and you'll be saying nothing at all but will sound very fluent: \emph{Ándale
pues} ("Have a nice day"). Remember as well the use of \emph{andar} for "to
run" or "to work" in reference to objects. \emph{¿Qué tal anda tu coche?} =
"How's your car running?" (literally, "How's your car walking?").
Some wags have even argued that the different conception of time in
Spanish-speaking countries is due to the fact that in Spanish clocks
walk rather than run!

\section{\emph{antojar}}

Used generally as a reflexive (\emph{antojarse}), this is an exceptionally common verb and one that you should get to know well. To translate it, dictionaries offer "to long for" or "to desire earnestly," but its
use in Spanish covers a lot more ground than that. Closer to the mark
would be "to get a hankering for" or simply "to feel like." An \emph{antojo}
is a "craving" or an "urge" and covers both intense longings (like the
kind that make pregnant women eat pickles) and simpler pleasures.
You'll probably run across the verb frequently in these same situations. Some examples: \emph{Se me antoja una pizza} = "I'm dying for (could
go for) a pizza." \emph{¿Por qué no vas a ir al cine? Porque no se me antoja}.
= "Why aren't you going to the movies?" "Because I don't feel like it."
\emph{Déjame una dona; luego se me va a antojar} = "Leave me a doughnut;
I'll probably feel like having one later."

\section{\emph{bajar}}

This verb means "to go down," "to put down," "to get off,"
and so on. Most of its uses are predictable, but a few that may not be
include "to go downstairs," "to get out of a car (bus, train, etc.)," and
"to lose weight" (\emph{bajar de peso}). It also means "to get (something)
down," as when you ask someone to get your suitcase down off the
rack (\emph{¿Me baja la maleta, por favor?}). \emph{Bájale}, by itself, is usually "Turn
it down," referring to the volume or the general noise level; in the
right context, it can also mean "Slow down."

\section{\emph{caber}}

An irregular verb that you should learn. It means "to fit," but
only in the sense of "fit into" or "fit onto." It is not used for clothing.
In the first-person present, it's \emph{quepo}, and you're likely to hear it in
\emph{¿Quepo yo?}---meaning "Will I fit?" or "Is there room for one more?"
Otherwise, you may run into it in set expressions like \emph{cabe decir} ("it's
worth mentioning") and \emph{no cabe duda} ("there's no doubt").

\section{\emph{caer}}

It means "to fall," of course, and "to drop" when used reflexively (\emph{caerse}): \emph{Se me cayó el vaso} = "I dropped the glass," It's also
very frequently heard in the phrases \emph{caer bien} and \emph{caer mal} to express
likes and dislikes (see Chapter 4). You may also run across \emph{caer} for "to
visit unexpectedly" or "to drop in on." \emph{Te caigo en la tarde} is an informal way of saying "I'll drop in on you in the afternoon." Sometimes
it's used to suggest that someone's arrival was not only unexpected but
also unwelcome. "What are your in-laws doing here?" might be answered by \emph{Es que me cayeron} ("They just kind of showed up").

\section{\emph{cambiar}}

Meaning "to change" as well as "to make change" in the sense
of "Can you change a twenty?" (\emph{¿Me puede cambiar un billete de a
veinte?}), \emph{cambiar} also crops up in a number of common expressions.
These include \emph{cambiar de idea} or \emph{cambiar de opinión} ("to change
one's mind"), \emph{cambiar de ropa} ("to change one's clothes"), and \emph{cambiar de casa} ("to move").

\section{\emph{coger}}

This is one of those words that many dictionaries handle with
more discretion than clarity. The simple fact is that \emph{coger} is a vulgar
term for "to fornicate" in several countries (Argentina, Mexico, Uruguay, and others), where as a result it is rarely used in proper company
(see Chapter 10). That said, it is also one of the most commonly used
verbs in some other countries (especially Spain). What's a poor student
to do when faced with the choice? That will depend on where you are
learning the language and with whom you expect to be communicating. But if you want to use substitutes for coger right from the start---in the sense of "to get," "to take," "to grab"---it may not be such a bad
idea. The word that usually replaces it is \emph{tomar}, as in \emph{tomar el tren}. In
Mexico, particularly, \emph{agarrar} is often heard. Both substitutes are 
understood even where \emph{coger} is used, and both can save you considerable
embarrassment.

\section{\emph{conocer}}

This verb is often confused with \emph{saber} by students of Spanish;
both mean "to know," but \emph{conocer} is used in the sense of "to be familiar with." A limited rule of thumb: use \emph{conocer} for proper nouns and
all specific people, places, and things; use saber for everything else. Do
you know Paris? \emph{Conocer}. Do you know the French Quarter? \emph{Conocer}.
Do you know the old lady there who sells flowers on the street? \emph{Conocer}. Do you know her name? \emph{Saber}. Do you know what flowers she
sells? \emph{Saber}. Do you know what she's saying about you? \emph{Saber}.

\emph{Conocer} also means "to meet," but keep in mind that it only
works for the first time you meet someone. English speakers use "to
meet" to describe routine encounters, such as "I met my mother at
the train station." They also tend to say \emph{la primera vez que lo conocí}
to convey "the first time I met him" (instead of the less redundant
\emph{cuando lo conocí}). In Spanish you can only \emph{conocer} someone once, and
needless to say it would be difficult (not to mention dramatic) to \emph{conocer} your mother at a train station. Finding the right Spanish word for
"to meet" in these other situations will give you an idea of how overworked this poor verb is in English. When you are referring to a first
meeting, as noted, use \emph{conocer}. \emph{Lo conocí en París} = "I met him (for
the first time) in Paris." All chance encounters after that are handled
by \emph{encontrar}. \emph{La encontré en el cine} = "I met (ran into) her at the
movies." Meeting a plane or a train would be covered by \emph{recibir}. \emph{Me
recibieron en la estación} = "They met me at the station." For a
planned get-together you would use \emph{quedar en verse con} or \emph{quedar en
encontrarse con}: \emph{Quedé en verme con unas amigas en el centro} = "I
met (up with) some friends downtown."

\section{\emph{creer}}

It means "to believe," but it is also almost always the word
you want for "to think." Beginning students, incorrectly, often prefer
\emph{pensar} (see below). The distinction is subtle, but it works out roughly
as follows: if the emphasis is specifically on the thought process or the
act of thinking, use \emph{pensar}. If you're stating a personal belief or opinion, use \emph{creer}. Thus \emph{Creó que tienes la razón} = "I think you're right."
\emph{Creer} is also used in many interjections and phrases, such as \emph{¿Qué
crees?} ("Guess what?"), \emph{Créeme} ("Trust me"), \emph{¿Tú crees?} ("You really
think so?"), and so on. A good phrase to learn is \emph{ni creas}, which could
be translated as "don't expect" or "no way." \emph{Ni creas que te voy a ayudar} = "You're crazy if you think I'm going to help you" or "Don't expect me to help you," \emph{Creo que sí} and \emph{Creo que no}, finally, should be
on the tip of your tongue for "I think so" and "I don't think so."

\section{\emph{cuidar}}

Technically, this means "to care for," though in English we
would generally use some other word to translate it. \emph{Ana se quedó
para cuidar a los niños} = "Ana stayed home to watch (take care of)
the kids." \emph{Cuide su cartera en ese barrio} = "Watch your wallet in that
neighborhood." \emph{Cuídate} is "Take care of yourself" or, slangily, "Take it
easy"; it is sometimes used as a parting comment. \emph{Cuidar} in general is
the word you want for asking someone to "watch over" or "to keep an
eye on" something, as when you want to leave your luggage in the bus
station for a few minutes (not a recommended practice). \emph{¿Me puede
cuidar la maleta unos minutos?} = "Can you keep an eye on my bag
for a few minutes?" See also \emph{guardar}.

\section{\emph{dar}}

'To give" and much more. A common additional meaning of
dar is "to hit," giving rise to a number of expressions. \emph{Dar en el blanco}
is "to hit the bull's-eye" and is often heard for "to guess right," "to hit
the nail on the head." \emph{Dar en la torre} is a common idiomatic expression that covers physical beatings as well as more metaphorical thrashings. "How did the Cubs do against the Mets?" \emph{Le dieron en la torre}
("They beat the pants off 'em"). \emph{Dale}, by itself, is "Hit it" (or "Hit
him," "Hit her"); you can let off steam by shouting it at boxing matches.
More metaphorically, it means "Give it your best" or "Give 'em hell."
Often, in this sense, it's paired with \emph{duro} and said to give encouragement: \emph{¡Dale duro, Juan!} ("Give 'em hell, Juan!"). By the same token,
you can use \emph{dándole duro} to convey the intensity of an action or an
effort. "How are you coming with the term paper?" \emph{¡Dándole duro!}

Here are other dar expressions you'll want to know:

\bsk

\indu \emph{dar a la calle} (\emph{al patio, a la alberca}, etc.) = "to face the street
(the patio, the pool, etc.)"

\indu \emph{da igual} or \emph{da lo mismo} = "it's all the same to me" or "it
doesn't matter"

\indu \emph{dar coraje} = "to make mad"

\indu \emph{dar (la) lata} = "to pester" or "to be a pain"

\bsk

\emph{Dar} is also used in some parts of the Spanish-speaking world for
"what's on"---in the sense of showings on television or at the movie
theater. \emph{Están dando la segunda parte de} Arma Mortal = "They're
showing the second part of \emph{Lethal Weapon}." (See also pasar.) Finally,
though you won't find it in the dictionary, an increasingly common expression in American Spanish is \emph{dar chance} for "to give a break." \emph{Vamos, oficial, denos chance} = "Come on, officer, give us a break." Purists will tell you that this is a horrible barbarism and that you should
say \emph{denos una oportunidad} instead. But purists should consider cutting their losses, since more and more speakers of slang are already bypassing \emph{dar chance} in favor of \emph{dar un break}, which is more barbaric yet.

\emph{Darse}, the reflexive form, is also used for a few handy phrases,
such as \emph{darse cuenta de}, which means "to realize." \emph{Perdón, no me di
cuenta de que estaba estacionado en su pie} =: "Sorry, I didn't realize
I was parked on your foot." \emph{Darse por vencido} is the phrase you want
for "to give up," "to surrender." To say "I give up" in response to a
riddle, for instance, you can even use \emph{Me doy} (more formally, \emph{Me rindo}) all by itself.

\section{\emph{decir}}

"To tell" or "to say." You'll probably also be using it a lot in
the phrase \emph{querer decir}, or "to mean." \emph{¿Qué quiere decir esa palabra?}
= "What does that word mean?" \emph{Decir} also pops up in a lot of cute
little phrases, such as \emph{No me digas} ("You don't say), \emph{Dime} ("Tell me"
or simply "Yes?"), and \emph{¿Qué decías?} ("What were you saying?"---after
an interruption). You should also be alert to \emph{Dile} (and \emph{Dígale}) as a typical preface to an indirect command, and thus the subjunctive. \emph{Dile que
venga} = "Tell him to come here." If you're stating a fact instead of
issuing a command, it takes the indicative. \emph{Dile que estamos aquí} =
"Tell him (or her) we're here."

\section{\emph{dejar}}

Meaning "to leave" or "to let," this verb is often found in
other expressions and is not always a reliable vehicle for English "let"
phrases. For example, \emph{Déjenos entrar} means "Let us in," but \emph{Entremos}
(or \emph{Vamos a entrar}) means "Let's go in." \emph{Dejar} is heard in such phrases
as \emph{Déjame en paz} ("Leave me alone"), \emph{Déjalo} ("Leave it" or "Drop it"
or "Skip it," often referring to a sensitive topic or one best dealt with
later), and \emph{Deja ver} or \emph{Déjame ver} = "Let me see."

\section{\emph{disfrutar}}

"To enjoy." Only mentioned here to dissuade you from saying
\emph{Disfrútense} for "Enjoy yourselves." It does mean this, but probably not
in the way you intend. "Have fun with yourselves" or "Take pleasure
in yourselves" would probably be a more accurate translation. Instead,
you should say \emph{Que lo disfruten}, usually in reference to a specific
event. For the best translation of "Enjoy yourselves," forget about \emph{disfrutar} altogether and use \emph{divertirse}: \emph{Diviértanse} or \emph{Que se diviertan}.

\section{\emph{dormir}}

\emph{Dormir} is "to sleep" while \emph{dormirse} is "to fall asleep." Just a
reminder: "sleep," the noun, has to be expressed by \emph{sueño} (see \emph{soñar}).

\section{\emph{durar}}

This is one of the words you will need to ask how long things
take (movies, bus rides, flights, speeches, etc.). You can think of \emph{durar}
as a generally safe translation of "to last." \emph{La fiesta duró toda la noche}
= "The party lasted all night." Generally \emph{durar} is used for things that
have a specific duration---or \emph{dura}tion, if it helps you remember. (See
also \emph{tardar} and \emph{hacer}.)

\section{\emph{echar}}

\emph{Echar} is one of those words that take up about three pages in
the dictionary. Not all of those expressions are that common or useful,
though. Almost all of them have to do with a forcible casting out or
expulsion. File away this division of labor: \emph{meter} is for putting in, \emph{sacar} for taking out, and \emph{echar} for kicking out, more or less. Buses \emph{echan}
smoke, teachers \emph{echan} students, and so on. An unexpected use of
\emph{echar} is for "to pour," as with a liquid into a glass or from a pitcher.
Some of the idiomatic expressions using \emph{echar} are very handy. \emph{Echar
de menos a} is "to miss (someone)" (though in the Americas you are
more likely to hear extrañar). \emph{Echar a perder} is "to spoil," be it children or food in your refrigerator. \emph{Echar ganas} is a very common
expression for "to show enthusiasm" or "to give it a good effort: "Mom,
I can't understand my math homework." \emph{Vamos, échale ganas, hijo}.
\emph{Echar una mano} is the expression you will need for "to lend a hand,"
and \emph{Échame la mano} is "Gimme a hand." \emph{Echar un ojo} is "to have a
quick look," and \emph{echar la culpa} is "to blame."

\section{\emph{encargar}}

This verb, meaning "to entrust" or "to commission," is far
more common than these awkward English translations suggest. Almost always it conveys some notion of "charge"---to take charge, to be
in charge, to charge, and so on. With \emph{de}, for instance, it means "to take
charge of" or "to put someone in charge of" something: \emph{Yo me encargo
de la ensalada} = "I'll be in charge of (take care of) the salad." \emph{El encargado} is the all-purpose word for "the guy in charge." Without the
\emph{de}, \emph{encargar} means "to entrust" but often with more colloquial English equivalents. For example, if someone's going off to the store, you
might say \emph{¿Te encargo unas aspirinas?} ("Can you get me some aspirin?"). Many usages could almost be translated "to order." \emph{Le encargué
dos litros al lechero} = "I ordered two liters from the milkman." Sometimes, what you're ordering or entrusting is not spelled out, and "to
count on" would make for a better fit. If someone offers to fix your car
by nightfall and you plan to leave town that very night, you might say
to the mechanic, \emph{Se lo encargo mucho} ("I'm really counting on you").

\section{\emph{equivocar}}

Used as a reflexive (\emph{equivocarse}), this is the common verb for
"to make a mistake" (although \emph{estar equivocado} works as well). Usually used with the preposition \emph{de}, it means "to get something wrong."
For instance, when you've dialed the wrong number, you would say,
\emph{Perdón, me equivoque de teléfono} (or \emph{número}). If Pedro shows up on
the wrong day for a party, you would tell him, \emph{Te equivocaste de día}.
And so on. As a subtle distinction, \emph{estar equivocado} generally suggests someone is mistaken; \emph{equivocarse} means he or she has made a
mistake.

\section{\emph{esperar}}

Meaning both "to hope" and "to wait," the verb will usually
take the subjunctive in both senses. \emph{Espero que venga} = "I hope he (or
she) comes." \emph{Estoy esperando que regrese} = "I'm waiting for him (or
her) to get back." Unless the context makes it clear, you would use the
preposition \emph{a} with \emph{esperar} to say "to wait for." Thus, \emph{Espero a que
regrese} = "I'll wait for him (or her) to get back." (\emph{Esperar por} or \emph{esperar para} should never be used for "to wait for.") \emph{Esperar} is frequently
heard in the imperative for "Wait a minute" or "Hold on." \emph{Espérese,
por favor} is the form you're most likely to hear. In familiar usage it's
\emph{Espérate}, which often comes out sounding like \emph{'pérate}.

\section{\emph{estar}}

The other verb for "to be"---the one that covers transitory
states. The \emph{ser} versus \emph{estar} confrontation was covered in detail in
Chapter 5. So here it's enough to glance at one use of the verb you
might not have been expecting---about the only one that is out of character for it from the standpoint of an English speaker---\emph{estar} when
used to ask "What's today's date?": \emph{¿A qué estamos hoy?} or \emph{¿A cuántos
estamos?} The answer is phrased \emph{Estamos a} and the number: \emph{Estamos
a 25}, for example. Remember too that \emph{estar + de} is used for moods
and inclinations. \emph{El está de malas} = "He's in a bad mood." (This use
is covered in Chapter 4.)

\section{\emph{estrenar}}

This is not the commonest of verbs, nor is it one that you desperately need to learn. It means to use or display something for the
first time---to "debut" something, as it were. \emph{Juan está estrenando su
nuevo coche} = "Juan is trying out his new car." \emph{Voy a estrenar la camisa que me regalaste} = "I'm going to wear (for the first time) the
shirt that you gave me." \emph{Un estreno}, referring to films and theater, is a
"premiere" or "opening"; referring to an artist, it would be a "debut."

\section{\emph{guardar}}

Meaning "to guard" or "to save," this is the common verb to
use for telling someone to "hold on to" something or "put (something)
away." \emph{Guardar} is usually used when you give someone something
that you want them to put away until you need it again. If you're going
swimming and don't want to take your traveler's checks into the pool
with you, you might hand them to a friend and say \emph{¿Me los guardas?}
\emph{Guardar} is also commonly used by parents to tell their children, "Put
it away" (\emph{Guárdalo}), be it in their closet or in their pocket. \emph{Guárdame
un poco} is "Save a little for me," said of a favorite foodstuff, for instance. (See also "Save" in Chapter 11.) The word is a little tricky to
use in the sense of its English cognate. \emph{¿Me guarda la maleta?} could be
"Keep an eye on my suitcase" but also something like "Put my suitcase away"; so if you tell someone to "guard" it and they walk off with
it, you'll know why. Use \emph{cuidar} (see above) for "to keep an eye on,"
"to watch."

\section{\emph{haber}}

This is one of the megaverbs, with more uses than you might
ever be inclined to learn. Some are indispensable, though. \emph{Hay}, of
course, is the way to say "There is\ldots{}" and "There are\ldots{}" Say it
with a lilt and it becomes the questions "Are there\ldots{}?" and "Is
there\ldots{}?" \emph{No hay} is the typical curt response to questions about
availability: \emph{¿Hay café? No hay. ¿Hay cuartos? No hay}. \emph{Hay que} is the
impersonal way of saying "to have to"---that is, when it's not obvious
exactly who has to do something. \emph{Hay que ir a España para aprender
español} = "One has to go to Spain to learn Spanish." The imperfect
and preterit forms of \emph{haber} are \emph{había} and \emph{hubo}, respectively. \emph{Había}
is the more commonly used of the two. \emph{Había veinte personas en el
coche} = "There were twenty people in the car." \emph{Hubo} is for something that was there all at once or not for long. \emph{Hubo un choque en la
carretera} = "There was an accident on the highway." Remember never
to use \emph{habían} or \emph{hubieron} for "there were," regardless of the number
of persons or things involved: \emph{Había una monja en la lancha} and \emph{Había dos monjas en la lancha}.

\section{\emph{hablar}}

A straightforward word for "to talk" or "to speak," but keep
it in mind for use on the telephone, where its use is rampant. It's not
hard to imagine a phone conversation going something like this:
(R-i-i-i-ing.) \emph{Hola. ¿Quién habla? ¿Con quién quiere hablar? Habla
Juan. ¿Puedo hablar can Fred? El habla. ¡Ah! ¡Hablas español!} Of
these, the \emph{El habla} (or \emph{Ella habla}) response is the most important to
get straight. It will spare you countless episodes of saying \emph{Soy él} (or
\emph{ella}) or \emph{Hablando}, both of which are incorrect when you mean "This
is he (or she)" or "Speaking."

\section{\emph{hacer}}

Hacer means "to make" or "to do," but you know that already.
Where English speakers have to remember to use \emph{hacer} is in the many
weather-related expressions that in English are covered by "to be."
Some of the things that "make" in Spanish are \emph{frío, calor, viento}, and
\emph{sol} ("cold," "hot," "windy," "sunny"). \emph{Hacer} is also the way to say
"ago" in Spanish: \emph{Hace dos años nació mi hijo} ("My son was born two
years ago"). When did they leave? \emph{Hace un rato} ("A little while ago").
To say "We did it!" or "We made it!" in Spanish, you say \emph{¡Lo hicimos!}
With \emph{la} instead of \emph{lo} and intensified with \emph{ya}, it becomes a colloquial
\emph{¡Ya la hicimos!} ("We've got it made!"). Don't trouble your mind
searching for an antecedent to la---there is none. A very useful phrase
with \emph{hacer}, finally, is \emph{Hazte de cuenta} (or \emph{Haz de cuenta}), which introduces a thought with "Pretend\ldots{}" or "Let's say\ldots{}" (See also
Chapter 8 under "\emph{Haz de cuenta que}.")

\emph{Hacerse}, in the reflexive, also turns up in a lot of unexpected
expressions. It's a common way of translating "to become" and is also
common idiomatically for "to make like" or "to act like." \emph{El se hace el
payaso} = "He's acting like a clown." \emph{Se hace el loco para no ir a la
cárcel} = "He's pretending to be crazy to avoid going to prison." You'll
often encounter this expression in the negative as an exhortation.
\emph{No te hagas tonto} = "Don't play stupid." \emph{No te hagas la víctima} =
"Don't play the victim." In Mexico especially you'll encounter \emph{No te
hagas} all by itself, with the predicate understood. This is a good translation for common interjections like "Come off it!" or "Don't gimme
that!" or "Cut it out!"

\section{\emph{ir}}

The verb for "to go" comes into play in many situations that
parallel English usages, but you'll have to be on the lookout for them
to learn to use them well. It's widely used, for instance, for negative
imperatives, otherwise known as warnings, equating approximately
with the English "Don't go ... ": \emph{No te vayas a meter en líos} =
"Don't go getting yourself in trouble." \emph{No le vayas a decir} = "Now
don't go telling him" or just "Don't tell him." In many compound verb
forms, \emph{ir} is extremely useful to get your point across. It's used, as in
English, in the future (\emph{Voy a llamar} = "I'm going to call") and in the
imperfect (\emph{Iba a llamar} = "I was going to call"). \emph{}And \emph{Vamos a} means
"Let's\ldots{}" Other common expressions with \emph{ir} include \emph{irle a} ("to root
for," as in \emph{Yo le voy a los Orioles}), \emph{ir por} ("to go get" or "to go for," as
in \emph{Voy por el coche}), and \emph{Ahí te va} (meaning "Catch" or "Your turn").
\emph{Vaya} by itself is "All right" or "Omigosh," depending on tone and context. \emph{Vaya, vaya, vaya} is the common way to say "Well, well, well," as
in "What have we here?" \emph{Vaya} plus a noun is the equivalent of the
sarcastic comment "Some\ldots{}!" If you return home to find the plumber
has managed to flood the entire basement, you might say sarcastically
\emph{¡Vaya plomero!} ("Some plumber!").

\section{\emph{lograr}}

This is the word you want for "to manage" when used with
another verb in the infinitive. \emph{Logré reparar la tele} = "I managed to
fix the television." \emph{Si logro salir de esta reunión, estaré en casa en
media hora}, = "If I manage to get out of this meeting, I'll be home in
half an hour."

\section{\emph{llevar}}

\emph{Llevar} means "to carry" or "to take" and is often used for
"to bring," as we saw in Chapter 5. Here we'll concern ourselves with
\emph{llevar} in those expressions you'll want to have handy in your daily
doings. One common one is \emph{llevar} for expressions of time. A good way
to answer the inevitable question heard abroad, "How long have you
been here," is to say \emph{Llevo} + the number + \emph{meses} (\emph{años, días}) \emph{aquí}.
The question, in fact, will often be phrased \emph{¿Cuánto tiempo lleva
(usted) aquí?} Once you get a feel for this usage, you'll find yourself
needing it more and more. \emph{Llevo dos días en cama} is much smoother
and more colloquial than \emph{He estado en cama durante} (or \emph{desde hace})
\emph{dos días} for "I've spent two days in bed." \emph{Llevar} is also useful for "to
wear" or "to have on." \emph{Es el señor que lleva 1entes} = "He's the man
with the glasses on."

\emph{Llevarse}, the reflexive, is also handy for a couple of expressions. A common one is for "to get along." \emph{Ella y yo no nos llevamos}
= "She and I don't get along." \emph{No me llevo con ellos} = "I don't get
along with them." Another one you'll need, especially for shopping,
is \emph{llevarse} for "to take" something (after paying for it, naturally). It
means the same as llevar essentially, but the reflexive is added for emphasis. \emph{¿Dos mil pesos? Me lo llevo} = "Two thousand pesos? It's a
deal." \emph{Me llevo dos} = "I'll take two." \emph{Llévatelo} is "Take it," pure and
simple. Any usage that translates as "to take along" gets the reflexive
as well: \emph{Me llevo un suéter por si hace frío} = "I'll take a sweater in
case it gets cold."

\section{\emph{mandar}}

Meaning "to send" or "to order," this word can cause problems because of other words that can get the same message across.
Most uses of "to send," for instance, work better with \emph{enviar}, and "to
order" in a restaurant is \emph{pedir} or (increasingly) \emph{ordenar}. \emph{Mandar} is
used for real orders, of the sort that generals and bosses give, and you'll
come across it a lot in conjunction with a second verb. In these cases
it works as "to send out for" or "to have done." \emph{Manda hacer unas copias} = "Have some copies made." It usually implies an order to an underling, so don't use it freely unless you are in a position either to give
orders or to take orders. If you go to Mexico, your first encounter with
\emph{mandar} may be the ubiquitous expression \emph{¿Mande?} for "What?" or
"You called?" It's considered polite, but many foreigners (especially
from other Spanish-speaking countries) seem to think it servile and demeaning. If you feel that way, you can substitute \emph{¿Cómo?} or even the
brusque \emph{¿Qué?}---but you'll hear \emph{¿Mande?} just the same.

\section{\emph{meter}}

Meter, meaning "to put (something) in," is used in a far wider
range of circumstances. It's the common way of saying "to go inside"
(\emph{Vamos a meternos} = "Let's go inside") or "to go in" (\emph{Vamos a meternos al agua} = "Let's go in the water"). It can even be used for "to get
in" a car (as in \emph{Métete al coche}), but \emph{subirse} is preferred. The reflexive
form \emph{meterse} is also a good translation for "to get involved in" or "to
get mixed up with." \emph{No te metas} = "Don't get involved." \emph{No te metas
can mi hermana} = "Don't mess around with my sister." \emph{Meterse en
líos} takes the idea further and means "to get mixed up in problems,"
"to get into trouble." If you get caught in the middle of a family
squabble and find opposing sides of the squabble looking to you for
support, you might throw up your hands and say \emph{Yo no me meto} ("I'm
not getting involved in this").

\section{\emph{notar}}

This verb is worth learning in the reflexive form (\emph{notarse}) to
express "I can see that" or "It shows." It's a nice, dry comment that
says that you, too, can perceive the obvious. If someone in the midst of
a downpour reminds you that it's the rainy season, you might respond
\emph{Se nota} ("I figured that out" or "But of course"). Adding the personal
pronouns \emph{me, te}, or \emph{se} personalizes the phrase. Your friend, screaming,
tells you she's angry. You say, \emph{Se te nota} ("So I see," literally "One
notes that in you.") \emph{Notar}, incidentally, is not a good word for "to
note something down" or "to make a note of." For that, use \emph{anotar}
or \emph{apuntar}.

\section{\emph{parar}}

"To stop." \emph{Pare el mundo, quiero bajarme} = "Stop the world,
I want to get off." \emph{¿Dónde para el tren?} = "Where does the train stop?"
And so on. \emph{Párale} is sometimes employed to say "Stop it" or "Cut it
out," as when someone is talking too much or the kids are screaming.
In most of the Americas (and even parts of Spain), the reflexive \emph{pararse}
means "to stand up," and \emph{parado} is "standing up." Travelers will
sometimes ask if there is room on a train or bus and be told \emph{Si quiere
ir parado} ("If you want to travel standing up"). \emph{Párate que nos vamos}
would be a colloquial way of saying "Get up, we're going." Learn to
distinguish between \emph{parar} and \emph{pararse} for "to stop." The reflexive
form is appropriate for stopping unassisted, whereas \emph{parar} suggests
something stopping something else. \emph{Paré el coche} is "I stopped the
car." \emph{El coche se paró} is "The car stopped."

\section{\emph{parecer}}

Meaning "to seem," this verb is more frequently encountered
than its English equivalent. One of the most common ways of conveying likes and dislikes in Spanish, in fact, is with \emph{parecer}. Here are
some typical usages: \emph{¿Qué te parece?} = "What do you think?" or
"How does that strike you?" \emph{¿Te parece?} = "Is that okay with you?"
\emph{No me parece} = "I don't like it." \emph{Me parece bien} = "Fine with me."
The reflexive \emph{parecerse} is "to look like" or "to resemble." \emph{Me parezco a mi madre}
 = "I take after my mother." "To look like" in the sense of
 "to look as if" requires \emph{parece que}, not the reflexive. \emph{Parece que va a
llover} = "It looks like (as if) it's going to rain."

\section{\emph{pasar}}

One of several options for "to happen" (see Chapter 11), \emph{pasar}
is also usually safe for most uses of "to pass," although it's a bit slangy
at the dinner table (\emph{Pásame la sal}). One use you may not be expecting:
in some Latin American countries, \emph{pasar} is the verb to use when asking "what's on" television or at the local theater. \emph{¿Qué están pasando
en el 8?} = "What's on Channel 8?" \emph{Están pasando la nueva película
de de Niro en el Cine Colón} = "They're showing the new de Niro
movie at the Cine Colon." In South American countries the verb of
choice here would be \emph{dar}, usually in the phrase \emph{estar dando}.

As a colloquial greeting, both \emph{¿Qué pasa?} and \emph{¿Qué pasó?} are
used, though individual countries tend to have a preferred form. In
most situations, \emph{¿Qué pasa?} implies that something is wrong or abnormal; this is the question to ask when there are police cars parked in
front of your house. In Mexico, where \emph{¿Qué pasó?} is the preferred
greeting, saluting someone with \emph{¿Qué pasa?} may prompt raised eyebrows and the question \emph{¿Por qué?} So much for starting a pleasant
conversation.

Using \emph{pasar} can conveys "to happen to" in the sense of "to
become of." \emph{¿Qué pasó con Juan?} = "What's become of Juan?"---that
is, why is he late? In the present tense \emph{¿Qué pasa?} with an indirect
object pronoun (\emph{me, te, le, nos}, or \emph{les}) is like asking "What's (my, your,
his, her, our, their) problem?" In fact, \emph{¿Qué te pasa?} is roughly equivalent to "What's bugging you?" In the past tense, this same construction means "What happened to (me, you, him, her, us, them)?": \emph{¿Qué
le pasó?} = "What happened to him?"---for example, why are they carrying him off on a stretcher?

\emph{Pasarse}, the reflexive form, plus the preposition \emph{de} is very
handy in expressions meaning "to go too far," figuratively speaking. \emph{Se
pasó de listo} means someone "was too clever" or "was too sneaky,"
and implies that the person got caught at it. Sometimes it can be translated as "to get carried away." When a person \emph{se pasa de listo (lista)}
with another person, it can mean that he or she is making unwelcome
sexual advances. \emph{Ese señor se pasó de listo con María} = "That guy
made a (rude) pass at Maria." The formula \emph{pasarse de} can be used with
almost any adjective or quality, positive or negative. \emph{Usted se pasa de
generosa} = "You are being overly generous." \emph{Te pasaste de imbécil} =
"You were even stupider than usual."

\section{\emph{pedir}}

"To ask," "to ask for," and the correct verb for "to order" in
a restaurant (although \emph{ordenar} is gaining ground). \emph{Pedir} should make
you think of indirect commands and the subjunctive: \emph{Pídele que se
vaya} ("Ask him to leave"). It is also used in a number of stock phrases,
quite a few of which you are liable to need in the course of your dealings in Spanish. Some common ones include \emph{pedir permiso} ("to ask
permission"), \emph{pedir perdón} ("to apologize"), \emph{pedir informes} ("to ask for
information"), and \emph{pedir ayuda} ("to ask for help"). \emph{Pedir prestado} is
the correct phrase for "to borrow," but you'll often find \emph{prestar} (see below) handier. A rule of life for some in the Spanish-speaking world is
\emph{Es más fácil pedir perdón que pedir permiso}, or "It's easier to ask forgiveness than permission."

\section{\emph{pensar}}

The verb for "to think," though often \emph{creer} (see above) is preferred. To the extent that there is a rule for distinguishing them, use
\emph{pensar} when you might use "to have been thinking" or "to be thinking
about" in English. If it's just a simple statement of opinion, use \emph{creer}.
\emph{Pienso que debes irte} = "It is my feeling that you should go." \emph{Creo
que debes irte} = "You should probably go." Sometimes the two are
interchangeable. A usage of \emph{pensar} you should become very familiar
with is \emph{pensar} plus the infinitive to mean "to plan on" or "to intend."
\emph{Pienso irme mañana} = "I plan to go tomorrow." \emph{Pienso quedarme
unos días} = "I intend to stay a couple of days." \emph{Pensar + en} is "to
think of" or "to have in mind." \emph{Pensar + sobre} (or \emph{acerca de}) works as
"to think about" or "to consider" something. \emph{Pensar + de} is "to think
of," in the sense of an opinion, and is a lot like \emph{creer}. \emph{Estoy pensando en nuestras vacaciones}
= "I'm thinking of (remembering, daydreaming about) our vacation." \emph{Todavía estoy pensando sobre} (or \emph{acerca de})
\emph{nuestras vacaciones} = "I'm still thinking about (considering, analyzing) our vacation." \emph{¿Qué piensas de nuestras vacaciones?} = "What do
you think of our vacation (so far)? Stock phrases with \emph{pensar} include
\emph{¡ni pensarlo!} ("no way," "it's out of the question"), \emph{pensándolo bien}
("on second thought"), and \emph{sin pensar} ("without thinking,"
"unintentionally").

\section{\emph{poder}}

Meaning "to be able," this is in general a predictable word.
Aside from its quirks in the past tenses (see Chapter 5), it's just a matter of mastering a few stock phrases to get a hold on \emph{poder}. One such
phrase is \emph{poder + con}, which means something like "to handle" or "to
deal with." Examples will be useful here. A student who has trouble
learning biology might lament, \emph{No puedo con la biología}. In the sports
pages, you'll often come across headlines saying things like \emph{Los Leones
No Pudieron con Los Toros} ("The Lions Couldn't Handle the Bulls"---
that is, the Bulls beat the Lions, pure and simple). A Cuban postrevolutionary chant intones \emph{Fidel, Fidel, ¿qué tiene Fidel, que los americanos no pueden con él}, which means "Fidel, Fidel, what does Fidel
have, that the Americans can't handle (defeat) him?" Depending on the
context, \emph{poder + con} can also mean "to tolerate," and in this sense is
nearly synonymous with the verb \emph{aguantar}. \emph{¿Ay, no puedo con mi hermano!} = "Arrgh, I just can't stand my brother!" A fun way of describing an extremely irritating person is to say \emph{No puede ni consigo mismo}
("He can't even stand himself"). Other phrases using poder that you'll
want on the tip of your tongue: \emph{¿Se puede?} ("May I"), \emph{Puede ser}
("Could be" or "Maybe"), and \emph{Puede que} plus the subjunctive ("It
could be that\ldots{}" or "Maybe\ldots{}").

\section{\emph{prestar}}

\emph{Prestar}, "to lend," works for just about anything you might
want to borrow, just as "to lend" does in English. Thus \emph{Préstame tu
pluma} = "Lend me your pen." (For "lending a hand," though, you
would probably use dar or echar: \emph{Oye, échame una mano}.) What is
difficult for many English speakers is to switch between "borrow"
phrases and "lend" phrases in Spanish. This sometimes leads to convoluted constructions with \emph{pedir prestado}, like \emph{"Puedo pedir prestado tu
pluma?} Much more natural in Spanish is to turn it around (i.e., saying
"you lend" instead of "I borrow") and use \emph{prestar} by itself. If \emph{Préstame}
is too tactless for your tastes, say \emph{¿Me presta?} or \emph{¿Me prestas?}: \emph{¿Me
prestas tu pluma?} In very slangy speech you might hear \emph{Presta para
acá} or \emph{Presta pa'cá} for "Hand it over" or "Give it up." The latter example has sexual overtones, as it does in English.
\emph{Prestar} is also used for "to pay attention," which, if you think
about it, is much more realistic than the English concept (we don't really "pay" attention, we just lend it out). \emph{¿Niños, presten atención!} =
"Children, pay attention!" \emph{Prestarse}, the reflexive form, can be quite
useful for "to lend oneself," though it sounds much more idiomatic in
Spanish than in English. \emph{¿Tú crees que Juan nos deje copiar en el examen? No, él no se presta a eso}. = "Do you think Juan will let us copy
his exam?" "No, he doesn't lend himself to that."

\section{\emph{quedar}}

A megaverb that you'll want to have on your side as quickly as
possible. Its most straightforward uses revolve around "to stay" or "to
remain." \emph{Aquí me quedo} is "I'm staying here" and is sometimes used
as a name for cantinas. \emph{Quédate aquí} = "Stay here." Other uses
require "to have left" in English. \emph{Soló me quedan treinta dólares} = "I
only have thirty dollars left."

A host of other expressions with \emph{quedarse}, the reflexive, are
better covered in English by "to keep." \emph{Me quedé can treinta dólares}
= "I kept thirty dollars." \emph{Quédese con el cambio} would be "Keep the
change." \emph{Quédatelo} = "Keep it." For use in shopping, \emph{quedarse} is a
lot like \emph{llevarse} (see above). \emph{Me quedo can el azul} = "I'll take the blue
one." Often quedarse suggests a final or resultant state of affairs. \emph{Me
quedé helado} is, literally, "I was left frozen" and suggests you were frozen with fear. \emph{Me quedé en blanco} is to say "I ended up blank" or "I
didn't understand that at all." If someone asks you whether you understood an explanation of the theory of relativity, you could answer, \emph{Para
nada. Me quedé en blanco}. In English slang the equivalent might
even be"I spaced." A stock phrase you should remember for personal
dealings is \emph{¿En qué quedamos?} to mean something like "What's the
agreement, then?" or "So what's the deal?" Use it toward the end of
conversations to establish clearly the next step, be it the signing of a
multimillion-dollar merger agreement or a date to sip margaritas under
the stars.

Along these same lines of final or resultant states are the everyday expressions \emph{quedar bien} and \emph{quedar mal}. Like many of the expressions using \emph{quedar}, these seem to defy a simple English translation,
but the idea is "to end up well (or badly) with someone." Their use is
similar to \emph{caer bien} and \emph{caer mal}, and often they can be translated
with "impress," though that's a little strong. "To get on someone's
good side" might come closer for \emph{quedar bien}. \emph{Se puso corbata para
quedar bien can los suegros} = "He put on a tie to get on his in-laws'
good side." For \emph{quedar mal}, an example will be more helpful than an
English equivalent. \emph{Quedé mal con él porque no lo saludé} = "I've
gotten on his bad side (i.e., he's mad at me) because I didn't say hello
to him."

Finally, quedar has a couple of common uses that you should
be alert to since they don't fall into the "stay" or "remain" categories.
It is the common word for "to fit," for clothes and everything else. Remember to use it with the indirect object pronouns \emph{me, te, 1e}, etc. \emph{Este
saco no me queda} = "This coat doesn't fit." Also, and more natural to
a Spanish speaker, \emph{Este saco me queda grande} (or \emph{chico}) = "This coat
is too big (or small) for me." \emph{Quedar} also comes into play for describing locations. As a tourist, especially, you will hear it (and can even
use it!) a lot. \emph{Perdón, ¿donde queda la plaza? Adelante, a tres cuadras}.
= "Excuse me, where is the plaza?" "Three blocks up." \emph{Queda cerca}
and \emph{queda lejos} are both handy phrases for travelers. \emph{¿Queda cerca la
plaza? No, queda lejos}.

\section{\emph{querer}}

An absolutely vital word, and one this book gives a lot of space
to so you can get it right. It means "to like" or "to want" and, with
people, "to love" or "to want" (see Chapter 4). Like \emph{poder} and \emph{saber},
this verb acts a little strangely in the past tenses (see Chapter 5). Otherwise it's trustworthy, though it's worth rehashing a few tips for using
it well. \emph{Querer} performs many interesting tricks when paired with an
adverb and in the subjunctive. Before you faint, remember that we already went over that (Chapter 6) and you survived it quite nicely---\emph{cuando quieras, donde quieras, como quieras}, and so on, We've also
gone over how to say you "want" something (using \emph{traer}) without
dragging \emph{yo quiero} into your speech habits (Chapter 2). Another option
is to use \emph{quisiera} ("I'd like"). \emph{Querer} is also, in the phrases \emph{con querer}
and \emph{sin querer}, your best ticket for handling "on purpose" and "by
accident" (Chapter 12). \emph{Mamá, metí al gato en la piscina. ¿Fue con
querer o sin querer?} = "Mommy, I put the cat in the swimming pool."
"Was it on purpose or by accident?"

\section{\emph{repetir}}

"To repeat," of course, but also "to burp" or "to provoke
burps," The proper word for "to burp" is \emph{eructar}, which covers most
every burp, while \emph{repetir} is for those little, barely perceptible, good-eatin' burps. Just thought you'd want to know.

\section{\emph{romper}}

Remember that romper is "to break intentionally": \emph{Rompí
el vaso tirándolo contra la pared} = "I broke the glass by throwing it
against the wall." \emph{Romperse} is "to break" in the sense of an accidental
act: \emph{Se me rompió el vaso cuando lo estaba lavando} = "The glass
broke (on me) when I was washing it," In this construction the literal
meaning is "such-and-such broke itself to me (or you, him, her, us,
them)." The distinction is not dogma, but you should try to stick to
it. \emph{Romper con} is "to break up with" in the sense of lonely hearts and
whatnot.

\section{\emph{saber}}

A sometimes complicated verb, \emph{saber} bears watching for its
trickery in past tenses (Chapter 5) and in contrast to \emph{conocer} (see
above). An imperfect but useful rule of thumb: use \emph{conocer} with
proper and specific nouns and \emph{saber} or \emph{saber de} with the rest of them
and most clauses. \emph{¿Conoces París?} but \emph{¿Sabes dónde comen los parisinos?} and \emph{¿Sabes de su historia?} An exception to this rule are the
names of languages, which take \emph{saber}: \emph{¿Sabes inglés?} Another more
sweeping but also far-from-perfect rule: more often than not the word
you want for "to know" is \emph{saber}. \emph{Saber} is frequently followed by verb
infinitives; \emph{conocer} never is. \emph{Saber} carries with it the idea of "to know
how," so you don't have to say \emph{saber como}. \emph{¿Sabes esquiar?} \emph{No, pero
se caerme}. = "Do you know how to ski?" "No, but I know how to
fall down."
A few of the many stock expressions using \emph{saber} include
\emph{¿Sabes qué?} ("Know what?"), \emph{No sé} ("I don't know"), \emph{¿Quién sabe?}
("Who knows?"), \emph{¡De haberlo sabido!} ("If I had only known!"), \emph{¿Yo
qué sé?} ("What do I know?" or "Don't ask me!"). \emph{Un sabelotodo} is "a
know-it-all," and a useful phrase for "as far as I know" is \emph{que yo sepa}.
Finally, and just possibly to confuse you further, \emph{saber} is also
the word for "to taste," as in how something tastes to you. \emph{La sopa
sabe bien} is "The soup tastes good." In the first person (for use after
kissing or among cannibals), the correct form is \emph{sé}, but colloquially
you might hear \emph{sepo}. \emph{¿Que tal sé (sepo)? Sabes a pepinillos agrios}.
= "How do I taste?" "You taste like pickles." For the transitive "to
taste"---that is, to taste something---you need to use probar.

\section{\emph{seguir}}

"To follow," yes, but also frequently "to continue" or "to keep
(on)." The most common formula is \emph{seguir} plus the infinitive: \emph{Sigue
viniendo} ("He keeps coming"), \emph{Sigues comiendo} ("You keep eating")
\emph{Sigo llorando} ("1 keep crying"), and so on. \emph{Seguir} also works to translate a lot of the uses of "still" in English. In fact, using \emph{seguir} frequently sounds more natural in Spanish than using \emph{todavía}, which
is what native English speakers tend to resort to: \emph{Sigue creyendo en
Santa Claus}. = "She still believes in Santa Claus." \emph{Sigo enfermo} =
"I'm still sick." For the negative, you can use \emph{seguir + sin} and sound
very "Spanish" indeed: \emph{¿Sigues sin creer en Dios?} = "Do you still not
believe in God?" \emph{Seguir sin} and \emph{seguir con} are also useful constructions with a noun tacked on instead of the infinitive. \emph{Sigue sin trabajo}
= "He still doesn't have a job." A few odds and ends to take note of:
\emph{Síguele} is a handy phrase for "Keep it up" in both its genuine and
ironic senses; \emph{¿Quién sigue?} = "Who's next?"; \emph{¿Cómo sigues?} as a
greeting means "How are you getting along?" and implies that a person
has been sick or afflicted by some sort of trouble, even if it's only Spanish grammar exercises.

\section{\emph{sentir}}

Sentir means "to feel" and is a transitive verb used with direct
objects---that is, things you feel. \emph{Sentirse}, the reflexive, is used with
adjectives to express how you feel. Thus \emph{Siento frío} is "I feel cold" but
\emph{Me siento bien} is "I feel fine." \emph{Siento asco} is "I feel nauseous," and
\emph{Me siento mal del estomago} is "I feel sick to my stomach." \emph{Lo siento},
keep in mind, is "I'm sorry," and \emph{Lo siento mucho} is "I'm very sorry."
In case you need to explain further, you need only add \emph{haber} and a past
participle, dropping the \emph{lo}. Try to learn a couple of these by heart and
keep them at the ready: \emph{Siento haber llegado tarde} = "Sorry I'm
late." \emph{Siento mucho no haber podido ir} = "I'm very sorry I couldn't
come (or go)."

\section{\emph{ser}}

Except when you need \emph{estar} (Chapter 5), \emph{ser} is used for "to
be." It is the verb of permanent states, of the way things are, of telling
it like it is. There are a billion or so expressions using \emph{ser}, but most
of them are predictable and translate as "to be" in English. A few to
watch: \emph{¿De quién es?} = "Whose is it?" \emph{¿De qué es?} = "What is it
made of?" Like \emph{querer}, \emph{ser} in its subjunctive forms can be employed
to form "-ever" words or to say "any"; \emph{cuando sea, donde sea}, and so
forth. This is covered in Chapter 6, but a few examples here can't hurt.
Usually you employ \emph{que sea} after someone has already fed you the antecedent. It's better not to repeat the antecedent but think fast to get
the right gender. If someone asks, "What brand of beer do you want?"
you can answer \emph{La (marca) que sea}. "In what restaurant do you want
to eat?" \emph{En el (restaurante) que sea}. It's worth a little extra work to get
these expressions down pat, since otherwise you'll be inclined to say
ghastly things like \emph{Cualquier hora que quieres}, instead of the pure and
lilting \emph{Cuando sea}, for "Anytime."

\section{\emph{servir}}

"To serve," yes, but how often do you say "to serve" in an average day? "After I serve dinner, if it serves you, m'lord, I'll serve you
poisoned coffee and it'll serve you right!" In Spanish \emph{servir} is much
more commonly heard for "to work" in the sense of "to function."
\emph{No sirve mi teléfono} is "My phone doesn't work" \emph{¿Para qué sirve?} is
"What is it used for?" or even "What good is it?" When \emph{servir} is used
for "to serve," it is often dressed up in the stock phrase \emph{¿En qué le
puedo servir?} ("May I help you?"). \emph{Sírvase} (or \emph{Sírvete}), finally, can be
used for "Help yourself," but a Spanish speaker would probably say
\emph{Tome lo que quiera} instead.

\section{\emph{soler}}

There is no English equivalent for this word among the verbs,
which is odd considering how often you'll find yourself needing it. \emph{Soler} (meaning "to be in the habit of," "to be accustomed to") is useful to
describe something you usually do and is stuck before another verb
in the infinitive, making it a snap to use. \emph{Suelo comer a las dos} = "I
usually eat at two." \emph{Suele ir al cine saliendo del trabajo} = "He usually goes to the movies after work" See Chapter 12 (under "Usually")
for other ways to handle this concept.

\section{\emph{sonar}}

"To sound" or "to ring." \emph{Suena el timbre} = "The doorbell's
ringing," \emph{Eso suena dudoso} = "That sounds dubious." \emph{Sonar} is also
the verb to use to say "to ring a bell," as in what happens in your
memory when something sounds familiar. "Do you know Juan Pérez?"
\emph{No, pero el nombre me suena} ("No, but the name rings a bell"). \emph{Sonarse}, incidentally, is "to blow one's nose," and \emph{sonarse a alguien} is
"to smack someone."

\section{\emph{soñar}}

Meaning "to dream," the verb is used with \emph{con}. \emph{Sueño con serpientes} = "I dream about snakes." Here's something worth remembering: \emph{sueño}, the noun form, means both "dream" and "sleep." \emph{Una
falta de sueño} = "A lack of sleep." \emph{Soñado}, the adjective form, is a
fun word for "dreamy" or "ideal": \emph{la playa sonada} = "the beach of
your dreams."

\section{\emph{subir}}

The polar opposite of \emph{bajar} (see above), \emph{subir} works conversely
for gaining weight, getting on a bus, and so on.

\section{\emph{tardar}}

\emph{Tardar} works well for "to take" in time expressions. \emph{Durar}
(see above), on the other hand, generally works better for "to last."
\emph{Tarda el tiempo que quieras} = "Take as much time as you want."
\emph{¿Cuánto tardará en venir?} = "How long will he take in coming?"
Since we sometimes use "to take" and "to last" loosely in English,
there can be confusion in the correct Spanish choice. \emph{El avión tarda
media hora en venir de allá a acá}. = "The plane takes half an hour
to get here from there." \emph{El vuelo dura media hora} = "The flight lasts
half an hour." As a rule, use \emph{durar} whenever "to last" could work, and
use \emph{tardar} otherwise. \emph{Tardar} also means "to take time" with the suggestion of "to take too much time," "to dally," "to be late." \emph{Tardé en
llegar porque había mucho tráfico} = "I took a long time getting here
(I'm late) because there was a lot of traffic." \emph{El tren está tardando en
llegar} = "The train is taking too much time (is late) arriving."

\section{\emph{tener}}

A megaverb, \emph{tener} is almost worthy of a chapter of its own.
Many of its uses are predictable renditions of "to have" in English, including the indispensable \emph{tener que} for "to have to": \emph{Tengo que irme}
= "I have to go." It can also be used by itself in the negative to say "I
don't have any" when the antecedent is understood. \emph{Dame dinero. No
tengo}. = "Give me money." "I don't have any." \emph{¿Dónde está su boleto?
No tengo}. = "Where's your ticket?" "I don't have one." Another common use of tener with an implied complement is the question \emph{¿Qué
tienes?} (or \emph{¿Qué tiene?}), which can translate as "What's your (or his,
her) problem?" or "What's the matter with you (or him, her, it)?" Another example: "I don't like the house they've picked out for us." \emph{¿Qué
tiene?} ("What's wrong with it?")

Where you will have to pay special attention to \emph{tener} is in the
thousand and one expressions in Spanish that use \emph{tener} plus a noun for
what in English would be "to be" plus an adjective. In English, for instance, you "are cold," whereas in Spanish you "have cold." Common
examples of this construction include \emph{tener hambre} ("to be hungry"),
\emph{tener sed} ("to be thirsty"), \emph{tener frío} ("to be cold"), \emph{tener calor} ("to be
hot"), \emph{tener sueño} ("to be sleepy"), \emph{tener paciencia} ("to be patient"),
\emph{tener cuidado} ("to be careful"), \emph{tener razón} ("to be right"), \emph{tener prisa}
("to be in a hurry").
There are also another thousand and one expressions, many of
them quite colloquial, using \emph{no tener}. Here are some you should learn:

\bsk

\indu \emph{No tiene sentido}. = "It doesn't make sense."

\indu \emph{No tiene caso}. = "There's no point" or "What's the point?"
(indicating futility)

\indu \emph{No tiene chiste}. = "It's boring" or "What's the point?" (indicating insipidness)

\indu \emph{No tiene (nada) que ver}. = "That has nothing to do with it"
"That's irrelevant."

\indu \emph{No tiene vergüenza}. = "He (or she) is shameless."

\indu \emph{No tiene lógica}. = "It's illogical" or "It doesn't make sense."
(indicating incredulity)

\indu \emph{No tiene ni pies ni cabeza}. = "I can't make heads or tails of it."

\indu \emph{No tiene en donde caerse muerto}. = "He (or she) is flat
broke."
(literally, "doesn't even have anywhere to drop dead")

\indu \emph{No tiene remedio}. = "There's no way out" or "There's nothing to be done."

\indu \emph{No tiene} (\emph{ni la menor} or \emph{ni la más mínima}) idea. = "He (or
she) hasn't got the faintest idea" or "He (or she) hasn't
got a clue."

\section{\emph{tirar}}

The common verb for "to throw," though \emph{aventar} and \emph{arrojar}
are also used quite a lot regionally. \emph{Tirar} also has the implication of
"to throwaway" or "to throw out." \emph{Tira esa basura} = "Throw that
worthless thing out." It is also used for "to knock over," as in \emph{Tiré el
vaso} ("I knocked over the glass"). To convey "to toss," as in "Toss me
a pen," you could use \emph{aventar} (in the Americas) or \emph{echar} or even \emph{tirar},
but mostly you wouldn't use anything because it's considered rude
in the Spanish-speaking world to toss things (see Chapter 2). Unless
you're especially keen on seeing a particular object in flight, stick to
\emph{Préstame} (see \emph{prestar} above).

\section{\emph{tocar}}

"To touch," of course, but also "to play" (a musical instrument) and "to knock" or "to ring" (at someone's door). In Casablanca
Bogie would have told the pianist, \emph{Tócala, Sam}. An extremely common use of \emph{tocar} that is often glossed over in textbooks is for "to experience" or "to be one's turn." A simple and good translation is elusive,
but examples will get the point across. \emph{Me toca}. = "My turn." \emph{¿A
quién le toca?} = "Whose turn is it?" \emph{A mi no me toca decirle}. = "It's
not up to me to tell him." \emph{Al dueño le toca arreglar la casa}. = "It's
up to the owner to fix the house." Sometimes the best translation involves "to get": \emph{¿A quién le toca la última rebanada?} = "Who gets the
last slice?" \emph{A ti te tocó la más guapa de las hermanas}. = "You got the
prettiest of the sisters." \emph{A mi me tocó el más feo de los hermanos} =
"I got (stuck with) the ugliest brother." \emph{No me ha tocado verlo en concierto} = "1 haven't had (gotten) a chance to see him in concert." And
so on. A final note on \emph{tocar}: its past participle, \emph{tocado}, is a common
synonym for \emph{loco}---as in the English sense of being slightly "touched"
in the head. Children often adapt it to \emph{toca-toca}, translating perhaps as
"cuckoo."

\section{\emph{traer}}

\emph{Traer} is straightforward for "to bring," except when "take"
and "bring" (\emph{llevar} and \emph{traer}) get mixed up (see Chapter 5).

\section{\emph{tratar}}

"To treat," of course, but far more commonly encountered
with \emph{de} and meaning "to try." For some reason, though, many Spanish-English dictionaries refuse to acknowledge this fact, leaving the student to choose among \emph{ensayar, procurar, intentar}, and \emph{pretender}.
These are all worthy and acceptable verbs, of course, and someday you
might even want to learn them. But for now, remember: \emph{tratar de} =
"to try." \emph{Traté de dormir} = "I tried to sleep." \emph{Tratamos de llamarte}
= "We tried to call you." \emph{Trata de venir antes de las once} = "Try to
come before eleven." Only when you're using "to try" in the sense of
"to sample" or "to test" should you abandon \emph{tratar de}; the correct verb
here is \emph{probar}. An awkward but illustrative example: \emph{Trata de probar
el vino blanco 1985}. = "Try to try (i.e., make an effort to sample) the
1985 white wine."

\emph{Tratarse}, the reflexive form, is a useful verb for "to have to do
with" or "to be about." \emph{¿De qué se trata?} is the common way of asking
"What's it about?" (in reference to a film, a book, a scuffle, an argument, and the like). "To treat" in the sense of "to pay for someone
else" is usually handled by \emph{invitar}. \emph{Yo invito} = "I'm treating."

\section{\emph{valer}}

Meaning "to be worth," this verb is frequently encountered
in the stock phrases \emph{vale la pena} ("it's worth it") and \emph{no vale la pena}
("it's not worth it"). In addition, you may find \emph{valer} handy for asking
prices. \emph{¿Cuánto vale?} = "How much is it?" Another good use of valer,
preceded by \emph{más}, is to translate English phrases that use "better" or
"had better." \emph{Más te vale irte} = "You'd better get out of here." \emph{Más
vale preguntar} = "We'd better ask." And there's the old standby \emph{Más
vale tarde que nunca} ("Better late than never"). In Spain and less so
elsewhere, \emph{vale} by itself is a common interjection for "all right" or
"okay." In Mexico, especially, \emph{valer} with an indirect object pronoun
is a somewhat crude way of saying "couldn't care less." \emph{Me vale} = "I
couldn't care less." \emph{Le vale} = "He doesn't give a damn." Its crudeness
comes from the fact that it's a shortened and therefore euphemistic
form of another phrase, which you'll have to read about in Chapter 10.

\section{\emph{venir}}

Meaning "to come," naturally, and sometimes difficult to distinguish from "to go" (see Chapter 5). This verb has some common
but unexpected uses, as in \emph{No viene al caso} ("That's beside the point").
\emph{Que viene} is especially worth learning in stock phrases like \emph{la semana
que viene} ("next week") and \emph{el año que viene} ("next year"). \emph{Venirse},
the reflexive form, is usually an innocent intensifier for \emph{venir}, but it
has sexual overtones in some countries and should be used with care.

\section{\emph{ver}}

As "to see," this is a pretty straightforward verb. It can sometimes be confused with \emph{mirar}, since \emph{ver} also works as "to look at" in
many cases where you might be tempted to use \emph{mirar}: \emph{Ese señor se me
queda viendo} = "That man keeps looking at me." \emph{Estoy viendo tus
discos} = "I'm looking at your records." \emph{Mirar} would work fine in the
first example, but in the second would suggest you are gazing at the
records as if waiting for them to do something. An extremely common
expression is \emph{A ver}, which is simply "Let's see\ldots{}" but which is used
mainly to buy time while you think of a clever response. \emph{Vamos a
ver} means the same but is less common as an interjection or "crutch
word." \emph{Tener que ver con} is the easiest way to translate "to have to do
with." \emph{No tengo nada que ver con el asunto} = "I have nothing 'to do
with this business." \emph{No tiene que ver} = "That's irrelevant."

\emph{Ver} is sometimes used to express an opinion, in the sense of
how you "see" or "size up" a problem. \emph{La veo difícil} = "It looks difficult to me." \emph{Verse}, the reflexive, covers almost all uses of "to look"
that refer to the appearance of something or someone. You should etch
the phrase \emph{se ve} onto the end of your tongue and have it ready for such
common utterances as \emph{Se ve bien} ("It/he/she looks good") \emph{Se ve difícil}
("It looks difficult"), \emph{Se ve bonito} ("It looks nice"), and so on. \emph{Se ve que}
plus a clause is an easy way to communicate "You can tell that\ldots{}" or
"It's obvious that\ldots{}" \emph{Se ve que no han cambiado el agua en la piscina} = "You can tell they haven't changed the water in the pool." \emph{Se
ve que son grandes amigos} = "You can tell that they're great friends."

\section{\emph{volver}}

In the sense of "to return" or "to come back," \emph{volver} is interchangeable with \emph{regresar}. \emph{Volver} has another common use that you
will want to learn, though---one that \emph{regresar} does not share. \emph{Volver}
plus \emph{a} plus an infinitive is frequently found as a substitute for "to repeat" or "to do again." It will take some practice before you start to
use it properly---and as an alternative to \emph{otra vez}---but it's worth the
extra effort. Some typical examples of when to substitute: \emph{Gracias,
vuelvo a llamar más tarde} = "Thanks, I'll call back (again) later." \emph{Si
vuelves a pedírmelo, no te lo voy a dar} = "If you ask me for it again,
I'm not going to give it to you." \emph{Vuelve a intentar} = "Try again."
\emph{Volver} cannot be used transitively---to "return" a book to the library, for
instance, or "to give back" a borrowed item. Use either \emph{regresar} or \emph{devolver} in these situations. \emph{Devuélveme a mi chica} ("Give me back my
girl"), for instance, was the name of a pop song and movie of a few seasons ago. \emph{Volverse}, finally, is one of the common ways to handle "to
become" (or "to get"). Skip ahead to Chapter 11 for details.

\chapter{Cranking Up Your Spanish}

Not long ago, I owned a somewhat antiquated motor vehicle
that in fits and starts, quite literally, became dear to me. It wasn't
much to look at---antique buffs generally looked away instead---but
it more often than not got me where I was going. The car's main flaw,
and what finally led me to get rid of it, was its temperament on chilly
mornings. Instead of responding to the turn of the key with a loud
growl and churning pistons, it tended to respond with a weak series of
hiccups and the grinding of metal. Once started, it sputtered, stuttered,
gurgled, and spit. Only after a few minutes of this litany did it begin to
growl a little roughly, then purr hesitantly, then hum.

Beginning Spanish speakers remind me of that car. Once
warmed up, they chatter along merrily, constructing elaborate clauses
and communicating their meaning. But getting warmed up is another
story. Frequent false starts, "ums," hiccups, and the grinding of gray
matter punctuate their speech, and English interjections pop in and
out. It is obvious that they are still thinking in English and trying to
translate, rather than thinking in Spanish.

What beginning students often lack are the appropriate words
and phrases to start a sentence smoothly. As in English, these words
don't always contribute a lot to the meaning of the sentence, but they
do give the brain a chance to warm up and kick into gear. Skipping
over them can make speech sound stilted and abrupt-sometimes
even impolite or hasty. Even when you are past the beginner stage, you
should pay attention to your "sentence starters." Learn a few well, get
accustomed to using them, and your Spanish will sound more natural,
flow more easily, and hum along to its desired destination.

\section{\emph{bueno}}

%BUENO
Bueno means nothing more than "well," but it is nearly indispensable to good spoken Spanish. It simply prefaces comments and tends
to tone down their hastiness or prepare the listener for a message. The
difference between \emph{Ya me voy} and \emph{Bueno, ya me voy} is subtle but noticeable. The first is "I'm going now" and comes across as a sort of dare
to anyone who might try to stop you, while the second suggests "Well,
I think I'll be going now" and sounds more laid-back and more polite,

\emph{Bueno} often insinuates a "stop" into an ongoing conversation
or activity. You might sit down to chat with friends for five or ten minutes and then simply say \emph{bueno} to announce your imminent departure. Or you could sit down to talk business with colleagues and, after
a few minutes of small talk and niceties, signal the beginning of the
business talk with a \emph{bueno}. If the person speaking before you at a conference decides to launch a vitriolic attack against everything the audience holds dear, you might begin your presentation with a sighed
\emph{bueno}; here it would convey "anyhow" or "as we were saying" or
even "whew!" Usually it's much simpler: a pause while you get your
thoughts together, a signal that you have the floor, an acceptance of
what has gone before but a statement that you intend to begin something different.

\section{\emph{pues}}

%PUES
Similar to buena but simpler still. \emph{Pues} means "well, then"
but often replaces an English "ummm" or "let's see" It means, in
short, that you are thinking about your response and would like your
listener to hold on for a few seconds while you try to bring order to the
mild chaos reigning in your head. Often, \emph{pues} is accompanied by eyes
rolling about, looking up, or darting off to some distant object and
other telltale and foolproof signs of intense mental activity. "Wanna
go to a movie?" \emph{Pues}\ldots{} ("Will it mean staying out late? Can I find a
baby-sitter? What good movies are around? Shouldn't I be working on
my taxes? Can I afford a movie these days?")\ldots{} \emph{sí, vámonos}.

You'll sometimes hear \emph{pues} reduced to a shortened, or just
more mumbled, form: \emph{pos}, for instance, or \emph{pus} or \emph{pos'n}. In parts of
Spain you'll hear an emphatic \emph{¡pue!}; in parts of Mexico a barely audible \emph{pss}. When a foreigner talks this way, it can sound very natural
or very stupid, depending on the stigma (upper-class snob, lower-class
slob, rural hick, urban gang, etc.) that accompanies a form in each
given locale. Listen to how people around you speak, then contract
your \emph{pues} if you're so inclined.

\section{\emph{entonces}}

%ENTONCES
It means "then," but which "then"? Usually, \emph{entonces} crops
up in storytelling situations, as when you are relating what happened
in chronological order. \emph{Entonces yo llamé a la policía}. \emph{Entonces Juan
escondió la tu-ya-sabes-que}. \emph{Entonces llegó la senora de los rifles}\ldots{}
But you will also run across it in present-time situations where it
translates more like "so" or "so then." \emph{Entonces, ¿qué hacemos?} = "So
what do we do?" \emph{Entonces, yo voy par la Carretera 1} = "So then (what
you're saying is that) I take Highway 1." \emph{¿Entonces no van a querer
pizza?} = "So you're not going to want any pizza?" \emph{¿Entonces?} and \emph{¿Entonces qué?} as questions mean "And next?" or "Then what?" Both are
very common and, when reduced in a slangy context, can come out
sounding like \emph{¿'tons qué?} ("So what'll it be?" or "Now what?").
You
may even hear \emph{¿'tons?} by itself, but remember that these lazy shortcuts
have their enemies among the more refined speakers of the language.

\section{\emph{luego}}

%LUEGO
\emph{Luego} is a lot like \emph{entonces} in that it means "then" and is interchangeable with \emph{entonces} in many cases. \emph{Luego} generally tends
to be reserved more for time constructions and often works better as
"later" than as "then." A subtle distinction: \emph{entonces} implies "and
then (as a result)," whereas \emph{luego} is often simply "and then (in sequence)." \emph{Primero me voy a bañar, luego voy a la tienda, y luego a tu
casa} = "First I'm going to take a bath, then go to the store, and then
to your house." It works as "later" when you're putting something off
into an indistinct future. The most familiar example of this is \emph{Hasta
luego} for "See you later." You can also say \emph{Luego te hablo} ("I'll call
you later"), \emph{Termínalo luego} ("Finish it later"), \emph{Luego voy} ("I'll go
later"), \emph{Luego te digo} ("I'll tell you later"), and so on. Note that regionally \emph{luego luego} (paradoxically) means "right away" or "immediately,"
and not "later, later."

\section{\emph{a propósito}}

%A PROPÓSITO
"By the way." As in English, this usually announces a change
of subject or indicates that the speaker has been reminded of something else that needs relating. \emph{Son las nueve. A propósito, ¿no ibas a
llamar a Juan?} = "It's nine o'clock. By the way, weren't you going to
call Juan?"

\section{\emph{por cierto}}

%POR CIERTO
This also means "by the way." If there is a difference, this
phrase is better for changing the subject when no apparent cause for
it exists. In other words, if someone is talking about the philosophical
implications of modern cinema, and you're hungry and want to talk
about sandwiches, you could say \emph{Por cierto, ¿nadie tiene hambre?}
You'll still get a funny look, but you'll be more correct than if you use
\emph{a propósito,} which denotes "in reference to that."

\section{\emph{es que}}

%ES QUE
This is an extremely useful sentence starter, and one you
should learn and prepare to embellish. You won't appreciate the importance of \emph{es que} unless you stop and think of all of the utterly useless
words you tack on to the beginning of your conversational speech in
English. Rarely, for instance, do we say "He doesn't want to go to the
store." We say something like, "No, well, as it happens, he doesn't
want to go to the store." Okay, I'm exaggerating, but pay close attention to how you speak and you'll see I'm not exaggerating much. Are
you hungry? "No, it's just that I've got this craving for a Milky Way."
Did you sleep well? "No, well, sort of, but the thing was, there were
so damn many dogs howling that I kept waking up." It may not read
pretty, but this is the way we speak much of the time. In Spanish you
can add useless words and phrases to your speech, too, and sound more
fluent in the process. The best all-purpose add-on is \emph{es que}, which can
be used in all of the comments used above. \emph{No, es que no quiere ir a
la tienda. No, es que se me antoja un Milky Way. No, bueno, mas 0
menos si, pero es que estaban ladrando tantos malditos perros que
me seguia despertando}.

In these examples, es que introduces a negation or qualification of a preceding statement. It doesn't always, but it does require a preceding statement of some sort to build from. Once you get the hang of es
que, you will find that just about any preceding statement at all will do:

\bsk

"Why is Juan unhappy?"
Es que esta enamorado de Maria.
"Why doesn't he tell her?"
Es que ella esta hablando con ese tipo.
"And when she's finished?"
Es que Juan tiene miedo.
"Of what?"
Es que ese tipo esta muy grandote.
CRANKING UP YOUR SPANISH
"So?"
Es que es e1 esposo de Maria.
"Ah. So hed better forget about Maria ..."
Es que no puede.
" ... and find someone else."
Es que no 1e gustan las otras.
"So he's got a problem."
Es que estti loco.
"And unhappy."
Es que estti enamorado de Maria . ..

\bsk

And so on.

\section{\emph{lo que pasa es que}}

%LO QUE PASA ES QUE
Just like es que, only more so. This is a very typical, very
natural, very unnecessary phrase to add on at the beginning of your
thought to make yourself sound more fluent. It can be used in the
same instances that es que can be used-which is to say practically
anytime. Why is Juan unhappy? La que pasa es que estti
enamorado ... and so on.

\section{\emph{la verdad es que}}

%LA VERDAD ES QUE
This means something like "the truth of the matter is, "but
you'll hear it-and need to use it-more often than that. In fact, this is
often a good translation for "actually," which many English speakers
use to start a sentence and which, as beginning Spanish speakers, they
tend to translate as actua1mente. Sadly, actua1mente doesn't mean
"actually" but "at present" or "currently." So you're left with 1a verdad es que or some other substitute. "Everything okay?" Bueno, 1a
verdad es que 1a sopa es un poco salada ("Well, actually, the soup is a
little salty"). "Can I drop by later tonight?" Pues, 1a verdad es que pensaba ir a1 cine ("Well, actually, I was thinking of going to a movie").
Note that 1a verdad es que, like "actually," is frequently used as a polite way to break some bad news or indicate disagreement.

\section{\emph{resulta (que)}}

%RESULTA (QUE)
Here's the word you've been wanting for "it turns out (that)."
Fuimos a1 cine, pero resu1ta que cierran los viernes = "We went to
the movies, but it turns out that they close on Fridays," Creia que
Juan era un sinvergiienza, pera resulta que es muy amable" = "I
thought Juan was a jerk, but it turns out he's a nice guy." Often it is
used to start off a sentence, especially in response to questions and
quizzical expressions. You walk in, empty-handed, after going to the
store. Your mother looks at you strangely. You say, prompted or unprompted and a little peeved, Resulta que ya no aceptan efectivo en
las tiendas ("It turns out that they don't accept cash anymore in the
stores"). Resultar que can also be used for "to work out" in the sense
of a mathematical result. Resulta que nos debes dos mil pesos = "It
works out that you owe us 2,000 pesos."

Resulta without a que clause and with an adjective or a noun,
and sometimes with ser thrown in for good measure, is also a common
construction. Usually, it still translates well as "to turn out." Result6
(ser) un desastre su fiesta = "His party turned out to be a disaster."
Result6 (ser) agradable la visita = "The visit turned out to be quite
pleasant."

\section{\emph{que}}

%QUE
You know of course that que is a key element of speaking
Spanish, and you even know where and when to use it-most of the
time. But if you are like most native English speakers, you will have
to train yourself to use que when answering questions that start with
ique~ "That" is not always used this way in English, though sometimes it is. Say your Spanish professor calls you at home, and after the
call your brother asks, iQue pas6~ ("What's up?"). Que manana va a
haber un examen ("We're having an exam tomorrow"), you would answer. Or say the terrorists have taken hostages and sent a list of demands. Trying to read over the police chief's shoulder, you ask, iQue
quieren? He would answer, Que liberemos a cien prisioneras ("For us
to free a hundred prisoners").

Another case in which "that" is rarely used in English but que
almost always is in Spanish is when you repeat something:

\bsk

Juan no estd en su casa.
iQue dices~
Que Juan no estd en su casa.

\bsk

In some cases, as when repeating a direct command, you'll have to
change from the imperative to the subjunctive (indirect command) the
second time around. This extra hurdle, in fact, may explain why many
foreigners don't take the trouble to learn this use of que:

\bsk

Vete a tu casa. = "Go home."
iQue dicest = "What did you say?"
Que te vayas a tu casa. = "I said for you to go home."

\bsk

Note that the police chief also used the subjunctive as he transformed
a direct command into an indirect command above.

\section{\emph{así que}}

%ASÍ QUE
Generally, this is the phrase you need to translate "so" at the
start of a sentence. "So you wanna be a rock 'n' roll star?" would be
expressed as iAsi que quieres ser una estrella de rockt "So you're
really leaving me?" would be iAsi que de verdad me vas a deiart And
so on. Note that asi que does not, however, mean "So what?" For that,
use iYquet

\section{\emph{fíjate (que)}}

%FÍJATE (QUE)
Again, not a phrase that will boost the intellectual content of
your comment, but one that will help you sound more fluent. It means
something like "Look, ..." and is used to call someone's attention to
something. For some reason, I associate it with reading a newspaper
and passing on bits of news to someone sitting nearby. It's used more
than that, of course, but that's a good example of how it should be
used. Fijate que van a cerrar la autopista manana = "Look, they're
going to close the freeway tomorrow." Fijate que Robert Redford usa
los mismos zapatos que til =. "How about that! Robert Redford wears
the same shoes as you." Sometimes it introduces a note of skepticism
into what you are relating. Fijate que el presidente dice que ya no hay
crisis econ6mica = "Whaddya know? The president says there's no
more economic crisis."

Fijate que plus the subjunctive means "make sure" and can be
used instead of the barbarism checar to say "to check." Fijate que este
bien cerrada la puerta = "Make sure the door's closed all the way."
Fiiate que no tenga gusanos antes de comerlo = "Check to make sure
it doesn't have any worms before you eat it."

Fijate without the que is simply "watch out" or "pay attention." Parents tend to use this phrase to scold children who mindlessly
wander into puddles, for instance. iFijate! or iFijate donde caminas!
They also might use it to get a child to concentrate on an explanation
of a math problem, say. Fijate, la dos pasa para alIa = "Look, the two
108 BREAKING OUT OF BEGINNER'S SPANISH
goes over there." Finally, if you want to be as rude as the klutz who
just bumped into you on the sidewalk, you can say iFijate, imbecil!

\section{\emph{mira}}

%MIRA
Meaning "Look!" or "Look here/' this imperative can be used
literally liMira! iNa trae ropa!) or figuratively (Mira, yo no quiero
problemas). In the figurative sense, it almost always is used to set the
record straight. Mira, yo nunca dije que no pudieras irte = "Look,
I never said you couldn't go," If you're keen on sounding macho, ies
also a good way to lead off a verbal assault, much like "Look here, ...
"Mira, a te largas a llama a la policia = "Look here, either you get
lost or I call the police."

\section{\emph{haz de cuenta que}}

%HAZ DE CUENTA QUE
This phrase is mostly to be found in the world of kids, but
grown-ups can use it, too. The best translation among kids is "Make
believe ... ,n as in "Make believe JIm Spiderman and you're Godzilla."
Since we grown-ups don't say things like that, in our mouths haz de
cuenta could translate as "Imagine .... "Haz de cuenta que sales del
bano y hay un alacrdn en tu toalla, = "Imagine that you get out of the
bath and there's a scorpion on your towel" Haz de cuenta que ems el
presidente, ique harias? = "Say you were the president. What would
you do?" It can also be used to cover a lot of uses of "to pretend,"
which pretender is notoriously unfit to do. "Pretend I'm not here" =
Haz de cuenta que no estoy. The formal form of this phrase, for the
record, is haga de cuenta, while in the plural it is hagan de cuenta.
Note also that sometimes it is used reflexively: hazte de cuenta, hdgase de cuenta, hdganse de cuenta.

\section{\emph{ni modo que}}

%NI MODO QUE
This expression isn't used everywhere in the Spanish-speaking
world, but in Mexico, for instance, it's so common that you'll wonder
how the other countries survive without it. It's a very handy expression, and one that can be considered "advanced" for beginners-especially since it calls for the subjunctive to follow it. But if you're feeling
ready to flex some Spanish muscles, give it a go. Ni modo que, more or
less literally, means "no way that." I say this reluctantly, because ni
modo definitely does not mean "No way!" as in "No way, Jose" (use
para nada for that), But ni modo que as a sentence starter can translate as "No way that, .. ," Better translations might be "Like hell ..."
or "You can't expect ...."

An example will help here. On your tax return you claim that
you made only two dollars the previous year, despite working full time
as an investment banker. A friend questions the ethics of your claim
and you say, jNi modo que 1es diga 1a verdad! ("No way I'm going to
tell them the truth!"). A few months later you get a call from an IRS
agent who says she found some irregularities in your return and wants
to talk to you. You grudgingly make an appointment. Ni modo que me
vaya a Mexico ("Well, I can't just run off to Mexico"), you shrug. The
agent chats with you, then a prosecutor chats with you, and then all of
a sudden a judge is chatting with you. Ni modo que te dejemos libre
("Well, we can't just let you go free"), he says. You make bail and decide to visit a travel agency, where you buy a one-way ticket to Cancun. Ni modo que vaya a prisi6n ("Like hell I'm going to jail"), you
say. Months later, on the beach in Cancun, someone asks you why you
came to Mexico. Pues ni modo que me quedara ahi, con tanto crimen
que hay ("Well, you couldn't expect me to stay there, with all that
crime"), you respond.

\section{\emph{menos mal que}}

%MENOS MAL QUE
Here's an extremely useful sentence starter that you should
learn and use. The best translation is "good thing" or "it's just as
well," and you'll find yourself needing it more and more as you become
comfortable with it. Menos mal que trajiste paraguas, porque va a
llover = "Good thing you brought an umbrella, because it's going to
rain." Menos mal que no fuiste, estuvo muy aburrida = "Just as well
you didn't go, it was very boring." Often it is used by itself, just like
the English comments "Good thing" or "Just as well." Juan no pudo
venir. Menos mal. Me cae mal. = "Juan couldn't come." "Just as well.
I don't like him." Este tren no para hasta San Jose. Menos mal.
"This train doesn't stop until San Jose." "Good,"

\section{\emph{lo bueno, lo malo, lo increíble, lo peor}, etc.}

%LO BUENO, LO MALO, LO INCREÍBLE, LO PEOR, ETC.
In English, it is common to begin a sentence by referring to an
unspecified "thing." It may not be glamorous, but we do it. The good
thing is that in Spanish there is an equivalent construction; the bad
110 BREAKING OUT OF BEGINNER'S SPANISH
thing is that it is done differently; the only thing to remember is the
word 10.

La plus an adjective is the formula in Spanish for "the good
thing," "the bad thing," "the only thing," and similar constructions:
La bueno, 10 malo, 10 unieo, etc. This formula saves you from ungainly constructions such as 1a buena eosa or 1a eosa mala, which
aren't natural to Spanish. Once you get the hang of this formula, I
promise that you will wonder how you spoke Spanish so long without
it. The usual formula is 10 + adjective + es que. Here are some of the
common adjectives to make use of this formula:

\bsk

10 buena = "the good thing"
10 malo = "the bad thing"
10 unieo = "the only thing"
10 difiei1 = "the hard thing"
10 pear = "the worst thing"
10 mejor = "the best thing"
10 raro, 10 extraiio = "the strange thing"
10 ehistoso = "the funny thing"
10 inereib1e = "the amazing thing"
10 (mas) absurdo = "the crazy thing"

\bsk

The formula es + 10 + adjective is also a handy one for quick and
fluent comebacks, of the sort that are dealt with in the next chapter.
For example, Es 10 raro = "That's the strange thing (about what has
just happened)."

\section{\emph{a ver}}

%AVER
Either by itself or followed by a clause, a ver is a useful sentence starter that translates well as "Let's see .... "By itself, it is used
almost as an interjection. "My computer is acting up." A ver ("Let
me have a look"). "I can turn water into wine." A ver ("Let's see you
do it"). "Do you have change for a dollar?" A ver ("Hold on, let me
check"). The clauses that usually follow A ver begin with que or si
and translate as "Let's see if" or "Let's see what." Often it provides an
easYJ painless translation for "to check," similar to fiiate (see above)
but usually implying that the speaker will help with the checking. A
ver si 1a puerta esta bien eerrada = "Let's check and see if the door's
closed." Aver si ya regresaron = "Let's see if they're back yet." A
ver que haeen los niiios = "Let's see (check on) what the kids are doing." When writing a ver, take care not to confuse it with its homonymhaber.

\section{\emph{con razón}}

%CON RAZÓN
This is an excellent sentence starter that can be used, like
a ver, either by itself or to introduce a clause. In both cases, its best
translation is "No wonder!" or "Little wonder." "As a child, John was
attacked by a Great Dane." Can raz6n les tiene miedo ahara ("No
wonder he's afraid of them now"). "My car broke down." Can raz6n
llegaste caminando ("No wonder you arrived on foot"). "1 ate three
chili dogs and now I'm feeling a little sick" Pues can raz6n ("Little
wonder" or "Well, whaddya expect?"). This notion can also be rendered
with No es para menos.

\section{\emph{por eso}}

%POR ESO
This phrase means "that's why" and frequently comes in
handy, by itself or with a clause, to draw a connection that may not be
immediately obvious. "1 hear you're going to marry my daughter." Sf
senor, par eso quiero hablar can usted ("Yes sir, that's why I want to
talk with you"). "There's a lot of crime in this neighborhood." Par eso
tenemos triple candado en la puerta ("That's why we have a triple
lock on the door"). "You want to go outside? But it's pouring rain!" Par
eso. Quiero estrenar mi nuevo paraguas ("For that very reason. I want
to try out my new umbrella").

\section{\emph{en fin}}

%EN FIN
A close fit for "So anyhow, ..." and very useful for leading off
a summation or a conclusion or for steering a conversation back to its
original point. En fin, no puedo visitar ahara = "So anyhow, I can't
visit now." The phrase is usually used as a sentence starter after you've
been talking for a while. For instance, after recounting all the different
plates you sampled at the restaurant, you might finish your description
with En fin, comi bien ("So anyhow, I ate well"). If you use it after
someone else has been talking awhile, you're saying, in effect, "Yes,
very interesting, but back to the subject at hand ...."

\section{\emph{total}}

%TOTAL
Total can also be used for "So anyhow, ..." and can be paired
with que to lead off a clause. Total que nadie fue a su fiesta = "So (as
it turned out) no one went to his party." Or, slangily, Total que fuimos
112 BREAKING OUT OF BEGINNER'S SPANISH
y no habia nadie = liSa anyhow we went and there was no one there. 1I
Both total and en fin sound more natural than the standard translations that dictionaries give for lIanyhowll and lIanywayll: de todos modos, de todas maneras, de cua1quier modo, en todo caso-at least at
the beginning of sentences. These translations are best saved for cases
where lIanywayll or lIanyhow" is used at the end of the sentence:
IIHe stole your car. Pero 10 quiero de todos modos ("But I love him
anyway").

\section{\emph{ya} and \emph{todavía}}

%YA AND TODAVÍA
Beginning students often have trouble distinguishing these
two invaluable sentence starters. As a rule, ya means lIalready" and todavia means IIstill" or "yet." Ya in the negative can equal IIno longerll
or "not ... anymore."

\bsk

Ya estoy comiendo. = "I'm already eating. II
Todavia estoy comiendo. = III'm still eating.II
Ya quiero comer. = III want to eat already (now)."
Todavia quiero comer. = "I still want to eat."
Ya no quiero comer. = III don't want to eat anymore,"
"I no longer want to eat."
Todavia no quiero comer. = "I still don't want to eat,"
"I don't want to eat yet."

\bsk

Let's say you're having a dinner party and the guest of honor is half an
hour late. You go to call his house and return with one of the following
messages:
Ya viene. = "He's on his way," "He's about to leave."
Todavia viene. = "He's still (planning on) coming (though not
immediately)."
Ya no viene. = "He's not coming anymore."
Todavia no viene. = IIHe still hasn't left."

\bsk

To learn ya well, keep in mind that it is often used for emphasis with
present-tense verbs. Regionally in American English "already" is used
in the same way: "all right already," for instance. Ya voy = III'm coming already." Sometimes "now" is the better translation. Ya nos vamos
= "We're going now." Ya estd 1a comida = "The food's ready now."
Sometimes the added emphasis is in the tone. Ya veras = "You'll see."
Ya intensifies imperatives as well. Gamete 1a fruta would be "Eat your
fruit"j Ya camete 1a fruta would be more like "Come on, eat your
fruit." Note that in most cases, ya indicates a change in the way things
have been up to that point.

Ya que is also a common sentence starter that students seem
to shy away from because "already that" doesn't ring any bells in their
head. It means "since" or, even better, "seeing that" or "now that." Ya
que terminaste este capitulo, ique vas a hacer~ = "Now that you've
finished this chapter, what are you going to do?"

\chapter{Snappy Answers}

Many native Spanish speakers will instinctively judge a foreigner's fluency in Spanish by the foreigner's ability to respond quickly
and conventionally to a wide range of stock situations and comments.
From such simple remarks as Mucho gusto and iC6mo estds! to more
complex utterances, many statements seem to have a traditional, almost automatic response built in. The sooner you master some of
these answers, the sooner you will achieve an enviable degree of fluency in Spanish.

Mastering the comeback means having a handful of stock expressions on the tip of your tongue, ready to be activated by the right
comment. Mucho gusto, someone says. A click goes off in your brain,
synapses slap together, and your tongue utters Igualmente. The whole
episode should take less than a nanosecond. You needn't think at all,
and in fact the less you thinK, the better. That may not be an ideal rule
for all verbal communication, but for learning comebacks ifis bliss.

\section{Affirmative}

For a perfectly adequate affirmative response you can always
say Sf, of course, But for true fluency you should shop around for alternatives. How often, after all, do you answer with a simple yes in
English? If someone asks if you want to go swimming, you are more
likely to respond "Sure" or "All right" than "Yes." If someone says "I'll
call you tomorrow," you may utter an "Okay" or "Good." "Can you
give me a hand?" elicits a "Sure" or "No problem," "Nice weather
we're having." "Really." "See you for dinner?" "Great." Likewise,
stock phrases exist in Spanish for all manner of situations. Get a
handle on a couple of them and you'll soon be saying yes in style.
\subsection{\emph{claro (que sí)}}
CLARO (QUE SI)
Generally defined as "of course," we would more colloquially
say "sure" in English. "Can you be here by four?" Claro. "Can you
lend a hand with the dishes?" Claro. It is also commonly used as a
sympathetic interjection for "obviously" or "naturally," and in this capacity it is often heard in one-sided phone conversations, with the listener repeating claro every so often to show he or she is still listening.
Somewhat gaudier alternatives for claro include desde luego and por
supuesto. iComo no~ is very natural-sounding, but you will have to
overcome your natural resistance to using a "no" phrase to say "yes."
Seguro works fairly well as "sure," but is also used ironically, as "sure"
is in English. To avoid doubt, use claro instead.
\subsection{\emph{está bien}}
ESTA BIEN
This is the phrase of choice for avoiding "okay," which is used
in Spanish and probably every other living language but is not native.
Furthermore, some uses of "okay" are simply wrong if translated into
Spanish. You would never say Esta O.K. for "It's okay," for instance. So
get used to saying Esta bien instead. "All right" can also be rendered
Esta bien. In fact, esta bien should be one of the commonest expressions to come out of your mouth when you speak Spanish. "I'm going
home now." Esta bien. "I'll call you later." Estd bien. "I'll pick you up
at eight." Esta bien. "Let's sell everything and elope to Paraguay." Estd
bien. To sound slangier, you can say 'ta bien.
\subsection{\emph{conste}}
CONSTE
This handy response comes from the verb constar, which is
one of those untranslatable words that frustrate dictionary writers.
Basically, the idea behind conste (from que conste, or "let the record
show") is roughly "for the record"; as a comeback, it means "That's a
confirmed fact" or "I'll hold you to that." Usually it is used in somewhat jocular fashion, as when a friend promises to visit you soon at
your new home in Arizona. Conste, you would respond, wagging a
friendly finger at your friend: "You've promised." Often it's followed
by ieh~ for emphasis. "I'll be home by eleven, Mom." Conste, ieh~
("Remember later that you said so" or "I'll be waiting up for you").
\subsection{\emph{de acuerdo}}
DE ACUERDO
Another good choice 'for the "all right" or "okay" group. It
is usually reserved for cases where some definite accord has been
116 BREAKING OUT OF BEGINNER'S SPANISH
reached, even if it's only an agreement to "see ya later." More formally,
you can use trato hecho ("it's a deal") or just hecho ("done"! in the
same cases. "1'11 pick you up at eight." De acuerdo. "Pick me up at
eight." Hecho.
\subsection{\emph{¿verdad?}}
lVERDAD?
A very simple remark but very common as well. It conveys the
idea of "Ain't that the truth!" but works well for "Really," especially
in response to questions that beg for an affirmative response. "This
weather's lousy." iVerdadt "Sure is nice to be here." iVerdadt "That's
a great movie." iVerdadt It is often preceded by sf: Sf, tyerdadt

\section{Negative}
Handy for requests made of you by friends, simple acquaintances, or even complete strangers, the negative phrase is everywhere
useful. The intensity of a rejection, even without straying over to the
side of obscenity, can range from the polite but firm to the impolite
but firm. As with sf and the affirmative, you could just say no for your
negatives. But read on and see what fun you'd be missing!
\subsection{\emph{para nada}}
PARA NADA
The correct translation for "no way," as in "No way, Jose."
"Wanna hear me play 'Feelings' on my kazoo?" Para nada.
\subsection{\emph{en absoluto}}
EN ABSOLUTO
Watch out-the phrase means "absolutely not," not "absolutely." In fact, it means the same thing as para nada, but is
more polite.
\subsection{\emph{¡qué esperanzas!}}
jQUE ESPERANZASI
Another fairly polite way of saying "Not on your life!" Literally, it means "What hope!" and suggests "You must be dreaming."
\subsection{\emph{en tus sueños}}
EN TUS SUENOS
A more impolite way of saying "You must be dreaming" or
"Dream on!" "How about a kiss, sweetie pie?" iBn tus sueiios! Add the
appropriate qualifier at the end and you have a rude, power-packed rejection. iBn tus sueiios, imbecil! = "Like hell, Dogbreath." Variants
include ini en sueiios! and ini soiiarlo!
\subsection{\emph{estás loco}}
ESTAs LOCO
This is the shortest way of saying "You're nuts," "You gotta be
kidding," or "You must be out of your mind," but it's not always the
most natural-sounding. For that, use Estas como loco or the slangier
Estas como operado, which is like saying "You ought to have your
head examined." Remember to make the phrase agree in gender and
number with the object of your derision. Estan como locos = "You
guys must be crazy."
r SNAPPY ANSWERS 117
\subsection{\emph{ni loco}}
NI LOCO
A very firm, very natural way of saying "No way." Literally,
you're saying "Not even if I were crazy (would I do such-and-such)."
"How about a quick game of Russian Roulette?" iNi loco! Remember to make gender and number agree: ni loco, ni loca, ni locos, or
ni locas.
\subsection{\emph{ya basta}}
YA BASTA
"Enough already!" or, in slangier situations, "Cut it out!" If
someone is relating a series of dubious anecdotes and ignoring all of
your newly learned expressions of disbelief, you may be able to shut
them up with a iYa basta! Said with a laugh, it means "No, don't go
on," "That's too much," or "Get outta here!"
\subsection{\emph{eso sí que no}}
ESO sl QUE NO
One you should learn just to impress your friends. It meansare you ready?-"no." But why just say "no" when you can say iEso s1
que no! Actually, it's a little stronger than just plain no and comes
closer to "That's out of the question" or "Forget it."
\subsection{\emph{fruits and other nonsense}}
FRUITS AND OTHER NONSENSE
Just for fun, you can substitute lots of irrelevant words for an
emphatic iNa! Two common ones are iNaranjas! (sometimes iNaranjas de la China!) and iMangos! Also used are iNarices! and iCuernos!
The latter expression refers to (and is sometimes accompanied by) the
two-finger "hook-ern-horns" gesture, which in turn can be traced back
to an indelicate reference to cuckoldry. Use it with caution.

\section{Surprise}

Surprises, by definition, catch you off guard. But that's no excuse to be tongue-tied as well. Here are some retorts that will enable
you to convey your surprise snappily.
\subsection{\emph{¿en serio?}}
lEN SERIO?
This is probably the best translation for the ubiquitous English
question "Really?" and usually suggests that you genuinely believe
what you are hearing, though you might prefer not to. "Mary's had an
accident." iEn serio~ It needn't be traumatic news. "I think my watch
has stopped." iEn serio~
118 BREAKING OUT OF BEGINNER'S SPANISH 1
:i
\subsection{\emph{¡n'ombre!}}
jN'OMBRE!
This universal slang phrase is an abbreviated form of no hombre but is pronounced very much like nombre. It is like ien serio~ but
is slangier and indicates more surprise: "No way!" Said laconically, it's
a simple negative, equating perhaps with "Nah."
\subsection{\emph{¡imagínese!}}
jIMAG/NESE!
These exclamations mean "Imagine!" or "Imagine that!" but
suggest stronger surprise: "Can you imagine that?" It's pretty formal
and is a good phrase to throw around when formality is expected of
you, as when your elders are describing shocking things that kids do
these days: "And girls these days let the boys kiss them on the first
date!" i1maginese! abuela. The familiar form is i1maginate!
\subsection{\emph{¡qué barbaridad!}}
jQUE BARBARIDAD!
The phrase means "Goodness gracious!" or the like but is
much more commonplace. "Good Lord!" or some other religious imprecation is often heard in English. It is used to respond to surprising
and generally unpleasant news, such as a natural disaster or a problem
in the extended family. It is not to be used when the surprising news is
also good news. If you're told that your mother-in-law has just won a
beauty contest, saying iQue barbaridad! might just earn you a fat lip.
iQue locura! ("What madness!") is used in much the same way when
the speaker is emphasizing the absurdity of the news as well.
\subsection{\emph{es el colmo}}
ES EL COLMa
This can be a tough phrase to translate, but you'll recognize
its use right away. It can mean "That's too much," "If that doesn't beat
all," "That's going too far," and so on. It is used for "last straw" situations, as when you can't fathom why someone has finally done something utterly outrageous. "Did you hear about the Van Gogh freak who
cut off his ear?" iNa! iEs el colma! "That rude son of mine forgot to
call me on my birthday!" iEs el colma!
Other interjections for surprise and shock include one-word
responses that cover a fairly narrow range of meaning-from "Gosh"
to "Omigosh," roughly. Two safe ones are iincreible! and icaray! Both
of these can also be used with Que to good effect: iQue increible! and
iQue caray! The first generally is in response to good news, the second
to bad news. Other expressions are euphemistic but safe, including
iHiio de! and iHiio! as well as (in Mexico) iHiiole! They all mean "Son
SNAPPY ANSWERS 119
of a gun!1I Common euphemisms for surprise in Mexico are iChin! and
iChihuahua! The Ca- family of euphemisms is also a good source for
mild, universally understood expressions of shock: iCaracoles! iCaramba! iCarachos! iCanarios! iCarape! and so on.

\section{Disbelief}

If you are traveling through the Spanish-speaking world and
are not yet accustomed to its idiosyncrasies, disbelief may well be one
of your most common reactions. Oh, how you will long for some good
expressions when your reservations are not respected, when the bus
doesn't arrive after all, or when the town you want to visit apparently
no longer exists! A good rule for visiting this magical kingdom is to
suspend your disbelief, but when you don't succeed, you'll need to express it. Here's how.
\subsection{\emph{no puede ser}}
NO PUEDE SER
Literally, "It can't be,lI even though of course it often can be
and is. This is the ultimate expression of disbelief, the one that conveys "I am not experiencing this!" or "This is not really happening! II
Good for when your breakfast bill tallies up to several thousand
dollars.
\subsection{\emph{no me diga}}
NO ME DIGA
This covers the laconic "You don't saY,1I as in English, but is
also widely used for "Don't tell me that!1I or "It can't be!" Slightly less
desperate than No puede ser, it indicates a less significant break with
reality. It can be used, for instance, when you arrive late to a concert
and are told that this performance is sold out. In the familiar, it's No
me digas.
\subsection{\emph{¡a poco!}}
jA POCO!
Slangy and not universal. It usually suggests that you don't believe what you are hearing and can even be a curt cutoff of what you
consider nonsense, like the English "Bull!1I but without the feistiness.
Maybe better English equivalents would be "You gotta be kidding!1I or
"Gimme a break!"
\subsection{\emph{déjese de cuentos}}
DEJESE DE CUENTOS
This means "Cut the crap" but is somewhat more refined.
"Come off itll might be closer. It does mean that you think you are
being lied to-or at least being kidded with-so don't use it too loosely.
A similar phrase is No me venga con cuentos. For a strong but per-
120 BREAKING OUT OF BEGINNER'S SPANISH
fectly decent way of saying "Like hell!" tell the neighborhood liar jEso
cuentaselo a tu abuela!-which means "Go tell your grandma."
\subsection{\emph{¿cómo?}}
i,COMO?
A simple but underrated form of expressing mild shock and
disbelief. If the cabbie says it will cost you seventy-five dollars to get
across town, you can very politely reply iComo?-as in, "I'm going to
pretend I didn't hear that, so let's try a new answer." Robert de Niro's
character in Taxi Driver, had he been speaking Spanish, might have
simply said iComo? for "You talkin' to me?" To embellish this phrase,
you can add a form of the verb decir: iComo dijo? or iComO dice?
\subsection{\emph{seguro}}
SEGURO
This means "sure," so it only works for disbelief if you ladle
irony all over it. Since "sure" often gets this tone in English, that
shouldn't be too hard to manage. The effect is like the English "Yeah,
right." Other expressions used ironically to express disbelief are claro,
ien serio? and como no. "Yes, Mr, Sanchez, your check is in the mail."
Camano.
\section{Indifference}
A lively and specialized vocabulary for indifference exists in
Spanish, and it's well worth your while to learn a few of its forms. Regional differences are often pronounced, so do pay attention to what's
being used around you and how people react to your usages. Keep in
mind, too, that often the best way of expressing your total lack of interest is-again-with a phrase for surprise or shock and plenty of
irony. iEn serio?-spoken as if you couldn't care less-works quite
well for "I couldn't care less."
\subsection{\emph{da lo mismo, da igual, es lo mismo}, and \emph{¿qué más da?}}
DA LO MISMO, DA IGUAL, ES LO MISMo, AND
i,QUE MAs DA?
These four everyday phrases are used for polite indifference, as
when you don't have a strong preference. They all mean "What's the
difference?" or "It doesn't matter."
\subsection{\emph{me vale gorro, me vale sorbete}}
ME VALE GORRO, ME VALE SORBETE
A considerably slangier and ruder form of indifference, especially in Mexico, where you will often hear Me vale by itself. It
roughly means "I couldn't give a damn." All of these me vale expressions are strong and can be considered substitutes for a rather rude ex-
lNAPPY ANSWERS 121
Hession discussed in Chapter 10. Don't worry about their literal meanngs-as euphemisms, they are essentially meaningless.
\subsection{\emph{me importa un bledo/comino/pepino}}
ME IMPORTA UN SLEDO /COMINO /PEPINO
More universal than the me vale constructions and slightly
tamer, it equals roughly "1 couldn't care less." Literally, you are saying,
"It is as important to me as a goosefoot plant/cumin seed/cucumber,"
respectively. In fact, add the minor foodstuff of your choice and you
will probably be understood.
\subsection{\emph{no me importa en lo más  minimo}}
NO ME IMPORTA EN LO MÁS MINIMO
Same idea using a less slangy construction. This is useful for
when you're worried that eloquence may detract from the directness of
your emotion.
\subsection{\emph{¿y qué?}}
"Y QUE?
Meaning "So what?" this can sound quite rude, depending on
your tone. In more friendly circumstances, it works as a slightly impatient "So what's the point?" iY eso qw?A is a variant.
\subsection{\emph{¿y a mi qué?}}
"Y A Mi QUE?
This means "What's it to me?" Said with a snarl, it means
"What the hell's that gotta do with me?" Said with a shrug, it's just a
simple "1 don't care."
\subsection{\emph{¿y?}}
"Y?
All by itself, this word means "So?" Depending on the context,
it can be an interested "So (what happened next)?" or an indifferent
"So what?"
\section{Assorted comebacks}
Here are several more comebacks that don't fit into any category but are useful in certain situations.
\subsection{\emph{¿no qué no?}}
"NO QUE NO?
A useful and cocky expression that means basically "1 told you
so." Coming back with tickets for the midnight train after being told
that they were sold out, you might flash your prize and say iNa que
no? In certain circumstances, the phrase would be construed as "In
your face!"
122 I;3REAKING OUT OF BEGINNER'S SPANISH
\subsection{\emph{es un decir}}
ES UN DECIR
This is a very handy expression that students of the language
should try to learn. It means something like "It's just a way of speaking" but will help bail you out of any situation in which you think
you're being misunderstood. "His Majesty would like to know if you
just called the queen a cow." No, es un decir nada mas ("Well, yes, but
that wasn't what I meant" or "It's just a figure of speech").
\subsection{\emph{o sea}}
o SEA
A very fine "crutch word" ("you know" or "um"), but also a
useful response in its own right. Again, the student of Spanish should
learn it early, since it is a formulaic way of saying "In other words" or
"Sorry, I missed that." Someone explains to you a convoluted route
you must take to reach the museum, starting with a right at the stoplight. You listen, get mixed up, miss the rest of it, and then say, 0 sea,
a la derecha en este semaforo ... , This generally gets the person to
repeat the directions or clarify the point,
\subsection{\emph{no es para tanto}}
NO ES PARA TANTO
Translated, this would mean something like "Let's not get carried away here." Somehow, though, opportunities for it arise more often in speaking Spanish than the English equivalent would suggest.
"You look pretty hot. You want a Coke?" Si, "Want me to pour it over
your head?" No es para tanto, In English you might say "I'm not that
hot!" This is quite a useful phrase once you get the hang of it. "I liked
that film." "If you want, we can sit through another showing." No, no
es para tanto, Sometimes a good translation is "It's no big deal."
\subsection{\emph{ni hablar}}
NIHABLAR
This common expression defies-or rather, embraces-categories. Depending on the context, it can convey "Don't mention it,"
"No way," "Oh well," and more. You're most likely to run across it
when confronting an insurmountable obstacle. "The IRS has ruled
that you can't count your trip to Bermuda as a business expense."
Nihablar.
\subsection{\emph{ni modo}}
NI MODO
This expression is heard primarily in Mexico instead of ni
hablar when dealing with major obstacles or crisis situations. A single
translation is hard to imagine, but a resigned "Oh well, what the hell"
conveys some of the sentiment. So does "Tough luck!" "A generator
blew and there'll be no electricity until July." iNi modo! "Mommy, I
don't want to eat the broccoli." Pues ni modo. This is not a very sympathetic expression, especially if the problem is someone else's, so be
careful how you use it.

\chapter{Invective and Obscenity}
Familiarity breeds contempt, as a sage once noted, but without
a little familiarity you can't breed anything. So it is, in a way, with
learning to curse in a foreign language. On the one hand, offering pointers may give the impression that I am encouraging students to blaspheme, excoriate, maledict, and heap billingsgate on their fellow humans. On the other hand, it is a simple fact that many students of
foreign languages will try their hand at four-lettered fricatives whether
I offer my counsel or not. And the danger of a little familiarity, one
fears, may be even greater than the danger of no familiarity at all. What
is a teacher to do?
My imperfect solution is to set forth a short checklist of Spanish swear words, making clear that it serves equally well as a short
course on how to sound rude, vulgar, and boorish. Most of these words
and phrases, used at the wrong moment or in the wrong company, are
definitely "fighting words," and fights-verbal or physical-in a foreign language, foreign country, or even foreign neighborhood are fights
you are predestined to lose. So swear if you must, but swear intelligently. If in doubt about a word's strength or appropriateness, try it out
first in the third person and on a neutral audience. For several of these
words and phrases, in fact, your best bet is to forget that they can be
used in the second person at all!
Two other hazards besides potential physical injury come into
play when you indulge in the crasser forms of the King's Spanish. First,
you will defile the image that some may have of you as a fine, upstanding man or woman-and not one of those reckless, godless, drug-crazed
norteamericanos that every Latin American and Spaniard knows so
124 BREAKING OUT OF BEGINNER'S SPANISH
well from u.s. television programs and movies. Foreign women are especially vulnerable to being "downgraded" or "depreciated" in this
way in Latin cultures; it may not be fair, but it happens to be true.
Second, you risk sounding like a complete fool, as anyone who
has heard a foreigner curse in English can imagine. I have seen U.S.
schoolkids maliciously teach foreign classmates the wrong way to
"cuss" only to reap pleasure from hearing Ivan or Pierre saYJ "I am so
shit on myself I could piss dog farts!" So don't think you'll sound too
much better cursing in beginner's Spanish. That out of the way, read
on for some pointers in this most dangerous game. What follows is not
meant as a comprehensive treatment but as a bare-bones primer. That
said, I can assure you that if you follow the advice offered here, you
will at least never sound as bad as Ivan or Pierre.
\section{\emph{}}
CABRON
A feisty fighting word. First, the etymology: a cabron is a "hegoat" and, from there, "one with horns," "a cuckold," "one whose wife
sleeps around." Its common use on the street, though, is usually to
refer to an aggressive, unlikable, mean-and-ornery kind of personexactly the sort of person whom you wouldn't want to call a cabron
face to face. (A mean and ornery woman would be called a cabronaagain, from a safe distance.) Among friends you may hear it, accompanied by a laugh, when one friend learns of a practical joke or dirty trick
(cabronada) committed by another. If Juan is desperately waiting for
Maria to call and Jose sneakily turns off the ringer on the phone, Juan
might notice and say icabron! In some parts of the Spanish-speaking
world, cabron can also be heard among tough-talking friends in an almost casual fashion, not unlike the way "motherfucker" is sometimes
heard on the streets of the United States. Be very, very careful about
imitating this quaint local custom: almost everywhere, saying Oye cabron is like asking, "Hey, wanna knife me?" Cabron is also widely
used as an adjective for "difficult," generally with estar. Esta cabron
would work nicely for "It's a bitch" when referring to a homework
problem or a stubborn bolt. Like "It's a bitch," cabron with estar can
apply generally to any unpleasant extreme. "How about this heat?"
Esta cabron.
CAGAR
The rude way to say "to shit" everywhere in the Spanishspeaking world. In some places it is ruder than in others, but nowhere
I ~
INVECTIVE AND OBSCENITY 125
will you use it to explain to your grandmother where you're going. For
that, use simply Voyal bano or, to be witty, Voya donde nadie puede
ir por mi ("I'm going where no one can go in my place").
You may find that cagar is quite common in rude and figurative senses. Me caga is a crude way of saying "I hate" something. Me
caga el futbol = "I hate soccer." Le caga ir al cine = "He hates going
to the movies." Gagarse en means "to shit on" and, figuratively, "not
to give a shit for." Me cago en tu opinion = "I don't give a shit about
your opinion." This phrase reaches fullest expression in the classic
phrase of frustration and rage, jMe cago en Dios! Save that one for real
special occasions. Gagar is sometimes used transitively in the sense of
"to screw (something) up." All of these usages should be considered
grade-A vulgar, considerably stronger than "to shit" in English (see
also mierda).
CARAJO
This is a useful expletive-universal, usually not too offensive, and plenty strong. In intensity, it is about on par with "goddam"
in English. Thus No oigo (veo) un carajo = "I can't hear (see) a goddam thing." Garajo (literally, but locally, "prick") is often heard in certain fixed expressions, such as jMe importa un carajo! ("I don't give a
damn"). In this sense, it can also be used with valer: jMe vale carajos
(un carajo) tu opinion! = "I couldn't give a goddam about your
opinion!"
In Mexico, the me vale format is especially common for expressing intense indifference. By following it with meaningless euphemisms (sorbete, gorra, etc.), you approach "I don't give a damn." Add
madre to it-i.e., Me vale madre(s)-and you have a crude, common,
and colorful way in Mexico of saying you don't give a flying french-fry
for anything. It is definitely obscene, but on the other hand you'll find
it on T-shirts sold to tourists. By extension, in Mexico a valemadrista
is someone who just couldn't care less about things and calls to mind
a person halfway through a week-long binge. These can be fun expressions, but they are also quite offensive. Use them with care.
Another set carajo phrase is Vayase al carajo, which equates
well with "Go to hell." Also very common for "Go to hell" is Vdyase
al diablo, literally "Go to the devil." In all of these constructions, if
you know the person you are cursing, or if it's a young person, you
would use jvete! instead of jvdyase! Mandar al carajo is "to tell (someone) to go to hell" or "to tell (someone) to get lost." "Where's your kid
brother?" Lo mande al carajo.
Garajo or carajos can also be inserted into almost any ques-
126 BREAKING OUT OF BEGINNER'S SPANISH
tion to give it a touch of rage and impatience. iCuando me van a atender/ ("When are you going to wait on me?") is what you ask after waiting five minutes, iCuando carajo(s) me van a atender/ is what you ask
after waiting forty-five minutes. Likewise, iQue carajo(s) estas hacienda = "What the hell are you doing?" iQue carajo(s) esta pasando/
= "What the hell is going on?" To tone things down just a little bit,
you can substitute diablos for carajo in most questions.
Finally, iCarajo! or iCarajos! by itself makes for a robust expletive. Euphemistic equivalents are endless: caracoles, caramba,
caray, cascaras, and so on.
CHINGAR
As Mexican as the margarita, this word has been studied and
written about in considerable detail by some of Mexico's foremost
scholars, Octavio Paz and Carlos Fuentes, for instance, dedicate sizable
sections of The Labyrinth of Solitude and The Death of Artemio Cruz,
respectively, to the psychosociological nature of the word in Mexican
life. Compare an English text with a Spanish one, and you'll get an excellent overview of how these words are used in Mexico.
In general, the word equates remarkably well with forms of "to
fuck," and all of its uses should be considered just as potent and perilous, Chingue a su madre is the classic Mexican curse: fists or bullets
answer it, not other curses. An hijo de la chingada is about the nastiest thing you can call a person, and they're definitely fightin' words as
well. iVete a la chingada! is the same as iVete al carajo! (see above),
only more so. iChingada madre! is as strong but can be used as a generic expletive (that is, a word for a hammer-smashes-finger situation).
If you must use it, make absolutely sure that no one around you even
suspects that you're referring to them or their maternal ancestor. iChingados! (pronounced chingaos to rhyme with "house") is safer as an expletive, but just as strong. In fact, among the chin- alternatives, only
iChin! can be considered an acceptable euphemism for this wordj the
others are too rude to count as euphemisms.
A rare few chingar usages and derivatives can almost be rated
PG-13, including chingar ("to pester"). I've heard a twelve-year-old tell
her father, Ay papi, no me estes chingando. That phrase and No chingues mean "Layoff!" or even just "Cut it out!" when used among good
friends (see joder), A fun and useful word, meanwhile, is cbingaquedito, meaning someone who pesters (cbinga) in a sly, unobtrusive fashion (quedito). A chingadera is a rudish word for a "dirty trick," "nuisance," or "major annoyance." A common euphemism for cbingar in
Mexico is fregar (see joderl, which can be extended to substitute for
INVECTIVE AND OBSCENITY
the naughty word in many of its various forms: una fregadera = "a
nuisance."
COGER
127
Coger is, of course, a perfectly ordinary, acceptable, and decent
verb meaning "to take" or "to grab." Unfortunately, it also has the
meaning of "to fuck" in many areas, especially the Southern Cone
countries and Mexico. More precisely, coger is a very crude way of saying "to have sex/' but supports none of the figurative uses of "to fuck"
(See chingar and joder.) The more poetic (and polite) way to say this,
incidentally, is hacer el amor ("to make love"). You can, by the way,
continue to use coger for "to take" in the countries mentioned above
(Spaniards, for instance, do it all the time), but expect a lot of snickering and strange looks.
COJONES
Mostly heard in Spain, though universally understood, this
word means "balls" in the sense of "testicles" and has many of the
same figurative uses. IQUe cojones! is "What balls!" Used by itselfICojones!-it's just a generic expletive. In both usages this word ranks
as one of the crudest in Spanish. It's a good word to avoid, especially
given the existence of so many good euphemisms. Two common ones
for the first usage are iQUe agallas! (roughly "What guts!") and iQUe
pantalones! ("What pants!"). The second makes the association between "pants" and "manliness." In fact, if the woman is perceived as
the one in charge in a relationship, it is said that ella lleva los pantalanes ("she wears the pants").
In the physical realm, "balls" can be called-equally
crudely-huevos, pelotas, or bolas in addition to cojones. The first
and second are mostly American (i.e., of the Americas), and the third is
quite rare. Polite alternatives are scarce, but then so are most polite
references to this part of the anatomy. The proper word is testiculos,
but you can play cute and refer to the general zone as a man's hombria,
or "manhood."
Huevos, in Mexico especially, is so widely understood to have
a crude second meaning that people generally avoid using the word in
the plural. You too should learn to say huevo instead of huevos whenever possible and to use the neutral iRay huevo! instead of the more
risque iTiene huevos! to ask a waiter or merchant if eggs are available.
The word blanquillos ("little white ones") is also heard in Mexico, and
128 BREAKING OUT OF BEGINNER'S SPANISH
though it seems unlikely that it has arisen as a "safe" alternative to
huevos, you can use it that way if you wish. Also in Mexico you can
refer folklorically to the testicles as los aguacates ["the avocados"),
especially since aguacate comes from ahuacatl, the Nahuatl word
for "testicles." Note that "Los Aguacates" is also the name of a traditional Mexican song, so be careful how you ask the band to play-Le.,
tocar-it.
eONO
A common interjection of the "Jesus H. Christ!" variety, but
considerably cruder. It is sometimes used alone, but more often to introduce a comment. ;Coiio! ;Que iugada! = "Christ, what a play!"
This word is popular in Spain but only selectively so in the New
World, so by using it where it's not common you may sound a little
strange-like someone saying "Gor Blimey!" in a Texas roadhouse.
eULO
This is the crude but common word for "ass" in Spanish. In
some countries it is cruder than in others. In Spain, culo is used freely
and easily, extending even to the back end of a car, for instance. In
Mexico, on the other hand, it is downright rude. You will certainly
want to suss out the prevailing mood before unleashing a word like
this. In the meantime, you can safely use trasero (roughly, "rear end")
for most everyday anatomical references and 1a parte trasera for the
rear end of cars. Other relatively polite words for the same corporal region include las na1gas ("buttocks"), 1a popa ("rear"), and e1 pompi or
las pompis. The last three are somewhat infantile. If you want to have
fun, refer to this region as donde 1a espalda pierde su digno nombre
("where the back loses its worthy name").
A related and quite offensive word, used mostly in the Americas, is cu1ero, loosely ascribing homosexual tendencies to its target
and most commonly heard in loud, lilting chants from the rowdies at
soccer matches: ;Cuuu-1eeeeeeeroooooo! The target here is usually the
referee or a dirty player on the opposing team.
HIJO DE
Name-calling in Spanish often has less to do with insulting
your opponent-a true macho doesn't care what you say about him-
INVECTIVE AND OBSCENITY 129
than with insulting your opponent's mother. And the common way to
do that, syntactically speaking, is to refer to your adversary as the son
of something offensive. What with the expression "S.O.B." in English,
the concept is not a hard one to grasp.
That said, it should be noted that the hi;o de complex of insults includes some of the strongest epithets in Spanish. English's
"sonnuva" selection is tame in comparison. In the most obvious case,
a beginner might equate hijo de puta with "son of a bitch," reasoning
that the structures are similar. Perhaps so, but the sentiments are a
world apart. "Son of a bitch" hardly warrants a raised eyebrow these
days, but hi;o de puta is as strong as they come: "motherfucker" is
probably the closest to it in intensity and offensiveness. Es un hi;o de
puta is quite simply the worst thing you can say of someone in most of
the Spanish-speaking world. (See also puta.) Should you feel inclined
to use this phrase, note carefully the correct pronunciation: in most
countries, the d gets swallowed altogether and the phrase ends up sounding like iHi;o'e puta! Incidentally, a fun word to describe the sort of
actions you would expect from a hi;o de puta is hi;oeputez, a word so
amusing-sounding that it almost avoids being obscene.
In Mexico, the form hi;o de la chingada is sometimes preferred, but hi;o de puta is universally used there as well. iHi;ole! is
also common as an innocent and euphemistic interjection there. In
various parts of Latin America, hi;o is sometimes turned into ;i;opronounced "hee-ho"-for no particular reason. Either ;i;o or hi;o can
be used alone or with nothing but de to convey the anger of the full
expression without having to employ vulgarities. Also popular as euphemisms are the harmless hi;o de su madre and the regional, and fun,
hi;o de su diez de mayo. Keep in mind that phrases like iJi;o de! can
still be interpreted as an offense, even though it is quite general.
JODER
Outside of Mexico, this is the common vulgarism for "to
fuck." Even in Mexico, most expressions using ;oder are readily understood, and some are used. Perhaps the commonest use of ;oder is by
itself as an expletive: iJoder! Also heard is No me ;odas, meaning
roughly "Don't fuck with me." iNa jadasl-like iNa chingues! (see
chingar)-is less forceful, about like "Cut the shit!" or even "Bullshit!" This can be quite strong, but said lazily, it carries no more force
than "Stop bugging me." For this concept you'll also hear the presentparticiple form: No me estes jadiendo. Again, strength depends largely
on the tone. A euphemism for this usage in much of the Americas is
fregar, as in No me friegues and No me estes fregando.
130 BREAKING OUT OF BEGINNER'S SPANISH
foder in the past-participle form conveys a slightly different
idea-closer to "fucked over" than "fucked with," though the difference is slight. Los jodidos of any country are the poor, downtrodden,
miserable masses. A road in very bad shape might also be called jodido. Likewise, you and your friends would say Estamos jodidos if you
ran out of gas fifty miles from town on a dirt road at night. Variations
on this phrase will depend on the assessment of blame. If your group's
poor planning skills were at fault, you might say Nos jodimos ("We
fucked up"). If someone stole your gas to put you in that predicament,
you could say Nos jodieron ("They fucked us over good!"). In Mexico,
these forms of joder are used as a somewhat weaker substitute for
chingar, and in general in Latin America fregar can again be substituted as a harmless euphemism: iEstamos fregados! ("We're screwed").
Lio
This word is neither rude nor obscene and, in a sense, doesn't
belong here at all. But it is a useful word to group together a number of
rude or rough-sounding expressions whose basic idea is "It's a mess."
That would be Es un lio in polite Spanish; it could be many other
things in less polite Spanish.
A lot of the "messy" words and phrases are regional. In Spain
un jaleo may heard; in Mexico, un relajo. Neither of these is particularly strong, Bronca is safe and universal for "mess," "altercation,"
or "violent confrontation." Inversely, No hay bronca works well as a
slangy "No problem" or "No sweat," and in slangy speech La bronca
es que is a common sentence starter meaning "The problem is ...."
Ruder words that you are likely to hear in Mexico include desmadre (a "bloody mess" or worse), for which euphemists can substitute despapaye. In Mexico and some other countries, a very crude
word is pedo ("fart"), meaning "big trouble" or "deep shit." (In other
countries, peo is used instead of pedo.) The equally vulgar No hay
pedo (roughly, "No problem, everything's cool") can be used-and
often is-to back out of an ugly scene. In Mexico pedo also means
"blind drunk" or "shit-faced."
MADRE
The mother is the paragon of the Latin family and is supposedly held sacred by the Latin male (especially). To show their deep respect for motherhood, Latin males (especially) have taken to inventing an entire vocabulary based on insulting other people's mothers. In
INVECTIVE AND OBSCENITY 131
Mexican street slang, for example, madre shows up in just about every
phrase, and almost invariably and paradoxically refers to the worst of
everything. Some examples of what you might hear: hasta Ia madre =
"wasted," "fed up," "drugged or drunk"; dar en Ia madre = "to beat
the shit out of"; romper Ia madre or partir Ia madre = "to beat the
shit out of"; en Ia madre = "in deep shit." The only exception to this
linguistic paradox seems to be a toda madre, which means "great" (or,
more correctly, "real fuckin' great"). Less vulgar alternatives include
a todo mecate and a todas margaritas; the original expression is a
todo dar.
Beyond madre the Mexicans have invented a host of crude
words. A desmadre, for example, is a "total mess," and desmadrar
means "to make a total mess of," Madrazo, besides being a common
last name, is a crude word for a physical "blow" or even an "asskicking." The situation is such that the word madre is best avoided
even in polite conversation in Mexico-use mama instead. Finally, as
on certain U.S. streets where you may hear "yo' mama" as a sort of allpurpose comeback, in Spanish tu madre can be handy. It is plugged
into a sentence to rebuff a rude or inappropriate suggestion. For instance, jHagase a un Iado! ("Move over!") can be answered with jQue
se haga a un Iado tu madre! (literally, "Have your mother move over! ").
jTermina rapido! ("Hurry up and finish!") jQue termine rapido tu
madre! For those reluctant to delve into the subjunctive, jTu madre!
by itself generally gets the point across. So does a simple jLa tuya!-
which packs quite a useful, all-purpose punch and translates roughly
as "Up yours!" For a watered-down version, jTu abueIa! can be used
instead. Sassy kids (like Mafalda, for instance) might use this. "Eat
your soup, Mafalda," jQue Ia tome tu abueIa!
MALDICI6N
As a mild expletive, jMaldicion! is equal to "Damn!" For
"Damn you!" use jMaldito seas! (or jMaldita seas! for a woman). This
works also for objects: jMaldito sea tu cache! = "Damn your car!" For
"damn" as an adjective, maldito is one of many words that fill the bill:
tu maldito cache = "your damn car." Mugre is also in widespread use,
though it is much more polite-akin roughly to "lousy" or "no-good."
Condenado and recondenado are also heard, though less often. And in
a few choice expressions, even puta is used as a modifier, chiefly in the
phrases jQue puta vida! and jQue puta suerte (Ia mia)! Its euphemistic substitute in these expressions is perra. jQue perra vida! = "Life's
a bitch."
In Mexico the word pinche serves this intermediary function
132 BREAKING OUT OF BEGINNER'S SPANISH
as well. Although it's a good bit more vulgar than maldito, it is still
exceedingly common. When the Mexican bandido cop in Treasure of
the Sierra Madre says to Bogart, IIBadges? We don't need no stinkin'
badges!" what he would have said in Spanish for IIstinkin'" is pinches.
Literally, pinche is a noun referring to a lowly kitchen employeej the
noun is still used, though now it can describe any low-level employee
or "gopher." Pinche gringo, incidentally, is one of the typical remarks
that sullen, gringo-hating Mexicans like to mutter under their breath
as you walk by. Ignore it, or, if you're feeling especially feisty, say
Gringo sf, pinche tu abuela. Better yet, buy the guy a beer.
MARIC6N
In a macho culture like the Latin one, there is no getting away
from the epithets that attack an opponent's manhood. Most common,
most unequivocal, and strongest is maricon, which is roughly equal to
"faggot." Used by a woman against a man, it is a truly debilitating insultj that is, it can be a good way in certain public circumstances to get
a man to leave you alone. Used by men, it is either a challenge equal to
"coward" (or "pussy"!, in which case it is a fighting wordj or it is dismissive, as when the target of the insult really is effeminate-and thus
not a worthy opponent for a "true" macho. Equally rude and offensive-though not as universal-are puto and jato. Somewhat less rude
choices include marica, mariquita, mariposon, and barbilindo. They
range in intensity roughly from "queer" to "queen." Poco hombre is a
common expression for any "unmanly" man. Both it and marica mean
little more than "wimp"-that is, without any allusion to sexual preference. In some parts of the macho's world, such epithets are appropriate for "unmanly" men who help with the dishes, allow women to go
out by themselves, or show emotion when a woman leaves them.
MEAR
A vulgar word for urination and the equivalent of "to piss."
The cutesy, children's way of describing urination is hacer pis or hacer
pipf, which are essentially the same as "to go wee-wee." Hacer aguas
is also heard. The only widespread and correct way of saying "to urinate" is orinarj note that it's spelled and pronounced with an a, not a
u, at the start.
MIERDA
The universal vulgar word for "excrement" in Spanish is
mierda, and you'll find that it is usually an accurate substitute for
INVECTIVE AND OBSCENITY 133
"shit." Some people consider mierda very offensive, while others use it
quite freely. The kiddies' word is caca, as in caca de perro or caca de
vaca. Excrementa is the safe way out.
Overall, mierda is probably less frequently used in rude expressions than is "shit." Neither caca de taro nor mierda de taro
would be understood as "bullshit" in its figurative sense, for instance.
Nor is mierda used commonly as an expletive. About the only phrase
in which it does take on a figurative meaning is de mierda, tacked
on after a noun, to mean "worthless" or "piece of shit." iCoche de
mierda! = "Piece-of-shit car!" Likewise, una mierda by itself can
mean "a piece of shit." iEste reloj es una mierda! = "This watch is a
piece of shit!" The epithet comemierda (literally, "shit-eater") is regionally favored (in Cuba, for instance) to describe just about anything
or anyone disagreeable.
PENDEJO
This strong, wide-ranging insult literally means "pubic hair."
Figuratively, it is almost universally used to mean "asshole" or "shithead," generally highlighting the target's stupidity more than his or
her maliciousness. Thus a common phrase is verle la cara de pendeja
(a alguien), which means roughly "to note someone's stupidity in their
face" or "to size them up as an easy mark." If a sidewalk merchant
tells you a Coke costs four hundred dollars in local currency, you could
rightfully ask iDe que me viola carat ("What do you take me for?"). If
you pay the money and don't realize it until later, how you feel about
yourself is a pretty good summation of pendeja. By extension, a pendejada ("screwup") is any stupid behavior of the sort expected from a
pendeja. Extended further, this helpful word can be stretched into pendejez, describing someone's general stupidity. "You paid four hundred
dollars for a Coke?" Fue un ataque de pendejez. Polite substitutes for
pendejada are disparate and payasada.
PUTA
Puta is a close fit for the English "whore," and possibly the
most universal profane oath and expletive of all is iPuta madre! or
simply iPuta! Shortening it further, to iPut! or (uta! or even to just
(ut! has the effect of cleaning it up a tiny bit for polite company. And
as it can be trimmed down and thus sugarcoated, it can also be embellished for effect. Quality cursers can drag the curse on for umpteen syllables: iLa puta que te pario! ("The whore that bore you!") is an example; La reputisima madre que te pario a stronger, even IOllger one.
134 BREAKING OUT OF BEGINNER'S SPANISH
These phrases are fine as generic expletives, but as with other
curses, be careful to make it clear that you are not referring to someone else present. If in doubt in the preceding examples, change the te
to me, converting your oath against an unjust world to an oath against
your own stupidity or clumsiness-and your own mother, incidentally.
Note also that there is a thin line between a generic expletive and a
foul insult, so be forewarned. IPuta madre! is an unaimed expletive,
but iTu puta madre! is a ballistic insult. You will want to be sure that
both you and your audience are aware of the difference.
\section{\emph{}}
\section{\emph{}}
\section{\emph{}}

\chapter{Which Is Which?}
For the foreigner, the path to fluency in Spanish is marked by
constant dilemmas. Ser or estar? Imperfect or preterit? Subjunctive
or indicative? Do you want the word on the left or the word on the
right? Do you want to keep the word you have, or do you want to trade
it for the word behind Curtain Number One? Or do you want to swap
it all for the word Linda is bringing down the aisle in a lovely giftwrapped box?
For foreigners learning English, one of the greatest bugaboos
is the tendency for one English word to have three or four (or a dozen)
different meanings. The problem is worse with verbs, which team up
with prepositions in a million-odd ways to produce new meanings altogether. As an exasperated Mexican friend once noted, "What can you
expect from a language where first you chop a tree down and then you
chop the same tree up?"
The converse of these difficulties are your difficulties, as a native English speaker, in learning Spanish. Every single English word, it
seems, requires a bundle of totally different foreign words to describe
it. And worse, you're expected to learn them all!
In this chapter, we look at some of the English words that
don't have a single Spanish equivalent. Instead, you must pause in your
palaver and search for the correct Spanish word. And while no book
can present and resolve all possible word-choice dilemmas, this chapter can help you sort through the choices so you will know which ones
you need to learn.
\section{\emph{}}
BACK
The body part, as you probably know already, is la espalda..
But what about all those other uses of "back?" For these, you'll need
to study tras, atras, and detras, plus de regreso and de vuelta and, last
but far from least, fonda (or al fonda).
Generally, to express "in back of" or "behind," atras and detras are the words you'll use most. Tras, for all intents and purposes, is
just like detras and can be safely discarded for the time being. As for
the difference between atras and detras, native speakers will tell you
they distinguish between these two words "by which one sounds right."
That is true, but of little help to you, the student. So we must turn to
rules, though they are far from perfect, to clarify the situation.
Here's the main rule to remember: use detras de instead of
atras de, and you'll usually be right. In fact, it is hard to think of a case
where atras de would be correct and detras de wrong. So you are free
to forget about atras de altogether. Or almost. One exception is in reference to someone coming behind or after someone else. Here, detras
de suggests "in pursuit of"; atras de just means "back there somewhere" or even "later." iD6nde esta tu mama~ Viene atras de mi.
"Where's your mom?" "She's coming behind me (later)." Change atras
de mi to detras de mi and you have "She's coming after me"-in
hot pursuit.
The underlying intuitive difference between atras and detras
is hard to pin down. In most cases, detras suggests a spatial relationship, or one thing "behind" another or being blocked by another. Atras
is much vaguer, meaning "in back/' "back there somewhere" or "behind (the speaker)." You might say, for instance, that there is a mountain detras de mi casa. But if someone inside the house asks where the
mountain is, you might say alla atras.
In fact, if we accept detras de as (almostl always the correct
choice, then we could invent a second, equally imperfect rule to cover
atras: when you don't need the de, use atras. That is, when standing
alone as an adverb, atras is correct and sufficient. This usually covers
backward motion, as in va bacia atras or va para atras. In keeping
with its vague "backness/, atras is usually the word of choice in figurative expressions. Se esta ecbando para atras = "He's backing out."
No mires atras = "Don't look back."
For other uses of "back/' de regreso and de vuelta both mean
"back" in the sense of "returned." Both are used with estar. Estd de
vuelta despues de su viaje = "He's back now after his trip." Of course,
regresar and volver convey the same thing.
Fonda, or al fonda, is one of the handiest words for "back"
WHICH IS WHICH? 137
and one you should try to learn. EI fonda, as a noun, refers to the
"back" of a bus, movie theater, or building (as well as the "bottom"
of a glass, swimming pool, and so forth). You'll hear al fonda for "to
the back" or "all the way in the back." Ask the bus driver or theater
usher if there are any seats and the answer may be Sf, al fonda. Atras,
especially with cars, can be used this way as well. It means something
like "in the back" as opposed to "at the very back," which is al fonda.
Al fonda is extremely common when giving directions, particularly
indoors. Ask where the bathroom is in a restaurant and the reply may
be, for instance, Al fonda ala derecha ("All the way back on your
right").
BECOME
This is perhaps the gold-medal winner for troublesomeness
in the English-to-Spanish olympics, especially when "to become" includes the sense of "to get." As a rule, ponerse is the handiest word for
"to become," but you should learn at least some of the others. Ponerse
in general is for fleeting states of mind or conditions, as in ponerse furioso ("to get/become furious"), ponerse viejo ("to get/become old"),
ponerse nervioso ("to get/become nervous"), and so on.
For longer-term, usually nonreversible conditions, hacerse is
often the verb you want. Me hice rico = "1 became (got) rich"; EI penique se esta haciendo inutil =' "The penny is becoming useless"j EI
nuevo baile se esta hacienda, popular ("The new dance is becoming
popular"). The distinction is' subt,le, sometimes to the point of near in-
.visibility. Thus one can also hacerse viejo, but it suggests "becoming
an old person" or "becoming elderly," whereas ponerse viejo hints at
"starting to feel (or look) old."
Volverse, another word for "to become," is perhaps best translated as "to turn into." Se ha vuelto una verdadera molestia = "He's
become (turned into) a real nuisance." It involves a more sudden and
unexpected change than hacerse, though it shares with that word a
change into a more or less lasting state. Se volvi6 rico, for instance,
can also be used to say "He got rich," but it gives the impression of
overnight wealth-winning the lottery, perhaps, or inheriting a fortune-and also carries an intimation of criticism, as if the change were
undeserved as well as unexpected. Volverse is also the only phrase to
be used for "to go crazy" in the permanent sense: volverse loco.
Both transformarse and convertirse can also be rendered
"turned into," but usually in connection with a more thoroughgoing,
physical change. EI agua se transforma en vapor = "Water becomes
(turns into) steam." Clark Kent se convierte en Superman = "Clark
Kent becomes (turns into) Superman." Sometimes these phrases can
138 BREAKING OUT OF BEGINNER'S SPANISH
cover more figurative conversions: Se ha convertido en un buen escritor = "He's become a good writer." Llegar a ser, meanwhile, works
well as "to turn out to be," as in La formula llego a ser muyutil =
"The formula turned out to be very useful." This phrase also covers
"to become" in the sense of "worked his (or her) way up to." Despues
de muchos esfuerzos, llego a ser medico. = "After a lot of hard work,
he became a doctor." In many cases, an easier construction can be
made with the verb resultar, which also means "to turn out." Este libro resulto ser muy interesante. = "This book turned out to be very
interesting."
A very common but often overlooked way of saying "to become" in Spanish is through the use of reflexive verbs, especially those
that are little more than an en- or a- prefix attached to an adjective or
noun: acalorarse ("to get/become hot"), enfriarse ("to get/become
cold"), entristecerse ("to get/become sad"), acercarse ("to get/become
close"J, and so on. Sometimes no prefix is needed with the adjective,
as in mo;arse ("to get/become wet"J and cansarse ("to get/become
tired"J. Pay attention to which nouns and adjectives can be made into
verbs and how, and you'll be well on your way to conquering the "get/
become" problem.
BURN
In English, this is a fairly simple if destructive processj in
Spanish it's a little more complex. Arder refers to the actual act of
burningj think of it as "on fire" or "burning up." Thus houses and
candles arden, and so do fever victims: Estoy ardiendo en calentura.
Out of context, Estoy ardiendo can sound like double entendre-"I'm
on fire, baby, sweep me off my feet." Figuratively, arder comes into
play in the expression esta que arde, meaning someone is "burning
mad" or a situation is "boiling over."
Incendiar, another choice for "to burn," is usually used for
things that should not be burning: forests, buildings, the lawn, the
lawn furniture, and so on. Think of it as "to set on fire." Even in the
case of pasion incendiaria, the implication is that the fire has gotten
out of hand and may end up being destructive.
Quemar also translates as "to burn," but generally brings to
mind (in the reflexive quemarse) "to burn up" or "to burn down," (another case of contradictory prepositions in English saying almost exactly the same thing). If you go away for the weekend and come back
to find your toolshed in ashes, a neighbor might explain simply, Se
quemo. Quema in its nomeflexive form is the usual way to convey the
transitive "to burn (something)." Estas quemando el bistec = "You're
WHICH IS WHICH? 139
burning the steak." (If you were to use incendiar here, you would be
saying, "You're setting fire to the steak.") The general idea behind quemar is "to consume (or damage) by fire." It is the correct verb for when
one thing "burns" another: the sun quema your skin, the stove quema
your fingers, and so on. The painful and unappealing marks left on
your skin to mark these mishaps are called quemaduras ("burns"). To
say "I burned myself," whether in the sun or the kitchen, you would
say Me queme.
CORNER
Ninety-five percent of the time, just use esquina for "corner."
For that matter, use it for "intersection" too, although intersecci6n
is increasingly heard these days. For a "crossroads," or intersection
of highways, use either entronque, encrucijada, or cruce de caminos.
Dictionaries will give rinc6n for "corner" as well, but it generally refers to an inside corner where two walls meet-the corner of a room,
in other words. By extension, it is used to mean a "nook" or "cranny,"
and is a favorite word to assign to cocktail lounges, piano bars, and
the like.
DATE
Here you'll have to slow down and analyze what exactly you're
trying to say. Do you mean "the date," as in July 13, 1984? Then the
Spanish word you want is fecha. If you mean "date" as in "appointment," you have a couple of words to choose from. A "date" of the sort
that thrills teens (and, let's be frank, some postteens as well) is almost
universally a cita. A cita is also what you have with a doctor or lawyer, however-in other words, an "appointment." That is, in Spanish
you can have a cita with your doctor without upsetting your spouse. A
compromiso is also an "appointment," usually an unspecified one. It
can almost mean simply "something to do" and implies an obligation
to be present at a certain time-but no more. Thus, to escape from a
dreary luncheon, you can excuse yourself with Perd6n, tengo un compromiso. All you're really saying is "Sorry, gotta go."
FAIL
The best all-purpose word for "to fail" is fracasar, though one;
1952 edition of a heavyweight Spanish-English dictionary I consulted
140 BREAKING OUT OF BEGINNER'S SPANISH
defines it as "to crumble, to break into pieces: applied commonly to
ships." A ship that crumbles into bits would certainly meet my definition of "failure," but fracasar in common usage covers-much more
ground than that. Fallar is also used for many general senses of "to
fail," though it seems a little kinder and less permanent than fracasar.
Fa1l6 en su intento suggests "He did not succeed"; Fracas6 en su intento hints "He was a total flop." Often you'll hear fallar in the form
estar fallando to mean "not to be working well," "to be on the blink,"
in reference to some appliance or gizmo. When someone or something
"fails" you in the sense of not living up to your expectations, either
fallar, defraudar, or decepcionar can be used. To "fail" an exam or a
course, reprobar is much used in the Americas, while suspender is
more frequent in Spain. Thus un suspenso on your exam in Madrid is
beyond suspense: it's an F. In Mexico City receiving an F would be expressed as reprobar e1 examen or ser reprobado par e1 profesor.
FRONT
This word opens an unruly can of worms on its journey into
Spanish. The basic problem is the excess of believable cognates, which
then have to be sorted through. For instance, you can say there is a
fountain a1 frente de 1a casa, enfrente de 1a casa, frente ala casa, or
de frente a 1a casa, and you won't necessarily be saying the same
thing. What's the difference?
A1 frente de is usually the expression for the English "in front
of." It means "at (in) the front part of (something)." In the earlier example, a fountain that is a1 frente de 1a casa is on the property at the
front of the house. Next come a host of prepositional expressions that,
in many circumstances, mean "in front of" but generally only in the
sense of "facing." Enfrente de is one of these expressions, and it is better regarded as a trickster (see Chapter 3). Think of it as asynonym for
"facing" or "across the way." A fountain that is enfrente de 1a casa is
likely to be across the street from the house. Frente a means much the
same and should be viewed as just as unreliable as enfrente de in most
cases. Sometimes the act of "facing" is implicit. Lo tienes frente a los
ojos = "You have it in front of your eyes"-that is, "staring you in the
face." De frente a, finally, states very clearly that something is "face
to face," literally "with its front facing (something)." Estoy parada de
frente a1 sol = "I'm standing facing the sun."
Perhaps the safest word of all for "in front of" is de1ante, usually followed by de. Generally, whenever you can use "ahead" or
"ahead of," you should probably be using de1ante or de1ante de. But it
also covers a wide range of "in front of" situations. The fountain can
WHICH IS WHICH? 141
be delante de la casa, for instance. Bob se sienta delante de mi en la
c1ase = "Bob sits ahead of me in class."
The classroom example may prove, well, instructive. Ana's
seat is in the third row; Bob sits delante de ella. Pedro sits in the front
row, delante de Bob and al frente de la fila ("at the front of the row")
and frente ala maestra ("in front of-i.e., facing-the teacher); when
Ana gets up to read her book report, she goes al frente del salon and
stands facing the class, or de frente a la c1ase. Ana's sister is in the
classroom across the hall, or enfrente del salon de Ana.
Finally, whereas atrds is "in back" (see "Back"), adelante is
usually "up front"-in the front seat of a car, for instance. To refer to
"the front" or "the front part" of something, el frente can usually be
used. El frente de la tienda estd sucio = "The front of the store is
dirty." La frente, in the feminine, refers exclusively to "the forehead."
FUNNY
Does it make you laugh or wonder? That's the first question
you'll need to ask (and answer). If it's the former, you can choose from
chistoso, gracioso, and comico. If the latter, stick to extraiio and raro.
Pick your favorite and use it; they are virtually interchangeable. Anything having to do with "fun," incidentally, should be carefully distinguished from things that are "funny," since what's fun isn't always
funny. "Fun" needs either the adjective divertido or the noun diversion to describe it.
HAPPEN
"To happen" can almost always be covered by some form of
pasar in Spanish. Other counterparts exist-including acontecer, ocurrir, suceder, and acaecer-but there's really no need to learn them
other than to recognize them when you hear them. You will have to
learn some of their noun forms, however, since pasar doesn't have one.
Acontecimiento, hecho, and suceso are three of the most frequently
encountered words for "happening," "occurrence," and "event."
Evento will be understood-and is gradually being incorporated in this
sense-but technically it means something that happened by chance
or with no advance planning.
HERE
Spanish is funny. Even when you know you're "here," you
have to know which "here" to use. Two words, acd and aqui, cover
142 BREAKING OUT OF BEGINNER'S SPANISH
the situation. Which is which? Textbooks generally promote the idea
that aca is the correct word for verbs of motion, as in Ven aca ("Come
here"), while aqui is the word of choice for stationary presence in a
given place, as in Estoy aqui ("I'm here"). After hearing one person on
the street shout Estoy aca, though, you might wonder why grammarians even bother with these distinctions.
The comparative forms of aca come into play when giving
guidance to, say, someone backing a truck down an alley. Mas aca
means "A little more this way," No tan aca means "Not so much this
way," and j¥a! means "Just right, hold it." jPraaac! ("Crunch!") means
you won't be asked to help truck drivers back down alleys anymore.
Finally, a very common use of "here" in English is to say
"Here you are" or "Here you go" when giving someone something (for
example, paying the cab driver the fare). In Spanish you can say pretty
much the same thing with aqui: Aqui esta and Aqui tiene both work,
as does Tome ("Take"). A somewhat archaic expression that still creeps
into written Spanish is He aqui, and some dictionaries will tell you
that it means "Here is ...." Actually, it's closer to "Behold!" Before
trying it out on the cabbie, try to imagine yourself saying "Behold, the
ten pesos!" in English. Then go back to Aqui tiene.
HOT, WARM
Native English speakers aspiring to fluency in Spanish should
be aware of the latter language's hierarchy of "heat" adjectives. Apart
from the several words representing various degrees of heat, different
words are used depending on the object in question. A soup fresh off
the fire, for instance, will drop in temperature from hirviendo ("boiling
hot" or "scalding") to caliente ("hot") to calientita ("pretty hot") to
tibia ("lukewarm") to temp1ada (not hot but not cold either). A cold
beverage goes from he1ada to fria, fresca, a1 tiempo ("room temperature"), tibia, and caliente. For weather, you can describe the day in order of ascending temperature: he1ado, frio, fresco, temp1ado, calientito, caliente, and ca1uroso. A good word for warm days is calido, often
in the sense of "unseasonably (or pleasantly) warm." As expressive alternatives for ca1uroso you can say Hace un calor barbaro (or tremendo, bochomoso, sofocante) or simply Hace un ca1m6n.
"Hot" in its many figurative senses in English can be a dangerous concept to convey in Spanish. Caliente in reference to other
people, especially of the opposite sex, is risky. Think of caliente as
"in heat" rather than "hot-looking," and you'll get the general idea. As
for the common novice mistake of saying Estoy caliente instead of
WHICH IS WHICH? 143
Tengo calor ("I'm hot"), those of you who have used it know the consequences: usually belly laughs all around, sometimes an unwanted offer.
HURRY
Several choices present themselves for the concept of "to
hurry," and mostly it's a matter of choosing one and using it. There are
some differences, though. Your choices are (the envelope, please): tener
prisa, apurarse, and darse prisa. The first means "to be in a hurry,"
and Tengo prisa is a basic Neurotic-Speak expression. Apurarse is less
common except in the phrase iApurate! (or iApurese!), meaning "Hurry
up!" and then only in the Americasj apresurarse replaces apurarse east
of the Atlantic. The adjective apurado is often heard, but implies more
"harried" or "frantic" than just "in a hurry." Darse prisa is sort of like
"to make oneself hurry" or "to get a move on," and iDate prisa! is
practically interchangeable with iApurate! for urging someone to get
the lead out. When you're in such a hurry that you can't even stop to
say "Hurry up," appropriate grunts include iYa! iAprisa! iDeprisa! and
in Mexico iAndale! and iOrale!
LOOK
"To look" in English covers a lot of ground. "To look for" is
not the same as "to look at," for instance, and "to look good" is a different concept from "to look like." As usual, one English verb can
combine with a score of prepositions to produce dozens of distinct
meanings-each of which in Spanish may require a different verb. The
most common Spanish words for the concept "to look" are buscar ("to
look for") and mirar ("to look at").
"To look" by itself is verse, a verb you should learn to use. Te
ves bien = "You look good." Se ve horrible = "It (or he or she) looks
horrible." ieOmO me veot = "How do I look?" And so on.
"To look like"-with the meaning "to resemble" or "to remind one of"-is either parecer or parecerse a. Ese sefior se parece a
mi padre = "That man looks like my father." When "to look like"
means "to look as if," parecer by itself is enough. Parece que va a
llover = "It looks like (as if it is going to) rain."
MISS
Your husband goes away on a bus and you miss him. He
misses the bus, so you don't miss him. On his way to the bus, another
144 BREAKING OUT OF BEGINNER'S SPANISH
bus just misses him, causing him to miss the bus and you almost to
miss him very much. Still with me? There's obviously a lot of missing
going on in this world. And in Spanish, you must take care to differentiate what you do to your husband and what the bus does.
Let's deal first with the verb that expresses what you feel
when any loved one goes away. This sense of "to miss" in the Spanish
of the Americas is handled by extranar. In Spain, you'll hear echar de
menos. As in English, you can "miss" not only a person but your bed,
your favorite brand of candy bar, or any once-regular activity, like your
nightly mud bath. All are covered by extranar and echar de menos.
Now let's look at "to miss" in the context of missing a bus,
a plane, a party, a movie, a lecture, and so on. In Spanish, this idea is
handled by perder, which of course is also the word for "to lose." Perdimas el tren = "We missed the train." If you miss something that
you were expected to attend, you would use faltar a, as in Falte a la
c1ase de espanol ("I missed the Spanish class"). Think of it as "to
miss" in the sense of "to be absent from." When something is "missing" or "absent," faltar is also the word of choice. Faltan once d6lares
de mi cartera = "Eleven dollars is missing from my wallet."
Then there is "to miss" in the sense of the close call-"The
bus barely missed the parked truck"-which is a fairly tricky concept
to translate into Spanish. Some dictionaries might have you use errar,
no acertar, or no dar en el blanco, but these imply that the bus was
trying deliberately to hit the truck-and may even have gone back for
a second shot. These are all useful expressions, but they are best saved
for when your jump shot goes astray or when you guess something
wrong. To convey the idea of the bus "missing" the truck, you would
have to resort to pasar rozando ("to graze") or pasar cerca ("to go by
close"). To get the point across even more clearly, you would need to
trade in "miss" for "almost hit." Un autobus par poco (or casi) atropella ami marido = "A bus just missed (running over) my husband."
NEXT
At some point in your progress as a student, you will realize
that pr6ximo is not the only word for "next." Soon afterward, you will
realize that it's often not even a very good word for "next." What are
the other words, and when are they used? For starters, you're safe with
pr6ximo when "next" means some indefinite "next time." In fact
pr6xima vez ("next time") is one of its commonest uses. But pr6ximo
does not work for "the next day," which is el dia siguiente (or al dia
siguiente). It works again for "next week" (la pr6xima semana), and
fairly well for "next year" (el pr6ximo ana), but you are probably more
WHICH IS WHICH? 145
likely to hear la semana que entm (or entrante) and el ano que viene.
Somewhere someone has probably thought up an explanation for these
deviationsj ignore it and just learn the phrases.
Another slight deviation to be alert for is the use of otro for
"next," which you will run across constantly. Al otro dia can mean
"the next day," for instance. More common is la otm for "the next"
street, stoplight, or corner when being given directions. Esta en la otra
calle = "It's on the next street over." What tends to confuse beginners
is that the noun is often implied, not stated. De vuelta no en esta sino
en la otm means "Don't turn here (at this corner) but at the next one."
For "next to" in a physical sense, pr6ximo is flat-out wrongj
junto a or allado de is correct. La tienda esta allado de su casa or La
tienda esta junto a su casa = "The store is next to his house." Note
that as an adverb of place, junto does not change with the gender of
the subject. "The girl next door" is either la muchacha de junto or la
muchacha de allado. Colloquially and quite commonly you will hear
pegado (literally, "stuck to") for "right next door."
OFFICE
Such an easy concept-and such a wealth of Spanish words to
cover it. Let's start with-and discard-oficio, which applies only to
"office" in the symbolic, conceptual sense: "the office of vice president." For a room where someone works, several words could fit, depending on what you want to say. The most generic term for "office" is
. oficina, and this can be used safely about 90 percent of the time. Getting pickier, we can distinguish between the larger office under the
boss's control and the actual office where he or she works, which could
be called his or her privada ("private office"). Doctors and dentists
have special offices called consultas or consultorios. Lawyers dwell in
bufetes, which works for both the physical office and the "firm." A
despacho is a workplace, usually nongovernmental, that is often involved in accounting or administration, though lawyers can labor in
despachos as well. A taller is a "workshop" where manual labor is
done. Taller mecanico means "worskhop" or "garage" and has nothing
to do with the height of the mechanic who works there.
OLD, OLDER
The main problem with this concept is the word viejo, which
you should probably be taught from early on not to use in reference
to people.
146 BREAKING OUT OF BEGINNER'S SPANISH
The other problem is the concept of "older," which in English
can refer to a two-year-old in comparison with a one-year-old, In Spanish, viejo means "old" in the very specific sense (for humans) of having
been on the planet for more than fifty Of sixty years. Mas viejo, naturally, means "older," in the sense of having been among us for sixtyplus years. Therefore, when referring to your "older brother," you can't
say he's your hermano mas viejo unless he is, officially, a potential
Gray Panther. He is your hermano mayor-your "greater brother," in
age if in nothing else. A very common way of phrasing this idea is with
grande. iTienes hermanos mas grandest = "Do you have older siblings?" A "grown-up," in fact, can safely be translated as una persona
grande-"a big person."
I say "safely" because, as always when talking about age, there
are numerous ways to stick your foot in your mouth. Almost all of
them have to do with the word viejo, which, as noted, you would do
well to forget except in reference to buildings and such. With senior
citizens, the tendency in all cultures is to avoid calling them "old"-
thus terms like "senior citizens." There's nothing wrong with being
old-most of us take special precautions to become old ourselves-but
nobody seems to like being reminded that they've achieved this status.
So in Spanish you say things like Mi abue1a ya es muy grande, Es
una persona ya grande or Es una persona mayor instead of saying Mi
abue1a es vieja. Other words given for "old" by the dictionaries are
also borderline rude except when talking about, say, pyramids. These
include anciano and antiguo, especially.
Mi viejo and mi vieja, meanwhile, come in for frequent colloquial use. They can mean "the old man" (or father) and "the old lady"
(or mother), but more commonly they refer slangily to one's "guy"
(husband, boyfriend, or lover) or "gal" (wife, girlfriend, or lover). Viejo
(without the mil is also used slangily in many countries for "buddy" or
"man," regardless of the age of the person addressed. Vieja is not used
this way ever, though, since una vieja is a rude way of referring to a
woman, similar to "a broad" or even "a bitch."
ORDER
In this case, your choice is the same word but with different
genders: 1a orden and e1 orden. Which is which? Well, if you're traveling through Spanish-speaking areas, you should first learn the word for
"an order" of the sort that you submit to a waiter or waitress at a restaurant. It is 1a orden, the feminine form, and you should invent some
mnemonic gimmick to make this stick. E1 orden, the masculine form,
means "the order" of things-that is, their organization or sequence.
WHICH IS WHICH? 147
Ordenar as a verb, to mean "to order (in a restaurant)," is being used
increasingly in Latin America, but the proper word for this is pedir.
Ordenar really only means "to put one's things in order." "To order
(someone about)" would be mandar. A final note on "order": "in order
to" should be handled with para, never with an expression using
orden.
PIECE
This is a tricky one, and even most Spanish speakers won't be
able to explain the subtle distinctions-though they will invariably
and instinctively use the correct form. First of all, banish pieza from
your mind for all but a few usages, usually having to do with a "part"
of a machine or car. Think of it as a self-contained "piece," and not so
much as an arbitrarily determined or broken-off "fragment" of some
larger whole.
For fragments, the two most common words are trozo and
pedazo. Sometimes, these two are interchangeable-with a "stretch"
of highway, for instance, either trozo or pedazo will work (though
tramo would be an even better word here)-and other times they're
not. The difference seems to reside in whether the piece in question is
a plainly three-dimensional chunk (trozo) or not (pedazo), whether the
piece in question is useful in itself (trozo) or not (pedazo), whether it's
an indistinct division between part and whole (trozo) or a fairly clear
one (pedazo), and so on. An object breaks into pedazos, and you noro mally would ask for a trozo of something at the dinner table.
In this last setting, however, bear in mind that many comestibles come in rebanadas ("slices"), and that this is the preferred word
for "pieces" of bread, cake, pizza, and so on. A cute word for a "little
piece," "smidgeon," or "tad" is nadita-"a little nothing." The word
taiada is also used widely to mean "piece" or "slice," but in many
areas it also carries the connotation of a "cut" or "percentage" of an
illegal or dirty business.
PLAN
As a noun, "plan" equals plan pretty consistently, but keep in
mind the difference between plan and plano (a building's "floorplan"
or "design," a town "map"). To be in a certain plan is a colloquial way
of saying to be "in a phase," to be "on a kick." Estoy en mi plan limpiador = "I'm in my cleaning mood." For the verb "to plan" it is almost always better to forget about planear and use pensar instead, un-
148 BREAKING OUT OF BEGINNER'S SPANISH
less you are consciously planning a scheme or a strategy. Pensamos
irnos manana = "We're planning to leave tomorrow."
QUIET
In general, quieta translates better as "still"-as in "keep
still"-than as "quiet." So what's the right word for "quiet?" That
depends on what you're using it to say. "Keep quiet" is Gallate (for
a single noisemaker) or Gallense (for.a band of rowdy tykes). But be
careful with your tone: Gallate can convey the rudeness of an abrupt
"Shut up!" and in fact can be difficult to say without sounding rude.
With friends, a clever way around this problem is to address them in
formal tones: Gallese usted. Paradoxically, this more formal construction tones down the harshness of Gallate quite a bit. The formal construction is also commonly used on children, much as a parent might
say "John Joseph Doe, you quiet down" instead of "Be quiet, Johnny."
Where all else fails, "Shhhh" translates quite well as Shhhh.
Gallado as an adjective means "hushed" or "not answering,"
as in how a stadium crowd reacts when the visiting team scores a
touchdown. Gallar, the verb, can be used to convey "speechless." La
callaron = "They left him speechless." (This could also be rendered
Se qued6 sin palabras, or "He was left speechless.") A person who is
"quiet" by nature could be called una persona callada or una persona
de pocas palabras.
Many times, "quiet" in English is used to describe serenity,
calmness, peacefulness, and ease: "Let's just have a quiet night at
home, dear." For these cases, tranquilo is the correct choice. It also
covers "quiet" places. Busquemos un lugar tranquilo para platicar =
"Let's look for a quiet place to talk."
SAVE
Once you learn that salvar is rarely the word you want for "to
save," it becomes tempting to use ahorrar all the time. And it does
work for many usages: "to save time" (ahorrar tiempo), "to save
money" (ahorrar dinero), "to save (making) a trip" (ahorrar un viaje),
and So on. Still, it's not always the smoothest translation of these and
similar common English idioms. For "to save a trip," for instance, evitarse ("to avoid") could be used comfortably. Me evite un viaje ala
tienda cuando lleg6 con azucar = "I saved (myself) a trip to the store
when she (or he) showed up with sugar."
Another common use of "to save" in English is in the sense of
WHICH IS WHICH? 149
"to set something aside," and in these cases ahorrar should be avoided.
Instead, use guardar. Guardame un trozo para mas tarde = "Save me
a piece for later." Guardar is also the correct verb for "saving" ticket
stubs, receipts, and other items, and can often be translated "to hold
on to." Salvar is also a perfectly good word for "to save," of course, but
only in the very specific sense of "to rescue." "Lifesavers," both the
candy and the flotation device, are salvavidas.
SHORT
Ask yourself: "short" as in "not tall" or "short" as in "not
long"? In English, we lump these concepts together: "A short man
took a short trip." In Spanish, you have to distinguish between the
two: Un hombre bajo hizo un viaje corto. Breve can generally be substituted for corto, never for bajo.
SIGN
As you travel through the Spanish-speaking world, you may
find yourself asking directions and, in the course of doing so, asking if
there are "signs" marking the way. After a period of traveling there, of
course, you'll realize the quaint innocence of this question in many
places, but in the beginning at least you will need a word for "signs."
.What shall it be? The best all-purpose word for "road sign" is letrero,
and that's what you would generally ask for: iHay letreros~The word
r6tulo is also used (in some countries more than others), but generally
this word refers to the sign over a shop.
What about signo, you wonder? For the most part, ignore itoutside of math and grammar it gets little use. You will find it in stock
expressions like signa de interrogaci6n ("question mark"), signa de
admiraci6n ("exclamation point"), signa negativo ("negative sign"),
signa (de) menos ("minus sign"), and so forth. A more all-purpose word
for "sign" in the sense of an intangible "signal" or "indication" is seiial.
Es una buena seiial = "It's a good sign."
Other words that you may want to incorporate as you gain fluency are huella ("sign" in the sense of "trace" or 'clue"), seiia (a hand
"signal" or "gesture," including but not limited to obscene ones), indicio (safe for "indication"), muestra ("sign" in the sense of "evidence
of something"l, marca ("mark"), and so on. There's much overlap in
the use of these words, and for the most part you're safe-if not excessively precise-with just letrero and seiial. "To sign," of course, is firmar, and una firma is "a signature."
150
SKIP
BREAKING OUT OF BEGINNER'S SPANISH
In spoken English, "to skip" is used more than you might at
first think You can skip breakfast, skip to school, skip school, skip a
grade in school, skip ahead in math or, if you don't like math, skip it
altogether. Still, many dictionaries and textbooks neglect "skip" as,
well, not very dignified English. In Spanish, most figurative uses of
"to skip" have a good equivalent in the reflexive saltarse, which-like
"to skip"-is much used in daily speech. Me salte el desayuno = "I
skipped breakfast." Se ha saltado un rengl6n = "He's skipped a line."
"Skipping school" would call for some other expression, of which
many local variants exist-irse de pinta in Mexico, for instance.
THERE
As in the case of "here" (see above), the several words that are
used for "there" differ only slightly from each other. Ahi, the linguists
say, means "there" when "there" is "over there"-near the person
being addressed. Alli is "over there" when this is not near either the
speaker or the listener, but is generally within sight. Alla is "way over
there"-that is, far away, out of sight, yonder. As in the case of aea and
aqui, the linguists overlook the fact that most Spanish speakers tend
to use the three "theres" almost interchangeably.
Still, in certain expressions there is a correct "there" to use.
One of these is mas alla, which means "beyond" or "past." When you
tell the cab driver where you live, for instance, you might say Vivo
mas alla del parque ("I live past the park"J. When he stops at the start
of your street and looks back at you, you might say Mas alla ("Farther
on"). Don't confuse mas alld with el mas alld ("The Great Beyond"J,
though you might think that, from the way the cabbie is driving, that
is his ultimate destination.
Another note on alld: it is often used as a kind of shorthand
way of referring to a foreign country. A visitor to Mexico, for instance,
might be a little befuddled the first time he or she hears, completely
out of the blue, a question like iY tienen tequila allat ("Do they have
tequila there?"). On the other hand, alla is a handy substitute for
stuffy-sounding expressions like "where I come from "or "in my country." Where the context is clear, it works in a way as "back home."
Alld tengo esposa y tres hijos = "Back home I've got a wife and
three kids."
Finally, remember that in Spanish "there" is usually left out of
the traditional phone query, "Is So-and-so there?" Instead, you simply
ask iEsta Fulanot If you must get the "there" into your question, you
WHICH IS WHICH?
can ask tEsta par ahi Fulanol-though this can come off sounding
fairly informal.
151
WORKER
What should you call your co-workers? That may depend on
your mood, but in Spanish it also depends on what kind of work they
do. Obrero may seem to work well as "worker," but it refers almost
exclusively to "laborers" doing manual work. Within this grouping,
there are albaiiiles ("construction workers"), labradores (usually,
"field hands"), and iornaleros ("day laborers"). Trabaiador is the generic term for "worker," though it too connotes some actual, sweatproducing labor. For "white-collar workers" or "office workers" (i.e.,
the sweat-free positions), empleado and oficinista are close fits. Note
that empleado means more than just "employee" and is by itself a
fairly respectable job description, often suggesting a job in the public
sector. In fact, many government forms offer empleado as an occupational category all by itself.
\section{\emph{}}
\section{\emph{}}
\chapter{Say It Right}
Some everyday concepts in Spanish just seem to resist translation. As a result, many students of Spanish never seem to learn to say
them right. Perhaps common usage doesn't conform to the bilingual
dictionary's definition, or maybe there's just not a good dictionary
translation to be found. Regardless, you'll need to express these concepts, and the sooner you learn how to say them, the better. Here is a
selection of the commonest hang-ups for students of Spanish.
\section{\emph{}}
CHOICE
The problem here is really quite simple. In English, you
"choose" between "choices," but in Spanish you can't escoger between
escogibles or escogisiones because these awkward, horrible-sounding
behemoths don't really exist. Escogimiento does exist, but if you use
it, you're liable to be hit. Instead, use opciones and selecciones when
you want to translate, but keep your eyes and mind open for alternative constructions. "What's your choice?" would be expressed as iQue
escoges~ "There's no choice" would be No hay de otra or No hay remedio. "Good choice" = Buena selecci6n and "What are my choices?"
= iCuciles son mis opciones~
While we're on the subject, "choosy" is another concept that
confounds many students of Spanish-especially when you add in
"fussy" and "picky." In Spanish you have to leave the universe of
"choice" altogether to express this concept. One of the most common
(if least expected) ways of expressing it is with especial ("special").
SAY IT RIGHT 153
Ella es muy especial para 1a comida = IIShe's a very picky eater.1I
More formally, exigente (lldemandingll) works well, and in some parts
you may hear fastidioso for someone who is annoyingly fussy. (Note
however that fastidioso does not mean IIfastidiousll so much as lIannoyingll or IIbothersome,lI though it may be taking on the English
meaning more and more. See Appendix B.) In Mexico a slangy word
for lIannoyingly pickyll issangr6n. Es muy sangrona para 1a comida.
Likewise, No seas tan sangr6n = IIOon't be so picky. II
ENOUGH
Most students of Spanish learn early on that lIenoughll is bastante, pure and simple. So it will come as something of a surprise to
discover that lIenoughll is not bastante-or rarely. Some dictionaries
try to gloss over this problem, giving you a variety of forms-from
jbasta! to even bastantemente-to let you wriggle out of a jam. In
fact, bastante in the sense of lIenoughll almost never works as well as
su{iciente, which you should carve into your memory as the correct
word for lIenough.1I
This is not to say that bastante is not a useful word; it is,
and immensely so. But it is much closer to IIplentyll than lIenough,lI
as a few examples make clear. When the waiter is serving you brussel sprouts and you think that six is enough, you naturally would say,
IIThat's enough, thank you.1I In Spanish, this would be Es su{iciente,
gracias. Es bastante, gracias sounds almost like a complaint-IIWhew!
That's plenty. II A waiter, on hearing it, might even offer to take some
of your sprouts back. Ya basta will get your point across, but it comes
off sounding like a rude IIEnough already! II Tengo su{iciente dinero
means III have enough moneyll (to buy gas, or food, or whateverl,
whereas Tengo bastante dinero sounds like a bit of a boast: IIl've got
plenty of money."
A good verb to keep handy for situations involving money, especially, is a1canzar, meaning lito reachII but also lito be enough.1I It is
a very common word, and you will probably hear iTe a1canza para una
Coca? more often than iTienes 10 su{iciente para una Coca? Similarly,
III don't have enoughll is usually just No me a1canza. Note that a1-
canzar requires inverting subject and object. In English, you have the
money; in Spanish, the money reaches for you. A1canzar is also probably the best way to translate lito afford,lI a notorious bugaboo in translators' circles. lIean you afford new shoes? II would be iTe a1canza para
nuevos zapatos?
Bastante is very common as a modifier of adjectives, much as
IIplenty" is in colloquial English. Es bastante guapo = IIHe's plenty
154 BREAKING OUT OF BEGINNER'S SPANISH
handsome" (and not "He's handsome enough," which sounds like a
backhanded compliment). Es bastante listo can las computadoras =
"He's real clever with computers" (and not "He's clever enough"). For
"clever enough" and "handsome enough," you have to dig down deep
for suficientemente, as in the phrase Es 10 sufiCientemente listo.
Of all the typical usages of "enough," only "Enough already!"
wouldn't be disposed of better by some form ofsuficiente and would in
fact be rendered iBasta!
FEAR
Here is a classic minefield for the foreign speaker of Spanish,
Not only are there many alternatives to choose from, but most of them
depart from our customary perspective in English. Asustar, dar susto,
temer, temor, pavor, espantar, dar miedo, tener miedo-all of these
are used, and it's quite a trick to learn exactly how.
To start with, keep in mind that "fear" in Spanish is something that you "have" or "give," Thus, you aren't "afraid" but "have
fear." And you don't "scare," you "give fear." With this in mind, lie
down on the couch over there and proceed to analyze your fears. Fear,
in Spanish, can be broken down roughly as follows. A temor is a very
specific fear, almost more like a deep-seated, haunting feeling; temer
is the verb that goes with this type of fear. Thus Temo que alga haya
pasado = "I'm afraid something has happened." Mi temor es que escapen los animales = "My fear is that the animals may escape." Temer
is also used in polite expressions, just as "fear" is in English: "I fear we
may have arrived at a bad moment" would be expressed as Temo haber
llegado en un mal momenta or Temo que hayamos llegado en un mal
momenta. Note that all subordinate clauses built off of temer and temar require the subjunctive, because fear of what may happen or what
may be happening is naturally the province of the subjunctive.
Miedo, meanwhile, is your run-of-the-mill sort of fear. You
"have this fear" of the dark, of scorpions, of large fellows who drool,
and of public speaking. Of these fears you would say, Me dan miedo
or Tengo miedo de ellos. Perfectly natural. These things always "give
you" this fear, regardless of the circumstances.
A susta is more sudden than a "fear." Properly speaking, it is a
"scare," even a "shock." iQue susto me diste! ("What a scare you gave
me!") you tell your roommate, whom you found hiding in the closet
with a wolf mask on. This type of fear even gets its own verb: asustar
("to scare"), which means exactly the same as dar susto. Here, though,
you have to be careful about your English, since "to scare" is commonly used for both sudden fears and for fears which you always have,
SAY IT RIGHT 155
regardless of the circumstances. "Bears scare me" would normally
be Tengo miedo de los osos in Spanish. You might hear Los osos me
asustan, especially from kids, but it suggests that certain bears are in
the habit of sneaking up behind you and yelling "Boo!" in your ear.
When "to scare" means "to scare away," espantar is used instead of asustar. Thus an espantapajaros-literally a "scare-awaybirds"-is a "scarecrow." Espanto in Latin America is also a common
word for "spooks" or "ghosts"; the "haunted house" at the amusement
park is in many countries called La Casa de los Espantos.
Another problem is the word "scary." Some authors cite "scary"
as one of those words that simply doesn't have a Spanish translation.
(And two Spanish-English dictionaries I have at hand simply mistranslate "scary" as miedoso and pusildnime-both of which mean "cowardly.") I remember one student of Spanish describing a film he had
seen as espantosa, thinking it meant "scary" or "frightening." Instead,
it usually means "frightfully bad" or "dreadful," far from the speaker's
intent. Possibly the safest way to convey "scary" is with miedo. "It's a
scary movie" = Es una pelicu1a de miedo or Es una pelicu1a que da
miedo (or una pelicu1a de horror in the specific case of Friday the 13thstyle horror flicks). You might also hear tetrico for "scary," though
it tends more to "spooky" or even "eerie." Still, it can be the perfect
word to describe a lonely alley late at night. Aterrador is like "scary"
but considerably scarier; "terrifying" hits the mark.
Two other words round out our study of fears. They are pavor
and panico, the second being a slangy addition to the list. Pavor is an
intense, almost phobic fear, mingled with equal parts dread and loathing. Me dan pavor las tarantulas = "I'm terrified of tarantulas." Like
miedo, pavor is a long-term, constant fear, not a sudden shock like
susto. Colloquially, you may hear me da panico to mean much the
same as me da pavor, and me da terror fits this mold as well. Me da
horror is also used to mean "It disgusts me" or "It grosses me out."
Along these lines, you'll sometimes hear Me da cosa, a slangy phrase
that could be rendered "It gives me the creeps." Ese senor me da cosa
= "That man gives me the creeps."
GUESS
This word crops up a lot more in spoken English than you
might at first think. And until you break the habit of translating your
thoughts, you'll need a good selection of phrases for it. "To guess" often can be handled by adivinar. Adivina quien es = "Guess who it
is." iAdivina que! = "Guess what!"
But in English "guess" is stretched well past its original intent
156 BREAKING OUT OF BEGINNER'S SPANISH
in many colloquial instances. Obviously, we aren't actually guessing at
our own actions when we say, "I guess I'll be going now." For "guess"
in this sense, you would need to resort in Spanish to some similar
qualifier. Supaner works fine: Supanga que ya-me vay. In some other
examples, imaginarse might be called for. iVas air esta naehe~ Me
imagina que sf. = "Are you going tonight?" "I guess so."
Two other verbs-atinar and aeertar-also come into play
here. Both mean, more or less, "to guess right." Le atin6 a mi nambre
= "He guessed my name." Aver si le aeiertas al ganadar = "Let's see
if you can guess the winner." Both of these words are commonly heard
in Spanish, and you would do well to learn to recognize them. iDe que
pais eres~ Adivina. Pues, de Canada. iAtinaste! = "What country are
you from?" "Guess." "Canada?" "You got it!" Many times, when we
would be tempted to use adivinar, we should probably use atinar or
aeertar. "Let me guess" could be rendered either literally, Dejame adivinar, or more naturally, Aver si le atina ("Let's see if I can guess it"l.
HALF
Learning about "half" and "middle" in Spanish is essentially a
matter of learning one word-media-and how to use it. Only when
"half" is a noun does another word-mitad-come into play. At all
other times, media or a related form covers "half" and "middle." Examples: la mitad de la pizza = "half of the pizza" but media pizza =
"half a pizza"; una pizza media coeida (or una pizza coeida a medias)
= "a half-cooked pizza."
Students of Spanish mostly err in this case when they overuse
mitad. Note how much easier and smoother it is to say media batella
de vina instead of la mitad de una batella de vina. Both mean "half a
bottle of wine." Mitad is much used but fairly specifically, and especially where no noun is present. Dame la mitad = "Give me half."
lrse a mitades is a slangy phrase for "to split the cost of something" or
"half-and-half." Si eampramas una pizza, nas vamas a mitades = "If
we buy a pizza, we each pay half."
The adverb media is extremely common as a modifier of adjectives. It equates with "kind of" in English in the sense of "kind of
ugly," "kind of happy," "kind of drunk": media fea, media feliz, media
barraeha. Since it's an adverb, it never changes genders. For example,
Ella esta media barraeha and La pizza esta media eaeida.
A couple of media-related phrases are a mediadas de, en media de, and a medias. The first one means "around the middle of"
and is almost always used with time constructions: a mediadas de dieiembre, a mediadas del mes, and the like. En media de is "in the
SAY IT RIGHT 157
middle of" in the physical or figurative sense: en media de la calle
= "in the middle of the street"; en media de un gran lio = "in the
middle of a big mess." A medias is an adverb describing how something is done, and its best translation is "half-way" or just "half." iHiciste 1a tarea? La hice a medias. = "Did you do the homework?" "I
kinda half did it." Note that this is not the same as Hice 1a mitad,
which means "I did half of it." You often hear a medias to describe a
poorly done job; in this sense, it equates with "half-assed." iEsos albaiiiles 10 hicieron a medias! = "Those construction workers did a
half-assed job!"
HOW
This innocent-looking word earns special mention for the
widespread but incorrect use of como to translate common questions
like "How do you like it?" Even almost-fluent foreigners, especially
those whose native tongue is English, make this mistake. Como
should never be used this way. iQue taU can often be used instead.
iQue tal estuvo 1a pelicula? = "How was the movie?" Shorter and
sweeter: iQue talla sopa? = "How's the soup?" You can also construct "How did you like" questions with parecer (iQue te pareClO 0' . . ..l) and ( gustar i 'T' L e gusto. ' . . .. l) But iC' omo te gusto'l . ... IS 0
always an absolute no-no.
MAYBE
At least once in each of his films, the Mexican comic Capulin would answer someone's urgent request for information with a
thoughtful, unhelpful No 10 se, puede ser, a 10 me;or, tal vez, iquien
sabe?-in other words, "Maybe." There's one other common translation that Capulin left out, and that's quizQS or quizQ. Now you too
are equipped to be noncommittal and unhelpful. Use anyone of these
phrases to say "maybe" in good Spanish.
But wait, you say breathlessly, isn't there any way to differentiate among these many alternatives? No 10 se, puede ser, a 10 me-
;or, tal vez, iquien sabe ... ?But seriously, you should be careful about
overusing quizQS and neglecting a 10 me;or, which in many countries
is the common conversational way of saying "maybe." iVienes esta
noche? A 10 me;or. = "Are you coming tonight?" "Maybe." Tal vez is
also extremely common and all-purpose. QuizQs, on the other hand,
somehow sounds loftier and more grandiose. Puede ser que, puede
que, a 10 me;or, tal vez, and quizQS can all be used to start sentences.
158 BREAKING OUT OF BEGINNER'S SPANISH
All but ala mejor are usually followed by the subjunctive to highlight
the uncertainty being expressed. Tal vez este enojado =ItMaybe he's
mad. 1t Ala mejor no quiere = ItMaybe he doesn't want to. 1t On the
streets, finally, it is increasingly common to hear chance for Itmaybe, It
but don't tell your teacher I told you. iVienes esta nochet Chance. The
words you want to avoid are probab1emente and posib1emente, especially the former. Both are perfectly legitimate Spanish words, but
they're not as common as their English cognates. Were they as common, Capulin would have thrown them in for good measure.
MOTIVE
Life being what it is, there will come times when you slip up,
trip up, or flub up. It can't be helped. You're walking down the street
in, say, Santiago, happy as a lark, when your hand knocks into the ice
cream cone of a six-year-old girl and sends the cone scattering on the
pavement. The passersby eye you with suspicion. The six-year-old eyes
you with disbelief, then with unbridled hatred, and then starts to bawl.
You look around at the crowd assembling on all sides. Think quick:
what can you say to them?
In English, you might breezily announce ItSorry, I didn't mean
to lt or ItIt was an accident. 1t You might even add, ItIt wasn't on purpose. 1t In Spanish, your simplest, most universal excuse-and one you
should have at the tip of your tongue at all times-is Fue sin querer
(literally, ItIt was without wantinglt). Sin querer can also be added on
to any verb to mean Itaccidentallylt or Itnot on purpose. It La patee sin
querer = ItI kicked it by accident. It Sin querer barre tu nombre de 1a
1ista = ItI accidentally erased your name from the list. 1t To vary your
vocabulary (if you flub up a lot, for instance), you can use sin pensar,
par distracci6n, or invo1untariamente. Of course, if you can pronounce
that last one, you probably don't have to worry about flubs.
When you want to say something was Iton purpose,lt a prop6-
sito (or in some places de prop6sito) works well. But don't overlook
can querer, the opposite of sin querer and every bit as common and
as handy. La hice can querer = ItI did it on purpose. It Adrede, which
means the same thing, is less common, but you will hear it. Save deliberadamente for effect.
Be careful with accidente and its derivatives. Although both
par accidente and accidenta1mente are used widely, they sound a bit
forced, and neither gets the heavy work assigned to its English cognate.
Fue un accidente will get you away from the maddened crowd and the
SAY IT RIGHT 159
six-year-old. But Fue sin querer will do the job more smoothly, without overdramatizing the incident.
OBLIGATION
Here's one you simply must learn. You have to. Really, you
should. You'd better, anyhow. That is, you ought to.
The difference in degrees of obligation are a subtle issue that
you will master only after many years of listening to Spanish being
spoken. As in English, there are many ways to convey the notion, but
there is no simple way to explain which is the correct choice for any
given sentence.
Perhaps the most useful way to classify the Itmustsll is by the
strength of the obligation they imply. Intuitively, you can sense the
different strengths in English: IIYou must go" is greater than IIYou have
to go/' which is stronger than IIYou'd better go,o which is a tad more
forceful than ItYou ought to go/' which is up a notch from IIYou should
go. II In Spanish, a parallel ranking, in order of decreasing obligation,
might go something like this: Tienes que ir, Mas te vale ir, Has de ir,
Hay que ir, Mejor vas, Debes/deberias ir. Of these, you can safely discard Has de ir, which has a scolding schoolmarmish quality about it.
Hay que plus a verb, meanwhile, is very common but is limited to impersonal situations. You cannot say Hay que ir tu, for instance; you
would have to say Tienes que if tu. Mejor vas and Mas te vale ir are
both good translations of IIYou'd better go/' with the second implying
IIor else . . . .II
Deber with an infinitive is IIshould, 1I and whether you use the
present indicative (debes), the conditional (deberias), or even the past
subjunctive (debieras) seems to be largely a matter of personal preference. For all intents and purposes, Debes if manana, Deberias ir manana, and Debieras ir manana are equivalents.
Adding a de to deber creates another construction that is
hard to find a consistent English equivalent for. It is used when the
speaker has some sign or evidence that something "is supposed toll
be or IIshouldll be. An example will help here. No debe de haber llegada means ItHe (or she) must not have arrived yetlt and would be used
when the speaker is looking around an untouched room or notices that
all the lights are turned off, for instance. That said, many Spanishspeakers will use No debe haber llegado, without the de, to say the
same thing, and in some countries debe de seems to replace a simple
debe in many cases. Note though that by changing the verb to the
past, you change the meaning significantly. No debi6 haber llegado
160 BREAKING OUT OF BEGINNER'S SPANISH
would be saying "He shouldn't have arrived"-that is, the person did
in fact arrive but against the speaker's better judgment.
For events in the past, stick to a past form of deber with a past
participle for most constructions. Debias (or debiste) haber ido =
"You should have gone." If you instead use the preterit plus the infinitive-debiste ir-you are stressing more the obligation tllan the
missed opportunity. Debias haber ido a 1a fiesta sugges-ts you missed a
good time at the partYj Debiste haber ido a 1a fiesta suggests the same,
only a fraction stronger, and hinting maybe at "I told you SO"j Debiste
ir a 1a fiesta means it was your duty to go-the party was in your
honor, for instance. The fourth combination-imperfect plus infinitive, or debias ir-is not used, so forget about it. Perhaps the best construction for past obligations-that is, your "should haves"-is also
the simplest: hubiera plus the past participle. (Te) hubieras ido = You
should have gone. Hubieramos llamado = We should have called.
A final note on words of obligation deals with the word ob1igar itself. In Spanish, this word is much more widely used than "to
obligate" is in English. In fact, in many cases where in English we use
"to make" or "to force/' in Spanish you should be thinking obligar. La
po1icia nos oblig6 a sa1ir = "The police forced us to leave." E1 diablo
me oblig6 a hacer10 = "The devil made me do it." One way to remember to use obligar is to remind yourself that forzar is relatively rarely
used. It's a perfectly legitimate word, of course, but until you're accustomed to using obligar, you're better off forgetting that forzar exists.
USUALLY
The basis of the confusion here is the word usua1mente,
which does exist but which you should treat as if it didn't. Dictionaries tend to give it as the translation for "usually;" but in actual practice you could go a month or a year without hearing a native Spanish
speaker use it. It is easy to remember, and technically it's accurate, but
it sounds stiff and strange. You decide.
If you decide to avoid usua1mente (smart choice!), you will be
left with the slight problem of substitutes. Never fearj they abound.
Genera1mente is safe, but a mouthfulj norma1mente is easier and just
as good. Por 10 general is also a good choice. A verbal construction that
works very well involves the irregular verb soler. Sue1e pasar a las seis
= "It usually comes by at six."
While we're discussing the frequency of things, note that if
you're still translating from the English, you'll find your brain feeding
you lines like "most of the time" and expecting you to translate them.
This leads to hideous constructions like 1a mayoria del tiempo, as in
SAY IT RIGHT 161
La mayoria del tiempo s610 bebo agua ("Most of the time I just drink
water"). In Spanish, this sounds as if you have made a scientific study
rather than an off-the-cuff remark. In general, forget about translating
"most of the time" and all other "most of" expressions-"most of the
people," "most of my life," "most of the class"-and look for ways
around them. The use of casi ("almost") often provides a way out: casi
siempre = "most of the time"; casi todos = "most of the people";
casi toda 1a c1ase = "most of the class."
WASTE
The tendency here, once again, is to translate our English
thoughts too literally into Spanish. The proper word for "to waste" in
Spanish is desperdiciar-a real mouthful that, once you learn it, you
are justified in wanting to use as often as possible. Unfortunately, for
most expressions requiring "waste" in English, desperdiciar is not the
word. Generally speaking, simple old perder ("to lose") is. Thus "to
waste time" is perder e1 tiempo and "to waste an opportunity" is perder una oportunidad. The use of perder carries over to the nouns: una
perdida de tiempo = "a waste of time." Ma1gastar can also be "to
waste/' but perhaps translates better as "to misspend." Some dictionaries list echar a perder as a translation for "to waste," but correctly
this phrase means "to ruin," as in Echaste a perder e1 fuego ("You ruined the game"). Its reflexive form, echarse a perder, means "to go to
waste" or "to spoil," generally in reference to perishable foodstuffs.
With people, it also means "to spoil." Esttis echando a perder a1 nino
= "You're spoiling the boy."
\section{\emph{}}
\chapter{Spanish Roots}
Until now, we have been treating "Spanish" as a single unit,
a monolith unchanging over time, an easily dissected collection of
words, phrases, syntax, and morphology. It makes sense to approach it
this way, of course. The book that presents the entire history of Spanish, while trying to make modern Spanish intelligible to foreign speakers, would be a much, much longer book than this. I'm glad I don't
have to write that book, and I hope you never have to read it.
But a little history is useful and, for some students, fascinating. For the history of a language is also the history of the people who
spoke it, the voyages they took, the castles and forts they built, the
wars they waged, and the ideas they attacked and defended. In the case
of Spanish, it is the family history of more than 300 million modernday Spanish speakers. But it is also the history of Europe, of the New
World, of Western civilization, of the entire planet.
You may not have much need for an understanding of the linguistic differences between today's Spanish and what was spoken in
the kingdom of Castile a millennium ago. But it somehow adds to the
thrill of speaking Spanish to know that the language we speak today is
pulsing with the life of centuries. The language in its present form is
but a point on a continuum in the evolution of Spanish; by learning it,
and learning about the worlds that gave birth to it, we join the process
in a small but exciting way.
Spanish, of course, came from Latin, as did French, Italian,
Portuguese, Romanian, and several other less widespread European
languages. In fact, a good way to grasp the origins of Spanish is to examine where, why, and how it branched off from Latin. And for that,
we have to step back in time more than 3,000 years.
SPANISH ROOTS 163
A thousand years before the birth of Christ, the Iberian Peninsula-what today is Spain and Portugal-was already home to distinct tribes and incipient civilizations. At the mouth of the Guadalquivir River, the Phoenicians founded the city of Tharsis, or Tartessus,
which became known for its wealth and commerce, according to Old
Testament reports. In the tenth century B.C., according to the Book
of Kings, King Solomon's navy went every three years to Tharsis
"and brought from thence gold and silver, and elephants' teeth, and
apes, and peacocks." Four centuries later the Hebrew prophet Ezechiel
talked of the "merchants of Tharsis," who apparently conducted a
busy trade with Phoenecia (modern-day Syria) at the other extreme of
the Mediterranean.
The Greek historian Herodotus, in the fifth century B.C., refers
to the inhabitants of "Iberia," located presumably near the mouth of
the Iber (Ebro) River in southern Spain. The Greeks also established
colonies in southern Spain, and by the third century the Carthaginians (from what today is Tunisia) began to conquer the entire Iberian
peninsula. Their capital was Cartagena-"New Carthage./I Not until
201 B.C., with the conclusion of the Second Punic War and Rome's victory over Hannibal's Carthaginian forces, did Spain start to fall under
Roman rule. That date also marks the formal arrival of Latin in Iberia-the arrival, that is, of the antecedent of the Spanish language.
The Iberians, as we call them today, had a relatively advanced
culture that peaked in about the fourth century B.C., or right before
the Carthaginian invasion. A few examples of Iberian coins and sculpture survive to this day, and the remnants of their written language
still largely defy translation. Tharsis also boasted of a written language dating back, they said, some 6,000 years! The extant examples
of this language are to~ally unlike ancient Iberian but just as indecipherable. Their only common trait is that both seem to have originated in Africa.
Spain's early history (and, indeed, much of its later history) is a
function of its location between the European and African continents.
If the Iberians and others came' from North Africa, then other tribes,
including the Celts and Ligurs, came from the European heartland. In
fact, the Celts, who began settling the northwest of the peninsula in
about the eighth century B.C., began mixing with the Iberians, creating
a "Celtiberian" culture in north-central Spain.
of the pre-Roman groups, only the Basques have survived
as a separate ethnic group, and only the Basque language has reached
modern times more or less intact. The state of scholarship on the
Basque language reflects the mystery surrounding pre-Roman Iberia
as a whole. Scholars who struggle to locate the origins of a now extinct
Iberian orTurdetan language must also recognize that they are unable
even to trace the origins of Basque, which alone among we~t~~~Euro-
164 BREAKING OUT OF BEGINNER'S SPANISH
.~
i
pean languages is from non- Indo-European roots. Where, then, did
Basque come from? Once it was thought the Iberians simply became
the modern-day Basques, but recent scholarship has poked holes in
that theory. Now scholars surmise that Basque may have come from
Caucasian, a language spoken in modern-day Georgia between the
Black and Caspian seas. Or, others say, perhaps it is from a Hamitic
source, like certain Sudanic languages, with which it has similarities.
Then again, it may be from Coptic, an Afro-Asiatic language of Egyptian stock still used today in the liturgy of the Coptic Church. One
modern scholar has even suggested that Basque is but another Romance (Le., Latin-based) language!
\section{}
THE FATE OF LATIN IN IBERIA
The combined effect of Iberia's pre-Roman cultures and languages on the birth of Spanish is small but significant. Words taken
from Celtic, from Basque, and even from Iberian still linger in the language. From Celtic, early forms of such common words as cerveza, caballo, carro, camisa, and camino were born. Celtic place-names on
the Iberian Peninsula include Segovia, Evora, Coimbra, and Coruna, although all of these names have undergone considerable changes from
the original Celtic words. Basque has contributed a handful of words,
most of them uncommon, to Spanish, with izquierda being probably
the most used.
Other everyday Spanish words that can trace their lineage to
Iberia before the arrival of the Romans are barra, conejo, gordo, muiieca, perro, and sapo. Certain suffixes are also viewed by today's scholars as typically "Iberian" in origin [though not necessarily from the
Iberians themselves). These include -rra, -rra, -ago, -era, -iego, and
-asco. The ending -ez also dates from pre-Roman times and can still
be found in names like Sanchez, Lopez, Ramirez, Gonzalez, Rodriguez,
Velazquez, Perez, and so on.
Even some changes in "spelling" between Latin and Spanish
have been blamed on the residual impact of pre-Roman languages. The
conversion of the Latin f to a now silent h in Spanish has been explained as part of the legacy of these languages and can serve to begin
our study of Spanish's slow but steady departure from its Latin roots.
The Spanish words hijo, hacer, hoja, hundir, and humo, for example,
come from the Latin roots filiiis, fiicere, falla, fiindere, and fiimus. In
these cases, French and Italian remained truer to their Latin roots with
filslfiglio, fairelfare, feuillelfoglia, enfoncerlaffondare, and fumee/
fumo. Though now we look on these variants as "spelling changes," at
the time they were changes in the spokenlanguage, which lexicographers
SPANISH ROOTS 165
only later would get around to assigning a spelling to. Mostly, though,
the pre-Roman languages, written and otherwise, bowed to Latin and
had vanished by the year 100 A.D., only three centuries after the Romans' arrival. And Rome's relatively unmolested control of the Iberian
peninsula was to last another three centuries, giving Latin a firm foothold there that may never be shaken loose.
Latin, of course, was far from a "pure" language itself, and
Latin speakers borrowed freely from other languages, especially Greek,
to supplement their native tongue. And Vulgar Latin-spoken on the
streets of Rome and by the soldiers in the invading Roman legionswas a far cry from the literary Latin that has come down to us in Roman and other texts. Latin speakers' willingness to improvise probably
accounts for the ready reception of Iberian, Celtic, and Basque words
and morphology. The result was that by the third century the language
being spoken popularly was an even less pure, less "literary" Latin.
Except for a handful of appearances of "common folk" in Roman comedies, few written examples of Vulgar Latin have survived
to the present. Scholars, though, have been able to recreate it based
on how Latin looked a few hundred years later. The "sloppiness" of
its vocabulary and some of its forms also prompted linguists of the
time to write tracts condemning developments in spoken Latin. One
of these was written in the third century by a man named Probus,
whose work shed considerable light on how Latin was changing. Probus also became somewhat famous among later scholars because
nearly everyone of the "errors" and examples of bad grammar that he
railed against was subsequently incorporated-Probus be damnedinto medieval Latin.
Students of languages amuse themselves by watching words
change. In this era of Latin, there was amusement enough to go
around. New words were absorbed and old words took on new meanings. Laborare, a perfectly functional Latin word for "to work," was
gradually replaced by tripliare, referring to a three-sticked instrument
of torture or punishment used on recalcitrant slaves. Thus as work
became more tortuous, a new word was needed to emphasize the unpleasantness of the experience. And so a word was "born" into popular
usage, eventually turning into the modern word trabaiar, "to work'"
Other common words have similarly intricate histories and
show in passing how Spanish began to emerge from shared Latin roots.
The Latin word for "grandmother," for instance, was aviaj in France
the word was abandoned in favor of a Germanicized grand-merej in
Italian, nonna, evidently a Late Latin derivative from child's talk (and
the source for the word "nun"), was adopted j only Spanish was to hold
on to the original Latin, and then only after "sweetening" the word
by attaching the diminutive suffix -ola. From there aviola became
166 BREAKING OUT OF BEGINNER'S SPANISH
abuela, the modern Spanish word for "grandmother." Knowing this
history makes it curious to hear the form abuelita, which has begun to
replace abuela in many parts of the Spanish-speaking world-and for
the same reasons that aviola replaced avia. The word, twice diminished, can hardly get any sweeter!
Some words, in use in the protected worlds of the church or
the law, survived this period of change intact. Sometimes these "original" Latin words continued to exist in these closed confines while
on the streets new words were being coined from the Classical Latin
roots. Sometimes, too, society's more cultured elements returned to
the Latin to express themselves, refusing thereby to "lower" themselves to the level of Vulgar Latin. This reintroduction of "cultisms"
would continue for centuries after Vulgar Latin had evolved into Spanish (or, strictly speaking, Castilian). As a result "Classical Latin" and
"Vulgar Latin" words survive side by side in Spanish today: frio and
frigido, oreta and auricula, delgado and delicado, entera and integra,
and so on. .
GERMANIC SPANISH
At this point, around A.D. 400, Spain and its language seemed
destined for a fairly uneventful development. Latin would continue to
be used, and abused, and the language that would evolve would be but
a regional, distorted, and "vulgar" version of the Classical Latin of
Cicero and Caesar.
It almost worked out this way, but not quite. In the year
409 A.D., the Vandals (the original ones, for whom all later vandals are
named) crossed the Pyrenees into Spain. The Vandals were a Germanic
people, and their invasion of Spain was to be the first of a wave of Germanic invasions. The Vandals themselves only lasted twenty years
on the peninsula. Hot on their trail-and fresh from sacking Rome in
41O-the Germanic Visigoths reached Spain only a few years later.
Their arrival, in fact, pushed the Vandals clean off the continent. The
Vandals crossed into North Africa in 429, conquered Carthage ten
years later, and sacked Rome themselves in 455. The Visigoths, for
their part, completed their conquest of Spain under their king, Euric
(466-484), and placed their capital in Toledo. Their stay was to last a
little longer than the Vandals'-some three hundred years, in fact.
The Visigoths' language was, of course, a Germanic one,
which meant that it shared an ancient Indo-European heritage-but
little else-with Latin. Still, the linguistic shock of the Germanic conquest was considerably less than might have been expected. First, the
Visigoths already spoke Latin and were themselves quite "Romanized" :
SPANISH ROOTS 167
as a result of being based for a full century in Toulouse before sacking
Rome and spreading into Spain. Second, the Visigoths kept to themselves ethnically, especially at first, swearing off mixed marriages. And
third, the Germanic languages had been filtering into Latin since at
least the first century A.D., which explains why many of the Romance
languages have parallel Germanic constructions.
Still, the Visigoths did leave their mark linguistically in proper
names, place-names, and in words associated with what they did best:
fighting. The word guerra ("war") is of Germanic origin, as are orgullo
("pride"), riqueza ("riches"), rabar ("to rob"), ganar ("to win"), and
bandido ("bandit"j. Of the place-names attributed to Germanic influence, Andalusia is probably the best known; it is but the Arabized
form of Vandalus, referring to the Vandals. And the proper names Alvaro, Rodrigo, and Fernando are "Hispanicized" versions of the Visigothic names Allwars, Hrothriks, and Frithnanth.
The lasting significance of the Visigoths on Spain, however,
had more to do with politics than with linguistics. Their conquest of
Hispania, for instance, effectively severed its ties with Rome for nearly
three centuries, leaving the Latin on the peninsula to develop in isolation and thus stray further from the Latin being spoken elsewhere. By
conquering most of the many disparate tribes on the peninsula, the
Visigothic lords also accomplished the first unification of Hispania,
giving birth to a larger political entity that later groups would struggle
to re-create.
ARABIC SPANISH
The three-century domination of Hispania by the Visigoths
ended in the year 711, when Arab forces (or Moors, as the Arab/Berber
Moslems on the peninsula were called) routed the Visigoth force.
Within seven years the Moslems, whose founder and prophet Mohammed had died only eighty years before, had conquered virtually the entire peninsula. The Visigoths stayed on under Moorish rule, enjoying
the tolerance of the new rulers. Science, art, and philosophy flourished
under the Moors, who promoted (or at least presided over) a bilingual
culture in which the Romanized groups became progressively more
Moorish and the Moors increasingly Romanized. The first group is
known as the Mozarabs (in Arabic, "would-be Arabs"j, the second, as
the Mudejars ("those who have been allowed to stay"-after the reconquest began, primarily). Hispania under Arab rule became a hotbed of
scientific activity, leading Europe in this sense and attracting some of
the best minds from around the continent.
The bicultural aspect of Moorish Spain loses none of its power
168 BREAKING OUT OF BEGINNER'S SPANISH
to astound more than a millennium later. Arab scholars wrote treatises
in Latinj Christians baptized their children with Arab names. Christian scholars, writing in Arabic, expounded on the fine points of their
religion. In Moslem Cordoba, churches, mosques, and synagogues
could be found in close proximity and operating openly.
The language, of course, opened its arms to embrace such
diversity. The legacy is a Vulgar Latin full of "Arabisms"-more
than 4,000 in all have survived in Spanish to our day. Many of these
words were to pass through Spanish (or more commonly, through
French) into English. A short list of the more indispensable ones
barely scratches the surface: azul ("blue"), escarlata ("scarlet"), limon ("lime"), naranja ("orange"), adobe ("adobe"L talco ("talcum
powder"L azar ("hazard/' "chance'l aceite ("oil"), cero ("zero"), cifra
("cipher/' "figure"L ajedrez ("chess"L cenit ("zenith"L nadir ("nadir"),
jazmin ("jasmine't azucar ("sugar't azafrcm ("saffran"L zanahoria
("carrot"), aduana ("customhouse"L tarifa ("tariff"), and arroz ("rice").
Ojala, which translates to English roughly as "I hope/' is but
a Hispanicized version of the Arabic expression wa-sa Allah (or in
sha'allahl, meaning "may Allah wish it."
Another vast group of Spanish words of Arab origin are those
beginning with al-, which is simply the definite article "the" in Arabic. This list also includes a number of words well known to English
speakers, such as algoritmo, algebra, alquimia ("alchemy"l, aleohol,
alcoba ("alcove"l, and almanaque ("almanac"). With a little imagination a few other al- Spanish words from Arab roots will look familiar
to us as well: aleanfor ("camphor"), algodon ("cotton"), and almirante
("admiral"). Some al- words you will need to learn in Spanish, if you
haven't already, include alllier ("pin"), almacen ("warehouse"L almohada ("pillow"), alcalde ("mayor'l alfombra ("rug't and almorzar
("to eat lunch").
Many of these"Arab" words are really just Arabized Latin and
Greek wordsj others are words that Arabic took from Persia, China, India, and Sumatra. But all of them entered Spanish (and some entered
French and Italian as well) through Arabic, and the vast majority of
them entered during the period of Moorish rule on the Iberian Peninsula. During this time the Moors also contributed mightily to Spanish
place-names, such that some of the places we consider most "Spanish"
are, in fact, Arab toponyms: Caceres, Gibraltar, Guadalquivir, and
Guadalajara, for instance. Many others are Arabized forms of Latin
names: Sevilla from Hispalia, for example, and Zaragoza from Caesaraugusta.
The kingdom of the Moors was to last almost 370 years in Toledo, some 500 years in Cordoba and Andalusia, and an astonishing
770 years or more in Granada. When the Moriscos, or "leftover" Span-
SPANISH ROOTS 169
ish Moors were forcibly expelled from Spain in the seventeenth century-more than 300,000 were "sent back" to Africa between 1609
and 1614-the "Arabs" had been in Spain three times as long as the
"English" have been in America.
CASTILIAN SPANISH
The stage is now set, in our lightning recapitulation of Iberian
history, for the "reconquest" of Spain by Romanized Christian forces.
But to understand well what was to happen to Spain and Spanish in
later centuries, we should pause here andJook at what else was going
on in Spain while the Moors held control of the bulk of the peninsula.
Beyond the mountains to the north and northeast, groups that
had managed to remain free of the Moorish yoke settled into small
kingdoms. To the northwest, the kingdom of Leon considered itself the
direct descendant of the Visigoth reignj to the northeast, the largely
Basque-speaking Navarre kingdom extended well into Francej south of
Navarre, a tiny kingdom called Aragon, speaking a Vulgar Latin much
like that of Leon, took hold under Navarre's protection and began an
independent expansionj Catalonia (Cataluna), to the east, was basically
part of France this whole timej Portugal set itself off as a kingdom and
eventually broke away altogether in the twelfth centurYj in Moorish
Spain itself, finally, the Christian Mozarabs continued to live in a variably bicultural and bilingual setting. Their language-the Mozarabic/
Visigothic/Vulgar Latin of southern Spain-evolved relatively little in
comparison with the tongues being spoken in the northern part of the
peninsula.
Lastly, a tiny kingdom carved out a niche for itself in the uninviting borderlands between Leon to the northwest, Moorish Spain to
the south, and Aragon to the northeast. This kingdom, called Castile,
was made up of a feisty, stubbornly independent people who fought
encroachment on all sides and with ever-increasing success. They resisted the Visigothic trappings and laws of Leon but refused to completely "Arabize" themselves either. Their language was coarse-unlike the relatively uniform dialects of Vulgar Latin being spoken on all
sides. In both Aragon and Leon, on either side of Castile, people said
mullerj in Castile they said mugerj in modern-day Spanish we say
muier. In both Aragon and Leon people said ferit, feito, and uelloj in
Castile these words were her-ir, hecho, and oio. In short, from the
standpoint of the rest of the peninsula's residents, the Castilians had
a strong and bizarre accent. And, thanks to political developments, in
a few short centuries almost all of Spain would have it as well.
From as early as about 1000 A.D., manuscripts from Castilian
170 BREAKING OUT OF BEGINNER'S SPANISH
monasteries show translations-in the form of margin notes or glossesfrom the Latin to the Vulgar Latin in use at that time. These translations are taken by many scholars to represent the first written record
of Castilian, and thus to mark the birth of the Spanish language some
1,000 years ago.
The Castilians were to become the "reconquerors" par excellence of the Iberian Peninsula. In 1029 the kingdom fell under Navarran control almost by default, and Ferdinand, the son of the Navarran
king, inherited it and declared himself Ferdinand I, king of Castile. His
neighbors in Leon didn't think much of Castile having a king and tried
to take the wind out of Ferdinand's sails, only to be defeated themselves in 1037. Ferdinand I thus became the king of Castile and Leon
and went on to annex a chunk of Navarre at the expense of his brother,
who had inherited it and whom Ferdinand murdered.
Castile thus became a regional power, and though the union
did not stay in place permanently, Castile was strong enough now
to begin expanding southward, into Moorish country. The advance
of Castile was like that of a wedge splitting the peninsula in two.
In 1085, Castile took Segovia, Avila, and Toledo and two years later
moved its capital to the last of these.
THE SPREAD OF CASTILIAN
The advance of the language was slower than the military and
political thrust. A century and a half after Castile's "reconquest" of Toledo, for example, Arabic was still being used for official documents
there. Castilian forms and pronunciations were being used in the "New
Castile" based in Toledo, but simultaneously many Mozarabic words
were infiltrating Castilian. Overall, Castilian was an open and innovative language, absorbing influences as its speakers advanced politically
and geographically.
Not until the reign of Ferdinand III (1217-1252) was the use
of Castilian imposed on New Castile. It was also Ferdinand III who renewed the spread of Castile's power, "liberating" Caceres to the west
(1227), Valencia to the east (12381, and Cordoba (1236) and Seville
(1248) to the south. By the end of his reign, Moorish Spain was backed
into one kingdom-Granada-and for his trouble and for his efficiency,
Ferdinand III was to be granted sainthood. Granada was defeated 250
years later (its liberation and the unification of Spain was the big news
on the peninsula in 14921 by Isabel. Isabel, queen of Castile, had a decade earlier married Ferdinand V, king of Aragon, thereby joining those
two kingdoms and setting the stage for Spain's final unification.
Castile's conquest of Spain, except for Granada, was essentially completed by the mid-thirteenth century. Consolidation came
SPANISH ROOTS 171
next, and in the case especially of the language, the man for the task
was Ferdinand Ill's son, Alfonso X, a.k.a. "The Learned" (1252-1284).
Alfonso was a studious sort, not given to military escapades and not
very successful in those he did attempt. Instead, he surrounded himself
with sages, scientists, poets, and historians, and set up his court at Toledo as a patron of scholarship. Though Alfonso took pains to assure
the harmonic mingling of Spain's special "triple heritage," made up of
Christian, Hebrew, and Arabic learning, he adopted Castilian as the officiallanguage of the court and the kingdom, replacing Latin.
The importance of the decision also resided in Alfonso's work
(with the help of his hired scholars) in unifying criteria for his new language. Thus for Castilian to have any meaning as a language, someone
had to decide which form of a word-the Mozarabic, the Old Castilian,
the Aragon, the Leon-was to be adopted thereafter for incorporation
into New or "Toledan" Castilian. His choices naturally became the official ones, not so much by decree as by consent-and by the fact that
the court commissioned much of the literature and published most of
the peninsula's written texts. The growth of Castilian literature, promoted by Alfonso, marked the definitive emergence of the Spanish
language.
In certain essential respects, the Spanish written in 1200 and
1300 is much the same as the Spanish of today and can be read with
relatively little difficulty by modern Spanish speakers. Cantar del mio
Cid, dating back to about 1200 but not written down until 1350 or so,
is quite intelligible to any modern Spanish-speaking personj Chaucer's
Canterbury Tales, on the other hand, was written close to 1400 but
is usually "translated" for present-day English speakers. Certainly by
1500, except for the influx of foreign words, the shape and structure of
Spanish was pretty well carved in stone.
In fact, foreign influences had never stopped. From roughly
1100 to 1300, a vast number of French words entered through the
north of Spain, brought by religious "tourists" on pilgrimage to Santiago de Compostela in Galicia. The route they took across the north
of Spain, through Navarre, Castile, and Leon, was littered with Gallicisms, some of which survived the intervening centuries and persist in
modern Spanish. These include numerous -aje words (mensaje, homenaje, coraje, viaje, salvaje) and other terms related to the traveling life:
meson, jornada, jardin, and ligero. The 1400s, in contrast, were years
of great Italian influence, as Spanish scholars strove to catch up withor even keep up with-the advances of Renaissance Italy. The incorporation of Italianisms was an ongoing process, lasting several centuries,
that counted among its harvest such everyday words as marchar, millon, banca, balcon, diseiio, modelo, capricho, novela, cortejar, charlar,
manejar, pedante, grotesco, piloto, soldado, and alerta.
The close of the 1400s was typified by that dramatic year
172 BREAKING OUT OF BEGINNER'S SPANISH
1492, which saw the expulsion of Spain's Jewish population (the dispersed Sephardic Jews), the discovery of America, and the fall of Granada and reunification of Spain. In this year as well Nebrija published
his Gramatica de la lengua castellana, which was an attempt to standardize Spanish and the first published grammar in any modern language. In short, while Spain was extending its reach geographically, it
was slowly closing itself off to "outside," non-Christian intellectual influences. These years were to witness the start of the Spanish Inquisition as well, although its influence was not fully felt in Spanish culture until the late 1500s.
INTERNATIONAL SPANISH
The 1500s began with Spain asserting itself internationally.
The conquest and colonization of America took place in these years,
and newly unified Spain was assured a prominent place in Europe.
Castilian was by the 1530s the language spoken throughout unified
Spain. Spain was ruled from 1516 to 1556 by Charles V (Carlos I of
Spain, known in the Spanish-speaking world as Carlos Quinto). Born
in Ghent, in present-day Belgium, Carlos didn't set foot on Spanish
soil until he was eighteen years old, but he is still considered one of
history's great defenders of the Spanish language. By 1550, Italian gentlemen took pains to learn Spanish, which for the next one hundred
years would reign as Europe's preeminent language. Around 1660, for
example, anybody who was anybody at the court of Brussels spoke
Spanish. .
The reign of Carlos's son, Felipe II, ran from 1556 to 1598 and
marked the high point for the Inquisition and Spain's xenophobia. It
also marked the low point in many other respects. Felipe II was a repressive, autocratic, and intolerant king even by sixteenth-century
standards, and the religious zeal that characterized his rule earned
Spain a reputation for backwardness and barbarism that it has never
overcome in some parts of Europe. Paradoxically, Spain's Siglo de Oro,
or Golden Age, began in earnest under Felipe. It was a century that saw
the emergence of such cultural greats as Cervantes, Gongora, Lope de
Vega, and Quevedo.
The exploration and conquest of America had a lasting effect
not just on Spain but on its language. Spanish became filled with
"Americanisms," many of which it then passed on to other European
languages. The names of fruits, animals, and American products came
into Spanish from the Carib, Nahuatl, Quechua, and other indigenous
tongues (see Chapter 14). From the ongoing exploration and slave trade
in Africa came words from that continent into Spanish, with banana,
SPANISH ROOTS 173
conga, bongo, and samba among the ones that have survived to the
present.
Spain's importance geopolitically in these years translated into
what we now view as an excess of pride and arrogance. The same attitude that permitted the mass expulsion of 300,000 Arabs from Spain
in the period 1609-1614 appears in the language as well. It was in
the early years of the 1600s, for instance, when a new form of respect,
usted, was invented. Until that time, Spanish speakers had used vos,
the second-person plural, in respectful forms of address. This construction shares the logic of the "royal we" in English, by which royals refer to themselves in the plural. At some point their vassals must have
figured that if the king or queen considered himself or herself to be a
plural, it was not the vassal's duty to inform the monarch otherwise.
And thus plural forms of address like vos (vous in French, voi in Italian) were born. For certain Spanish lords of the seventeenth century,
though, vos had become too widely used, and a new form of address
was "needed" to distinguish those who genuinely merited adulation.
The form proposed and adopted, in a matter of decades, was vuestra
merced ("your grace"l, which in just twenty years of popular use was
chopped down in size to a more manageable usted.
By the 1700s, the effects of centuries of intellectual isolation
and repression were clearly showing. The Siglo de Oro was long gone,
and nothing even close to its output was accomplished through the
1700s. Instead, the eighteenth century was a good one for lexicographers and grammarians, as creative energy and talent was channeled
away from more risque endeavors. For example, in the years 1756,
1759, 1762, and 1764, respectively, the Inquisition banned in Spain and
its possessions the works of Montesquieu, Diderot, Voltaire, and Rousseau. The Royal Academy of the Language, meanwhile, was founded in
1713, and spent the century publishing reference works-the equivalent of decrees-on grammar, spelling, and literature. These had the
effect of establishing official spellings for Spanish words (much as
Noah Webster did for American English half a century later), and although Spanish has certainly changed since this period, its changes
have been minor ones.
French influence heightened in the 1700s, and in fact by century's end Spain found itself controlled outright by Napoleon. By far
the largest number of Gallicisms entered the language during this period, a collection of words that includes, for starters, asamblea, burocrata, controlar, jinanzas, pantalon, chaqueta, and equipaje (another
-aje wordl. French phrases and constructions were imported in this era
as well; for example, the expression hacer el amor in the sense of "to
court" or "to woo" became popular. (Its other, more modern sense is
attributed to the influence of American English.) One wonders how
174 BREAKING OUT OF BEGINNER'S SPANISH
many more French words would have entered in this period if Montesquieu, Diderot, Voltaire, and Rousseau had not been blacklisted!
Napoleon's misadventures in Europe gave Spain's colonies the
opportunity to slip away in the early 1800s, but the effect of this on
the language was minimal. American Spanish was already firmly established, a product and stepchild of the Andalusian Spanish spoken by
many early settlers. Thus the practice of seseo, or pronouncing c and z
as s (meses rhymes with veces), is universal in America and dominant
in Andalusia, but uncommon in the rest of Spain. In some parts of the
New World, expressions that died out at home (in Spain) were kept
alive. Other expressions were born in the colonies and made their
home there, never to infiltrate Castilian Spanish. The result, today, is
an American Spanish that is considered a distinct (though by no means
uniform) dialect, one easily understood by Spaniards if not "approved"
at all times by their official institutions-particularly the archconservative, arch-Castilian Royal Academy of the Language.
CONSERVING SPANISH
Conservativeness in language is a practice that tends to make
its practitioners look silly a few centuries or even a few decades later.
Resistance to change is almost invariably a losing battle. In the case of
Spanish, scholars and students of the language understandably seek to
preserve as much as possible the legacy of one thousand years of written tradition. They want speakers of the language one hundred and five
hundred years from now to be able to read and enjoy Cervantes, Garcia
Lorca, and Neruda in the original-not in "translations to modern
Spanish." And that, most would agree, is a goal worth looking a little
silly for.
The defense of Spanish in our time has largely become a stand
against the influx ("contamination," some would say) of English words,
a reflection of the fact that English, for the better part of the twentieth
century, has been the dominant language internationally (as Spanish
was from 1550 to 1670, more or less). That defense, and its trials and
tribulations, is the subject of Chapter 14.
This defense of Spanish, like all defenses that seek to protect a
language from its speakers, has little chance of final success. That said,
it is even less likely that an upstart language like English-useful in
commerce and science, trendy in popular entertainment-will ever
supplant Spanish when it is time to talk of the stars, of the gods, and
oflove. For these concepts, for words of passion and glory, power and
drama, Spanish need look no further than its own vibrant history.
\section{}
\section{}
\section{}
\chapter{The big Mix}
Spanish, as we have seen, is in a way a melting pot of linguistic influences. All languages are. And few would deny that Spanish has
benefited from its contacts with other languages. Without its prehistoric Iberian, Celtic, and Basque influences, without the influx of Germanic terms during the reign of the Visigoths, without the thousands
of Moorish words imported from Arabic, without the Italianisms of
the Renaissance and the Gallicisms of recent centuries, Spanish would
simply not be Spanish as we know it and speak it today. It would be
Modern Latin.
In the twentieth century, English is the language that has most
influenced Spanish (and about every other world language, for that matter). U.S. commercial, political, and cultural dominance has in the field
of languages resulted in the steady infiltration of English words into
Spanish. What is thought up on Madison Avenue, put on screen in
Hollywood, or schemed in Washington almost inevitably turns up, in
one form or another, in spoken Spanish. And though the tendency to
adopt Anglicisms is often presented as an acutely Latin American phenomenon, any visitor to Spain or reader of Spanish magazines will see
that English has insinuated its way deep into the heart of Castile
as well.
As an English speaker learning Spanish, you are present on the
front line of this phenomenon. And the military metaphor is not out of
place. For there is a battle going on between perfectly adequate Spanish
words and aggressive English invaders. It is a battle that causes lexicographers to lament and purists' blood to boil.
Avoiding Anglicisms in your Spanish is generally synonymous
176 BREAKING OUT OF BEGINNER'S SPANISH
with being a careful and respectful speaker of the language. As a beginning speaker (or even a lazy advanced speaker!L you will find yourself
tempted to use English words as a shortcut or a crutch. At other times,
unknowingly, you may give Spanish words a meaning that their cognates have in English-but not in Spanish. Finally, even as a nearfluent speaker you will have to guard against using English constructions and translating English phrases "incorrectly." Often these are
borderline judgments, as when choosing between peor es nada and
meior que nada for "better than nothing/' or un dia de estos and
uno de estos dias for "one of these days." It is the difference, in other
words, between being correct and being more correct. Proceed attentively, and let your ear be your guide.
These adjustments take time and attentiveness. And some English loanwords you may find useful or even necessary to your style
of speech, especially when everyone around you is using them, For
the most part, though, it is probably better for the "contamination" of
Spanish to occur at its own rhythm, without your contribution.
Scholars have long debated the reasons for the adoption of
foreign words, but the only clear reasons are the obvious ones. When
something is invented or introduced to a culture, for instance, it often
comes with a word-or an entire nomenclature-attached. Computers
are a good example: software, hardware, and mouse are all "Spanish"
words now. Similarly, when the word toronia was adapted from the
Arabic, bur6crata from the French, and canoa from the Arawakan, it
was because the products or concepts they described were new to the
Spanish-speaking world.
The adoption of foreign loanwords is not always so simple
a matter. In the fifteenth century, for instance, the Spaniards found
themselves in need of a word to describe "mustache/' it not having occurred to a Spaniard until that time that such a thing could exist without the rest of the beard. At first, the Germanism bigote found favor
(itself a word that some claim came via Norman French from the English expression "by God/' presumably for what the English said when
they first saw a Norman wearing one). A century later, the Italianism
mostacho (originally from the Medieval Greek moustakiJ became the
rage. Eventually, the Germanism won out, though mostacho lives on
in Spanish as an infrequent archaism.
As the pace of inventions has accelerated in recent decades,
the need for new words to describe them has grown. But usually there
are numerous options for fulfilling this need. As the use of motorcars
spread, for example, the construction of special roads to accommodate
them began in earnest. In English, one of the words given to these roads
was "highway." In Spanish, new words were needed as well, Which
word would be used? Had the sway of English been as powerful then,
THE BIG MIX 177
Spanish speakers might have undertaken the direct importation of
"highway," as in the Spanishized jaigiiey or some similar monstrosity. A generation earlier "highball" had been imported and rewritten
as jaibol. Fortunately, perhaps, that didn't occur. Another possibility
might have been to translate the meaning of the English word into
Spanish, again presuming English was influencing the selection. This
would have produced a phrase like camino alto for the English word.
But this didn't happen either. A third choice would have been to invest
an existing word or words with the new meaning. This in fact did happen with two words-carretera and calzada-both of which predate
highways but are now used in modern Spanish to describe them. Other
words came and went. As late as the 1950s, for instance, bilingual
dictionaries gave camino real as a translation for "highway," but this
phrase has since plummeted from common use. Lastly, an altogether
new word, or "neologism," could be coined to handle the novel concept of a high-speed road. Autopista was just such an invention.
Often the importation of a concept can result in a combination
of linguistic responses, or even different responses in different places.
One word may be coined in Mexico, for instance, while another is
coined in Puerto Rico, a third is invested with the new meaning in
Chile, and the English word imported lock, stock, and barrel in Spain.
A 1974 study of Latin American students in the United States explored
this process and found, among other things, that the concept "sneakers" had led to seven different words in eighteen different countries:
tenis, zapatos de tenis, zapatillas, champion, zapatillas de goma, zapatos de caucho, and zapatos de lana. Other "imported" concepts had
as many as nineteen different Spanish names! Dictionaries-and in
particular the dictionary of the Royal Spanish Academy-are reluctant
to include a word until it has shown some staying power. As a result,
speakers of the language are left with no guidance in the meantime and
must invent, invest, adopt, or adapt as they can.
Practical necessity, moreover, is but one of the many mothers
of the invention of new words and the importation of foreign ones.
More often than many like to admit, foreign words are imported because using them makes the speaker or writer seem more suave, chic,
and debonair. Foreign words have a certain je ne sais quoi that can
make them simply irresistible. We use them not because English or
Spanish doesn't have the words we need, but because we are a little
bored with the words we have. We feel the urge to break out of the
angst and the Sturm und Drang of our daily drudgery, so voila, we import "new" words-ad nauseum. In short, we like to show off.
Sometimes, too, we use foreign words because we want specifically to convey a sense of the exotic. Advertisers are especially
guilty of this, but they are not the only ones-and they can't always
178 BREAKING OUT OF BEGINNER'S SPANISH
be blamed. Would you sooner choose Boca Raton or Mouse Mouth for
a Florida honeymoon? Need something dramatic to tell your daughter
and son-in-law as they head off for their week in Mouse Mouth? How
about "Bon voyage!" instead of "Have a good trip"? What is more explicit than "blitz" to describe eleven grown men attacking a quarterback? And after a nice, refreshing shower, would you rather sprinkle
yourself with eau de toilette or toilet water?
In Spanish, too, words are imported left and right, and the
state of world superpowers dictates that many of those words are English ones, for the time being. Someday, to paraphrase Kurt Vonnegut,
Chinese linguists and historians might find the phenomenon quite
interesting.
\section{}
ANGLICISMS
The Anglicisms that enter Spanish do so via different paths.
Some enter exactly as they are written and spoken in English and remain that way. Others are modified to meet Spanish rules of grammar,
pronunciation, and spelling. Some are misappropriated or misspelled.
Some are introduced and then acquire new meanings.
Though there is great variation, imported words tend to follow
a fairly predictable pattern, Imports in Spanish-speaking countries often start as the province of the educated middle and upper classes, who
generally speak a little (or a lot of) English and know how to spell it.
After a word is imported as a foreign term, it must withstand the test
of time. If there is a need for the word, real or perceived, it may survive
and spread into the general public. There it is twisted and shaped until
it slides more comfortably off the tongue and is written as it sounds to
each ear, half a dozen different ways, until a uniform spelling is settled
on or imposed. Then, much later, it may be approved by the gatekeepers of the Spanish language at the Royal Academy. And then again, it
may not.
In English-speaking or bilingual areas, the importation of
words seems to be almost a haphazard process: anyone can play, and
most do. It is these areas-the borderlands between Mexico and the
United States, the streets of New York, the barrios of Los Angeleswhere hundreds if not thousands of Anglicisms are bred. The words
that emerge from these cultural and linguistic cauldrons are often the
ones that seem most grating and most unnecessary to language purists.
As the two languages mingle freely, terms like edible (instead of comestible) and marqueta (for "market") begin to appear as a result of
the, well, melange.
These pochismos, as the border loanwords are called in Mex-
THE BIG MIX
179
ico, are hard to feel neutral about. On the one hand, it is difficult to
stand in the way of a language's inevitable and inscrutable evolution.
Linguists have found that the incorporation of loanwords follows a system, and argue that the invention or adaptation of English words by
immigrants is in many cases a defensible response to the trying conditions of immigration. Concepts that are common in the dominant culture but unfamiliar to the minority culture frequently require new
words to express them. The loanword-loaded Spanish of the border
areas, moreover, has become an integral part of the bilingual and bicultural identity of many of its speakers.
On the other hand, it is hard to justify pronouncing any English word that pops up with a Spanish accent and calling it Spanish.
Most people try to avoid the extremes and decide for themselves which
words are acceptable to them. A quick study of Appendix A~an annotated list of many of the more frequent imports-will help you draw
your own line.
An English word's transition into Spanish isn't always a
smooth one. In fact, there's many a slip twixt English and Anglicism.
The word crack, for instance, appears on the financial pages in the
Spanish-speaking world to refer to a stock market "crash." Similarly,
in Mexico a car's "brake lights" are called luces de stop, or "stoplights," which of course are a different thing altogether. "Hitchhiking,"
a practice almost certainly introduced to Spanish-speaking countries,
generally goes by the name of autostop, though it is not known by that
term in any English-speaking country.
Other cases of the "wrong" word being imported include lifting ("facelift"); clip ("paper clip"J; smoking ("tuxedo" but presumably
an abbreviated form of "smoking jacket"); dancing (usually a "dancing club," not the act of dancing itself); and happy (almost invariably
"drunk" or "tipsy" in Spanish slang).
Nor does the importation process seem to pay too close attention to word morphology and gender. "Punch" became ponche but the
similar ponchar comes from "to puncture." Spelling presents problems
as well. Aware of the penchant of Spanish speakers to place an e before English words beginning with st (estres, estereo, esteik, Rolling
Estones, etc.), some word-importers overcorrect and write "establishment" as stablishment.
What happens to the spelling once a word is in the public
domain is the stuff of orthographers' nightmares. Would you have
guessed, for instance, that estiur, biste, friquearse, and ndilon are your
old friends "steward," "beefsteak," "to freak out" and "nylon?" How
about uachimdn ("watchman") and yes ("jazz"J?
Ironically, using these foreign words is not at all frowned upon
in most Spanish-speaking circles, just as using "chic," "debonair," and
180 BREAKING OUT OF BEGINNER'S SPANISH
"suave" is considered perfectly acceptable if slightly snooty English.
Overall, different rules can apply for verbs, which require greater inventiveness and Spanish endings to adopt "correctly." Few verbs, in
fact, overcome the initial years of sounding downright frightful.
BARBARISMS
Perhaps it just reflects an elitist attitude, but English words
are usually not stigmatized until they reach the level of popular
use-and misuse. When lunch becomes lonche and every corner
boasts a loncheria, then some of the charm of using a foreign word
wears off. Shopping is used by Spanish speakers who can afford to take
shopping trips to the United Statesj pronounced chopping or written
chopin, it becomes a "vulgar" usage. Thus you can sound cosmopolitan talking about tonight's speaker at the clubj call him or her an espiquer, though, and you're liable to have your club membership revoked.
The distinction between cultured imports and vulgar imports is mostly a specious one. Is stand somehow "good Spanish" and
estand "bad Spanish"? Is hamburger fair and hamburguesa foul? You
can be your own judge. Is it acceptable to refer to your car's clutch but
vulgar to call it a cloch? Can you talk uprightly about containers but
should you bow your head in shame when you utter contenedores? Is
rin worse than rim, soquet worse than socket, and ziper worse than
zipper? Is using nipple crass only when you pronounce it "knee-play"
and write it niple? These, for the most part, are the "barbarisms" that
you will be warned away from using by more careful speakers of the
language. And as a rule, these warnings are well taken. You won't impress anyone with your command of English-you will only disappoint
with your laziness in not learning the correct Spanish word.
Other barbarisms are a subtler prey, requiring far greater alertness than is needed to avoid saying breakecito. As English contaminates Spanish in border areas and, indeed, around the world, odd phrasings and meanings begin to emerge from otherwise innocent-looking,
fully legitimate Spanish words. Phrases like vacunar la carpeta appear,
trying to mean "to vacuum the carpet" but in "real" Spanish meaning
"to vaccinate the folder." "Groceries" becomes groserias, which in
fact means "rude remarks" or "offenses." The nice new neighborhood
on the outskirts of town is called a suburbio, though that's what the
slummy part of town was called a generation ago. And on and on.
Most English-influenced barbarisms are considerably less astonishing than these, and Appendix B lists a number of them for your
perusal. Some are so subtle that it is really a judgment call whether
the English is influencing or not. For instance, admitir has as one of
THE BIG MIX 181
its accepted meanings "to accept, to consider provisionally an explanation, thesis, etc. as good or true." But is it correct to say, as newspaper
headlines throughout the Spanish-speaking world do, that e1 criminal
admite su culpa ("the criminal admits his guilt")? Technically, probably not, but you'll find it just the same.
If you think that this is but a semantic splitting of hairs, consider the word remover, which has as its second accepted meaning "to
take away an inconvenience or an obstacle." Still, its first and by far
most common usage in Spanish is "to move about, to stir, to jiggle."
If the person using remover is a doctor speaking to a nurse about the
catheter sticking out of your arm, you may not think it nitpicking at
all to explore the subtleties of this verb.
For many Spanish words, the culprit in their increasing confusion with English cognates may be the news media-specifically the
wire services, whose translated news reports are sent to newspapers,
radio stations, and television networks worldwide. When translating
vast amounts of text a night, as wire-service translators based in New
York or Miami must do, it is easy to accept an easy, ready-made translation instead of a more complicated one. Thus "to remove" becomes
remover instead of the more correct quitarj "admits his guilt" becomes admite su culpa instead of the more precise reconoce su cu1paj
and so on. These hairline infractions then travel the globe and reach
into people's living rooms at night and across their breakfast tables in
the morning. Their influence is incalculable.
Other scholars blame the movie industry-and especially the
movie-dubbing industry-for distorting Spanish grammar. When an
actor in a close-up says "I am waiting for him/' for instance, the temptation is to translate it Estoy esperando por e1 to match the lip movements. This is bad Spanish, but that may be the price we pay for good
dubbing.
REVERSE POCHISMOS
English has not always been the world's dominant language.
For most of the last three centuries, France held that position, and before that-for much of the sixteenth and seventeenth centuries-Spanish was the language that worldy men and women went out of their
way to learn. Those years of dominance have left their mark in the
Spanish we speak today. For just as border and other forms of Spanish
are being infiltrated by Anglicisms and English influences, so were Hispanisms and Spanish influences a "problem" for English and French in
centuries past.
The centuries of Spanish sway span much of the European ex-
182 BREAKING OUT OF BEGINNER'S SPANISH
ploration of the Americas, and it is no surprise to find that many of the
Europeans' discoveries bear Spanish names. Even when the original
name was Taino or Arawakan, it passed through a Spanish filter before
reaching French, English, and the rest of the Old World's languages.
Thus foods ("maize/' "tomato/' "chile"), animals ("peccary/' "llama/'
"jaguar/' "iguana"L and other concepts ("canoe/' "cacique/' "hammock/' "hurricane/' "cigarette/' "cannibal") reached English from native languages via Spanish. (The word "potato/' as a former U.S. vice
president would be happy to hear, has been misspelle~from the start:
it originated as a confusion between the Quechua papa and the Arawakan batata, becoming thereby the Spanish patata and the English
"potato.")
Spanish influence turns up unexpectedly in our everyday
speech. "Tuna" comes from the Spanish, as do two of the three classic
ice cream flavors ("chocolate" and "vanilla"). We have Spanish to
thank for "mosquito" and "cockroach/' as well as for "amigo" (of
course) and "comrade." The Spaniards' dedication to settling the New
World gave us landholding terms like "ranch/' "hacienda/' and "estancia." Early settlers' obsession with racial distinctions gave English
"mulatto/' "creole/, "mestizo/' "negro/' and "sambo." "Coconuts" are
so called because the first Europeans to see them-Spanish or Portuguese sailors-saw scary faces in the stained husks, and thus called
them cocos, or "bogeymen."
Many Spanish words entered English through the U.S. Southwest, which of course used to be the Mexican Northwest. Topographical terms such as "canyon," "barranca/' "arroyo/' "mesa/' and "cuesta"
reached English this way. So did ranching terms like "chaps," "stampede/' "buckaroo/' and "rodeo." The wildlife of the Southwest boasts
many Spanish names: "coyote/' "armadillo/' "chuckwalla," "chaparral/' "saguaro/' "yucca/' "pinyon," "locoweed/' and so on. The states
of the Southwest themselves point back to a Spanish-language heritage: Arizona, California, Colorado, and Nevada (plus Florida and Montana, of course).
From the Southwest and the Old West come some of English's
more colorful Spanish loanwords, a few of which lost more than a
little of their "Spanishness" in the translation. Still, no western would
be complete without some "desperado" "lassoing" a "bronco" (from
desesperado, lazo, and bronco). "Lariat" is but the Spanish la reata,
just as "alligator" is a lumping of ellagarto. If you were a bad guy in
those parts, you could end up in the "calaboose" or the "hoosegow,"
which originated as calabozo and ;uzgado, respectively, When we
call something a "cinch," furthermore, we are recalling the Spanish
cincha, which was the sure and steady "saddle-girth" used by southwestern horsemen. The bad guys, incidentally, wore black "sombreros,"
THE BIG MIX 183
whereas the good guys could be counted on to show up in white "tengallon" (from tan galan, or "so gallant") hats. When they did show up,
they "mosied" down Main Street, while the bad guys went "vamoose"
(both from vamos). After all, these were some tough hombres!
Some words are of more dubious Spanish origin. "Hoosier" and
"hoodlum" are words whose beginnings are lost to history, but which
some scholars speculate had Spanish origins. "Loafer" (and thus "to
loaf") may come from gallofero, an old word for "vagabond," and "palooka" may come from peluca, equally archaic for "severe reproof."
My own personal favorite folk etymology is for "cocktail," perhaps in
part because it's a word that is now considered an Anglicism in Spanish. According to one published version, King Axolotl VIII of Aztec-era
Mexico had a daughter named Xochitl, who invented the cocktail and
had it named after her. Xochitl gradually became Coctel, and the rest is
history-or nearly so.
The process of linguistic infiltration continues in modern
times, of course. With North Americans' more recent "discovery" of
Mexican food, words like "taco," "burrito," "tortilla," and "frijoles"
have entered everyday English. Spanish names playa prominent role in
recent automotive historYi Mustang, Barracuda, Vega, Granada, Matador, Pinto, and Fiesta are some well-known examples. And "macho" is
just as much English as Spanish these days, though the phrase "macho
women," using the masculine form, must sound odd to Spanish speakers. Just as when words travel from English to Spanish, some that have
gone from Spanish to English have had a rough trip. You may hear "mano
a mano" in English these days, for instance, but the usual meaning assigned to it is "man to man," not "hand to hand." Appendix C lists
other words that Spanish has contributed to English over the years.
Many other such words exist, and more still are being imported and
exported, across borders and within them, even as we speak.
As we speak, indeed.
\section{}


\begin{flushright}
    {\tiny{Nunc dimittis}, \DTMnow}
\end{flushright}

\end{document}
